---
title: "Pedigrees, relatedness, and inbreeding"
editor:
  markdown:
    wrap: 72
format:
  html:
    code-fold: true
---

Genotypes and haplotypes describe chromosome copies in an individual. A
[**pedigree**](https://www.genome.gov/genetics-glossary/Pedigree)
records the parentage links that explain how chromosome segments can
pass among individuals. It lets a breeding programme distinguish a close
relative from an unrelated candidate before any relationship matrix is
calculated. The previous page described transmission within a family;
this page follows those paths across a pedigree.

## Pedigrees describe possible transmission paths

A founder is an individual with no recorded parents in the pedigree. Its
offspring form the next generation, and their descendants form later
generations. Two relatives have a common ancestor when their pedigree
paths meet at an earlier individual. A parent and offspring have one
direct path; full siblings have both parents in common; half siblings
have one parent in common.

For an autosomal pedigree with non-inbred, unrelated founders, the
expected proportion of genome shared by a parent and offspring is
one-half. Full siblings also share one-half on average, whereas half
siblings share one-quarter on average. These are expectations over
meioses, not a claim that every pair has exactly that proportion at
every region. Independent assortment and
[recombination](linkage-recombination-ld.qmd) make the realised
chromosome segments inherited by particular relatives differ.

| Relationship | Shared pedigree paths | Expected autosomal sharing under the stated assumptions |
|----------------------|----------------------|----------------------------:|
| Parent and offspring | One parent transmits one of two chromosome copies | 1/2 |
| Full siblings | The same two parents contribute to both offspring | 1/2 |
| Half siblings | One parent contributes to both offspring | 1/4 |

The [**kinship coefficient**](https://pubmed.ncbi.nlm.nih.gov/11986874/)
uses a different scale from the expected sharing proportion. It is the
probability that one allele sampled from each of two individuals is
identical by descent. For these simple non-inbred relationships, kinship
is one-quarter for a parent-offspring or full-sibling pair and
one-eighth for a half-sibling pair. The corresponding expected sharing
proportions are twice those values. This distinction matters because
terms such as "relatedness" can otherwise refer to two different
numerical scales.

## Identical copies need an ancestry statement

Two allele copies are [**identical by descent
(IBD)**](https://pubmed.ncbi.nlm.nih.gov/11986874/) when both descend
from the same ancestral allele copy through the pedigree. IBD is
therefore a statement about inheritance history. Two copies are
**identical by state (IBS)** when their observed allele states match,
whether or not they came from the same ancestral copy. Common alleles
can be IBS without being IBD, and sparse marker data can fail to reveal
a shared ancestral segment.

For the same reason, counting matching genotypes is not a kinship
coefficient. Genotype matching depends on which markers were observed,
their allele frequencies, coding, and chance. Kinship describes a
probability of shared ancestry under a pedigree or genetic model.
Genomic marker data can help estimate realised sharing, but the
construction of pedigree and genomic relationship matrices belongs to
[Genomic prediction](../genomic-prediction/quantitative-genetics.qmd),
not this biological page.

## The principal family in the simulated project data

The project dataset is simulated. It contains 20 founders and 200
offspring in ten full-sib families. `SIRE01` and `DAM01` are the
recorded parents of `IND001` through `IND020`. The code below derives
these statements directly from the pedigree table, then displays the
complete first family.

```{r}
#| label: pedigrees-principal-family
d <- readRDS("../../demo-data/main/demo_data.rds")
founders <- subset(d$pedigree, is.na(sire) & is.na(dam))
offspring <- subset(d$pedigree, !is.na(sire) & !is.na(dam))
family_1_ids <- c("SIRE01", "DAM01", sprintf("IND%03d", 1:20))
family_1 <- subset(d$pedigree, id %in% family_1_ids)

c(founders = nrow(founders), offspring = nrow(offspring))
table(d$family_id)
family_1
```

@fig-principal-family shows a readable subset of this recorded family.
Every box is sex-neutral because the pedigree table records parent roles
as `sire` and `dam` but has no separate sex field. The horizontal line
joins the two recorded parents, and every vertical line shown leads to
one of their offspring. The other 14 offspring have the same two
recorded parents and are listed in the output above.

```{r}
#| label: fig-principal-family
#| fig-width: 8
#| fig-height: 4.2
#| fig-alt: "A sex-neutral pedigree diagram. Boxes labelled SIRE01 and DAM01 are joined at the top. A shared descent line branches to six boxes labelled IND001 through IND006 below, showing that all six have the same two recorded parents."
#| fig-cap: "A readable six-offspring subset of the simulated principal family. `SIRE01` and `DAM01` are the two recorded parents of `IND001` through `IND006`; the R output lists the remaining 14 offspring with the same parent pair. Boxes are intentionally sex-neutral because no sex field is supplied in `d$pedigree`."
display_ids <- c("SIRE01", "DAM01", sprintf("IND%03d", 1:6))
family_display <- subset(d$pedigree, id %in% display_ids)
stopifnot(all(family_display$sire[3:8] == "SIRE01"))
stopifnot(all(family_display$dam[3:8] == "DAM01"))

offspring_x <- seq_len(6)
plot(
  NA,
  xlim = c(0.4, 6.6), ylim = c(0.2, 2.6),
  xlab = "", ylab = "", xaxt = "n", yaxt = "n", bty = "n"
)
segments(3.05, 2.1, 3.95, 2.1, lwd = 1.5)
segments(3.5, 2.1, 3.5, 1.35, lwd = 1.5)
segments(1, 1.35, 6, 1.35, lwd = 1.5)
segments(offspring_x, 1.35, offspring_x, 0.81, lwd = 1.5)
rect(1.95, 1.9, 3.05, 2.3)
rect(3.95, 1.9, 5.05, 2.3)
rect(offspring_x - 0.42, 0.49, offspring_x + 0.42, 0.81)
text(c(2.5, 4.5), c(2.1, 2.1), labels = c("SIRE01", "DAM01"), cex = 0.75)
text(offspring_x, rep(0.65, 6), labels = sprintf("IND%03d", 1:6), cex = 0.7)
```

The diagram records expected paths of descent, not the exact lengths of
DNA segments that each offspring inherited. Full siblings share the same
parents, but each receives a different realised mosaic of parental
chromosome copies.

## Inbreeding and homozygosity

[**Inbreeding**](https://pmc.ncbi.nlm.nih.gov/articles/PMC6304688/)
occurs when the two allele copies in an individual are more likely to
trace to the same ancestral copy because its parents are related. It
increases expected **homozygosity**, the state of carrying matching
alleles at a locus. A homozygous genotype need not be IBD, because the
same allele state can occur independently in the population. Conversely,
inbreeding is an ancestry-based probability that may be estimated using
pedigree or genomic information.

Increased homozygosity can expose recessive deleterious alleles that
were masked in heterozygotes. The resulting average reduction in a trait
associated with survival, fertility, growth, or another fitness-related
measure is called **inbreeding depression**. It is a population-level
possibility, not an automatic diagnosis for one individual with a
related pedigree. Its magnitude depends on the trait, the population's
deleterious variation, its breeding and demographic history, and the
environment.

Pedigrees give a biological map of ancestry. The next page asks a
different question: when a phenotype varies among relatives, how many
loci and biological pathways may contribute to that variation?

## Exercises

1.  Under the stated non-inbred pedigree assumptions, why can full
    siblings have the same expected sharing as a parent and offspring
    but different realised chromosome segments?

::: {.callout-note collapse="true" title="Solution"}
Both relationships have expected autosomal sharing of one-half. A parent
transmits a recombined chromosome copy to each offspring, and two full
siblings receive separate meiotic samples from both shared parents.
Their actual inherited segments can therefore differ despite the same
expectation.
:::

2.  Two individuals have matching alleles at a marker. Why does this not
    by itself show that the copies are IBD or give their kinship
    coefficient?

::: {.callout-note collapse="true" title="Solution"}
Matching observed states establish IBS at that marker. The match may
arise without a shared ancestral copy, especially for a common allele.
IBD and kinship require information or assumptions about ancestry, not
only one matching genotype.
:::

3.  What does the principal-family diagram establish, and what does it
    not establish about the six displayed offspring?

::: {.callout-note collapse="true" title="Solution"}
It establishes the recorded pedigree links: `SIRE01` and `DAM01` are
parents of each displayed offspring. It does not show their exact
realised genome sharing, the parental origin of every allele, or whether
any individual has inbreeding depression.
:::