---
title: "Genetic variation in populations"
editor:
  markdown:
    wrap: 72
---

Suppose a breeding candidate carries one copy of the alternative allele at
a marker. That genotype describes the candidate, but it does not tell us
whether the allele is rare or common among the animals from which candidates
are selected. Answering that question requires a change of scale, from an
individual genotype to counts across a defined population.

The population must be named because frequency is not an intrinsic property
of an allele. The same allele can be common in one breed, uncommon in another,
and absent from a third. It can also change in frequency between generations.
This page develops the population quantities needed to describe those
patterns and uses them to construct a reference model for genotype
proportions.

## From genotype counts to allele frequency

The previous page introduced a biallelic marker genotype as the number of
copies of a chosen alternative allele: 0, 1, or 2. In a diploid population,
each individual contributes two allele copies at an autosomal locus. If

- $n_0$ individuals have dosage 0,
- $n_1$ individuals have dosage 1, and
- $n_2$ individuals have dosage 2,

then the population sample contains $n_0+n_1+n_2=N$ individuals and $2N$
allele copies. The alternative allele occurs once in every dosage-1 genotype
and twice in every dosage-2 genotype. Its count is therefore

$$
n_1 + 2n_2.
$$

The **[allele
frequency](https://www.ncbi.nlm.nih.gov/books/NBK232608/)** is the
proportion of all sampled allele copies that are the allele of interest. We
will write the alternative-allele frequency as

$$
f_A = \frac{n_1 + 2n_2}{2N}.
$$

The reference-allele frequency is $1-f_A$ because the locus is biallelic.
The **[genotype
frequencies](https://www.ncbi.nlm.nih.gov/books/NBK232608/)** are $n_0/N$,
$n_1/N$, and $n_2/N$. These two kinds of frequency answer different
questions. Allele frequency counts copies of an allele, whereas genotype
frequency counts individuals carrying each pair of alleles.

An allele is *fixed* in the sampled population when every sampled allele copy
is the same, so its frequency is 1 and the other allele has frequency 0.
Terms such as *rare* and *common* describe positions on the frequency scale,
but their numerical boundaries depend on the population, sampling design,
and analytical purpose. They should not be treated as universal biological
categories.

## A manual calculation for `SNP007`

Among the 200 offspring in the project data, the observed dosage counts for
`SNP007` are 74, 95, and 31. The number of alternative-allele copies is

$$
95 + 2(31) = 157,
$$

and the total number of allele copies is

$$
2(200) = 400.
$$

Consequently,

$$
f_A = \frac{157}{400} = 0.3925,
$$

and the reference-allele frequency is $1-0.3925=0.6075$. Notice that the
alternative allele is not necessarily the less frequent allele. *Reference*
and *alternative* remain coding labels, as explained on the [variation and
marker page](genetic-variation-markers.qmd#reference-and-alternative-alleles-are-coding-labels).

## Hardy-Weinberg proportions provide a reference

The **[Hardy-Weinberg
model](https://www.ncbi.nlm.nih.gov/books/NBK232608/)** connects allele
frequencies to expected genotype proportions. Consider an autosomal,
biallelic locus in a diploid population. If alternative alleles occur in a
fraction $f_A$ of both male and female gametes and gametes combine at random
with respect to this locus, the possible unions give

$$
\begin{aligned}
P(\text{dosage }0) &= (1-f_A)^2,\\
P(\text{dosage }1) &= 2f_A(1-f_A),\\
P(\text{dosage }2) &= f_A^2.
\end{aligned}
$$

The factor 2 in the heterozygous proportion appears because the alternative
allele can come from either gamete. Random union of gametes produces these
genotype proportions in one generation. For allele frequencies themselves
to remain constant across generations, the classical equilibrium model also
assumes no selection, migration, or mutation and a population large enough
that random sampling does not appreciably change frequencies.

For `SNP007`, multiplying the three proportions by $N=200$ gives

$$
\begin{aligned}
200(0.6075)^2 &= 73.811,\\
200\{2(0.3925)(0.6075)\} &= 95.378,\\
200(0.3925)^2 &= 30.811.
\end{aligned}
$$

Expected counts need not be whole numbers. They describe the average pattern
implied by the model, not a second set of observed individuals.

## Verify the calculation in R

The same calculation can be performed directly from the project genotype
matrix. Taking the mean dosage counts alternative alleles per individual;
dividing by two converts that mean to a proportion of allele copies.

```{r}
#| label: population-snp007-counts
d <- readRDS("../../demo-data/main/demo_data.rds")
g <- d$snp[, "SNP007"]
observed <- table(factor(g, levels = 0:2))
f_alt <- mean(g) / 2
expected <- length(g) * c(
  dosage_0 = (1 - f_alt)^2,
  dosage_1 = 2 * f_alt * (1 - f_alt),
  dosage_2 = f_alt^2
)
rbind(observed = as.numeric(observed), expected = expected)
f_alt
```

The R result matches the manual calculation: the observed counts are 74, 95,
and 31, while the Hardy-Weinberg reference counts are approximately 73.8,
95.4, and 30.8.

```{r}
#| label: population-snp007-plot
#| fig-alt: "Grouped bars compare observed and Hardy-Weinberg reference counts for alternative-allele dosages zero, one, and two at SNP007. The paired bars are nearly equal for every dosage."
#| fig-cap: "Observed genotype counts at SNP007 and counts calculated from Hardy-Weinberg reference proportions. The close bars describe this marker in this sample only."
barplot(
  rbind(observed = as.numeric(observed), HWE_reference = expected),
  beside = TRUE,
  names.arg = c("0", "1", "2"),
  col = c("grey75", "grey25"),
  border = "grey15",
  xlab = "Alternative-allele dosage",
  ylab = "Number of individuals",
  legend.text = c("Observed", "HWE reference"),
  args.legend = list(x = "topright", bty = "n")
)
```

## Why population frequencies and proportions change

Hardy-Weinberg proportions are useful precisely because real populations can
depart from the reference for interpretable reasons. Several mechanisms must
be distinguished because they do not all have the same immediate effect.

**[Mutation](https://www.genome.gov/genetics-glossary/Mutation)** creates new
sequence variants. At an individual locus its per-generation contribution is
usually small, but mutation is the ultimate source of alleles on which other
processes act.

**[Selection](https://www.ncbi.nlm.nih.gov/books/NBK232608/)** occurs when
genotypes differ in survival or reproductive contribution. In a breeding
programme, deliberate choice of parents is a form of selection. If the
selected trait is influenced by a locus, or by a linked locus inherited with
it, allele frequencies can change between the candidate and offspring
generations.

**[Genetic drift](https://www.genome.gov/genetics-glossary/Genetic-Drift)** is
random change in allele frequency caused by finite sampling of parents and
gametes. Even equally fit alleles are not guaranteed to contribute the same
number of copies to the next generation. Drift is stronger when relatively
few individuals contribute genetically.

**[Migration](https://www.ncbi.nlm.nih.gov/books/NBK232608/)**, also called
gene flow in population genetics, moves allele copies between populations
when immigrants reproduce. Introducing animals or germplasm from another
breeding population can therefore change local allele frequencies. The
direction and magnitude depend on the migrants' frequencies and their
reproductive contribution.

**[Non-random mating](https://www.ncbi.nlm.nih.gov/books/NBK232608/)** changes
which allele copies occur together in individuals. Mating relatives, for
example, tends to increase homozygosity and reduce heterozygosity relative to
random mating. It can change genotype proportions immediately even when it
does not, by itself, change the allele frequency in that generation.

**[Population
structure](https://www.ncbi.nlm.nih.gov/books/NBK232608/)** exists when
the sampled individuals come from subpopulations with different ancestry or
allele frequencies. Pooling such groups and calculating one frequency can
produce a heterozygote deficit relative to the pooled Hardy-Weinberg
expectation, even if each group is close to its own reference proportions.
Family groups are also a form of sampling structure because relatives share
segments inherited from recent ancestors.

The **[effective population
size](https://doi.org/10.1038/nrg2526)**, written $N_e$, describes the size
of an idealised population that would experience genetic drift at the same
rate as the population under study. It is not simply the number of animals
alive or genotyped. Unequal numbers of breeding males and females, unequal
family sizes, and large differences in reproductive success can make
$N_e$ much smaller than the census size. This distinction matters in
breeding programmes because using many candidates does not prevent drift if
only a small number become parents.

## Agreement is not proof of equilibrium or independence

The close observed and reference counts for `SNP007` show that one summary of
this marker is numerically consistent with Hardy-Weinberg proportions. They
do not prove that the breeding population satisfies every classical
assumption. A similar match can occur by chance, and a single-locus
comparison cannot diagnose selection, migration, genotyping error, or
population structure on its own.

The sampling design is especially important here. The 200 records are
offspring from ten full-sib families, with 20 offspring per family. Siblings
share parents and are therefore not independent draws from a freely mating
population. Any formal test or downstream model must account for the design
rather than treating numerical agreement at one marker as evidence of
independent sampling.

Later, the Genomic prediction section will use allele frequencies to centre
and sometimes scale marker dosages. This page stops at their biological and
population interpretation so that the modelling operation is introduced
where its purpose can be made explicit.

## Exercises

1. A sample of 20 individuals has genotype counts $n_0=8$, $n_1=10$, and
   $n_2=2$. Calculate the alternative-allele count and frequency.

::: {.callout-note collapse="true" title="Solution"}
The alternative-allele count is $10+2(2)=14$. There are $2(20)=40$
allele copies, so $f_A=14/40=0.35$.
:::

2. For $N=100$ and $f_A=0.30$, calculate the Hardy-Weinberg reference counts
   for dosages 0, 1, and 2.

::: {.callout-note collapse="true" title="Solution"}
The proportions are $0.70^2=0.49$, $2(0.30)(0.70)=0.42$, and
$0.30^2=0.09$. Multiplying by 100 gives expected counts of 49, 42, and 9.
:::

3. Suppose relatives mate more often than expected, but no allele copies enter
   or leave the population. What can change immediately?

::: {.callout-note collapse="true" title="Solution"}
The pairing of alleles into genotypes can change, commonly producing more
homozygotes and fewer heterozygotes. The allele frequency need not change in
that generation because the same allele copies are being paired differently.
:::

4. Explain the difference between allele frequency and genotype frequency at
   a biallelic autosomal locus. Which quantity counts allele copies rather
   than individuals?

::: {.callout-note collapse="true" title="Solution"}
Allele frequency is the proportion of all allele copies that are a chosen
allele, whereas genotype frequencies are the proportions of individuals with
dosages 0, 1, and 2. Allele frequency counts copies, so its denominator is
$2N$ for $N$ diploid individuals.
:::

5. A breeding population has many animals alive but only a few parents that
   contribute very unequally to the next generation. What does this imply
   about effective population size and genetic drift?

::: {.callout-note collapse="true" title="Solution"}
Its effective population size can be much smaller than its census size. The
idealised effective size reflects the rate of drift, and unequal reproductive
contribution increases that rate because relatively few parents transmit the
alleles represented in the next generation.
:::

6. Observed counts at one marker are close to its Hardy-Weinberg reference
   counts in the project data. Why does this not prove equilibrium or that the
   200 offspring are independent draws from a freely mating population?

::: {.callout-note collapse="true" title="Solution"}
A close numerical match can occur by chance and one locus cannot diagnose all
model assumptions. The records are offspring from ten full-sib families, so
siblings share parents and are not independent draws. Selection, migration,
genotyping error, and population structure also cannot be ruled out from this
one comparison.
:::

Allele frequencies describe one locus at a time. The next question is how
alleles at different loci become associated on chromosomes and across a
population.
