---
title: "BayesR"
editor: 
  markdown: 
    wrap: 72
---

BayesR extends BayesA by adding an explicit zero-effect class. BayesA
gives each marker its own continuous variance, while BayesR places each
marker in one of four discrete variance classes, and one of those
classes is an atom at zero. The atom at zero lets BayesR perform exact
variable selection. Small, moderate, and large effect sizes are still
covered by the other three classes.

## Prior

The marker-effect prior is a mixture of four normals:

$$
\beta_j \mid \boldsymbol{\pi}, \sigma_\beta^2
\;\sim\;
\pi_0\,\delta_0
\;+\; \pi_1\,N\!\left(0,\, 0.001\,\sigma_\beta^2\right)
\;+\; \pi_2\,N\!\left(0,\, 0.01\,\sigma_\beta^2\right)
\;+\; \pi_3\,N\!\left(0,\, 0.1\,\sigma_\beta^2\right),
$$ {#eq-bayesr-prior}

with $\sum_k \pi_k = 1$ and a Dirichlet hyperprior on the proportions:

$$
\boldsymbol{\pi} \sim \mathrm{Dir}(\mathbf{1}).
$$

The symbols are:

-   $\beta_j$ : the effect of marker column $j$ on the trait.
-   $\boldsymbol{\pi} = (\pi_0, \pi_1, \pi_2, \pi_3)$ : the mixture
    weights for the four classes. They sum to 1.
-   $\delta_0$ : a point mass at zero. A marker assigned to class 0 has
    $\beta_j = 0$ exactly.
-   $\sigma_\beta^2$ : the base scale of the per-marker effect-variance
    pool. masbayes initialises it from the user-supplied `sigma2_ah`
    (commonly `var(y) * h^2`) and then updates it from the non-zero
    effects each MCMC sweep. This is *not* the trait-level total
    additive genetic variance $\mathrm{Var}(\mathbf{W}\beta)$; it is a
    prior scale parameter.
-   $0.001\,\sigma_\beta^2,\ 0.01\,\sigma_\beta^2,\ 0.1\,\sigma_\beta^2$
    : the variances of the three non-zero classes. The largest is
    $100\times$ the smallest, so the three classes cover a two-order
    range. masbayes stores the scaling factors as `variance_class` and
    defaults to `c(0, 0.001, 0.01, 0.1)`.
-   $\mathrm{Dir}(\mathbf{1})$ : a flat (uniform) Dirichlet prior on
    $\boldsymbol{\pi}$. masbayes updates $\boldsymbol{\pi}$ from this
    prior plus the per-iteration class counts via a conjugate Dirichlet
    posterior.

masbayes sets the default initial mixture proportions with
`run_bayesr(pi_vec = c(0.95, 0.02, 0.02, 0.01))`. The prior starts 95%
of markers in the zero class and 1% in the large-variance class.

**What each class represents:**

| Class | Variance | Biological interpretation |
|----------------|----------------|-----------------------------------------|
| $\pi_0\delta_0$ | 0 | No LD with any QTL |
| $\pi_1 N(0,\, 0.001\sigma_\beta^2)$ | Very small | Background polygenic signal |
| $\pi_2 N(0,\, 0.01\sigma_\beta^2)$ | Moderate | Markers in LD with minor QTLs |
| $\pi_3 N(0,\, 0.1\sigma_\beta^2)$ | Large | Markers tagging major QTLs |

The three non-zero variances cover a $100\times$ range. That range is
wide enough to distinguish background noise from a major-QTL signal. We
do not need a continuous variance per marker.

::: callout-note
**BayesA vs BayesR, which to choose.** BayesA gives each marker a
*continuous* variance from a heavy-tailed prior, which fits traits where
effects form a continuum. BayesR uses *discrete* mixture classes with an
exact zero, and works best when the architecture is genuinely sparse. As
a practical heuristic from masgenomics development: if a BayesR run
returns $\hat\pi_0 > 0.9$, most markers sit in the zero class and the
sparse assumption holds. If $\hat\pi_0 \approx 0$ and effects are
diffuse, BayesA or GBLUP is the more natural fit for that architecture.
:::

------------------------------------------------------------------------

## PIP diagnostics

The **posterior inclusion probability (PIP)** for marker $j$ is the
posterior probability that $j$ is *not* in the zero class:

$$
\mathrm{PIP}_j = 1 - P(\gamma_j = 0 \mid \mathbf{y}),
$$ {#eq-pip}

with:

-   $\gamma_j$ : the class indicator for marker $j$. It takes one of the
    four values $\{0, 1, 2, 3\}$ (notation following Wibowo Chapter 1;
    avoid confusion with the Dirac mass $\delta_0$ above).
-   $\mathbf{y}$ : the phenotype vector for the training population.
-   $P(\gamma_j = 0 \mid \mathbf{y})$ : the posterior probability that
    marker $j$ was assigned to class 0 (the zero class). masbayes
    estimates this directly from the MCMC class assignments.

When the PIP is higher, the evidence for a non-zero marker effect on the
trait is stronger. PIP thresholds in the broader Bayesian-QTL literature
commonly use these cut-offs:

| PIP threshold   | Interpretation                          |
|-----------------|-----------------------------------------|
| PIP $> 0.5$     | Marker likely associated with the trait |
| PIP $> 0.9$     | High-confidence QTL region              |
| PIP $\approx 0$ | No evidence of association              |

**Architecture inference from** $\hat{\boldsymbol{\pi}}$. The posterior
distribution over the mixture proportions is itself informative.
Posterior mass concentrated in class 3
($\sigma_3^2 = 0.1\sigma_\beta^2$) signals a few major QTLs. Mass spread
across classes 1 and 2 signals a polygenic architecture closer to the
infinitesimal model. Inspect $\hat{\boldsymbol{\pi}}$ before settling on
a final model.

------------------------------------------------------------------------

## SNP vs Multi-allelic: W matrix differences

### SNP design matrix

For a biallelic SNP, each column of $\mathbf{W}$ is one centred locus:

$$
W_{ij}^{\text{SNP}} = M_{ij} - 2p_j.
$$

The symbols are:

-   $M_{ij}$ : the raw genotype dosage of individual $i$ at SNP $j$,
    coded $\{0, 1, 2\}$.
-   $p_j$ : the population frequency of the reference allele at SNP $j$.
-   $W_{ij}^{\text{SNP}}$ : the VanRaden-centred entry in $\mathbf{W}$.

The mixture variances $\{0.001, 0.01, 0.1\}\sigma_\beta^2$ scale with
$\sigma_\beta^2$, and masbayes updates $\sigma_\beta^2$ from the data
each MCMC sweep. So the per-column effect range follows whatever
effect-variance pool the training phenotypes currently support.

### Microhaplotype design matrix

For a microhaplotype block, each column is one haplotype contrast within
the block. masbayes builds those contrasts with the
$\mathbf{W}_{\alpha h}$ Da (2015) coding. A block with $h$ haplotypes
contributes $h - 1$ columns. The microhaplotype design matrix is
therefore denser in per-column biological information but smaller in
column count per block than the equivalent SNP panel.

**Consequence for** $\sigma_\beta^2$ scaling. For microhaplotype
markers, masbayes updates $\sigma_\beta^2$ from a design matrix whose
columns already encode *combinations* of SNP alleles in phase. Per unit
of effect size, each column captures more variance than a single SNP
column. At the same estimated $\sigma_\beta^2$, the absolute mixture
variances $\{0.001, 0.01, 0.1\}\sigma_\beta^2$ translate into *larger*
per-column effect ranges for microhaplotypes than for SNP. As a result,
the prior has more room for haplotype-level effects without any manual
tuning.

------------------------------------------------------------------------

## Biological implications of SNPs and microhaplotypes

For SNP markers, a high PIP on a column means that SNP sits in LD with a
QTL. The causal variant could be any of several nearby SNPs. For
microhaplotype markers, a high PIP on a haplotype column flags the
*specific combination* of alleles encoded by that column. By
construction, the PIP attaches to a haplotype block rather than a single
position. This is an interpretive consequence of the multi-allelic Da
(2015) coding.

In terms of $\hat\pi_0$ comparison, we expect microhaplotype panels to
push $\hat\pi_0$ lower than SNP panels on the same trait. The
expectation rests on the richer per-column information content, as
documented in our results (PIC 0.31–0.32 for microhaplotypes and
0.22–0.23 for SNP; 8–11% rare alleles in microhaplotypes and 0.2–7.9% in
SNPs). More columns carry detectable signal, so more columns should
escape the zero class.

Under directional selection, LD between SNP and QTL erodes over
generations. So, our study reported a direct measurement at the
GRM-diagonal level: the microhaplotype GRM diagonal stays at 1.000
across one round of selection, while the SNP GRM diagonal drops to
0.955. The PIP-level consequence follows the same logic. Recombination
rarely breaks haplotype blocks of \<200 bp within a single generation.
The microhaplotype column that tracks a QTL should therefore keep its
class assignment across rounds of selection. We treat the matching PIP
stability claim as a plausible extension of the GRM finding, not as a
measured outcome.

::: callout-note
The BayesR prior structure, the Dirichlet hyperprior on
$\boldsymbol{\pi}$, and the Gibbs / EM sampler are identical for SNP and
microhaplotype inputs. Only $\mathbf{W}$ changes:
`construct_snp_matrix()` for SNP, `construct_wah_matrix()` for
microhaplotypes. The `marker_type` argument signals which internal
convention to apply.
:::

------------------------------------------------------------------------

## See it in Code

**SNP — MCMC**

```{r}
#| eval: false
library(masbayes)

d       <- load_data("large")
y       <- d$pheno$y_cont_qtl_snp
names(y) <- d$pheno$id
X_fixed <- stats::model.matrix(~ sex, data = d$pheno)

W_snp   <- construct_snp_matrix(d$snp, encoding = "vanRaden")$W
wtw_snp <- colSums(W_snp^2)

# pi_vec: prior sparsity — 95% of markers start in the zero class
fit_r_snp_mcmc <- run_bayesr(
  w             = W_snp,
  y             = y,
  wtw_diag      = wtw_snp,
  X             = X_fixed,
  marker_type   = "snp",
  pi_vec        = c(0.95, 0.02, 0.02, 0.01),
  sigma2_ah     = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "mcmc",
  response_type = "gaussian"
)
summary(fit_r_snp_mcmc)
```

**SNP — EM (MAP estimate, faster)**

```{r}
#| eval: false
fit_r_snp_em <- run_bayesr(
  w             = W_snp,
  y             = y,
  wtw_diag      = wtw_snp,
  X             = X_fixed,
  marker_type   = "snp",
  pi_vec        = c(0.95, 0.02, 0.02, 0.01),
  sigma2_ah     = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "em",
  response_type = "gaussian"
)
```

**Microhaplotype — MCMC**

```{r}
#| eval: false
wah    <- construct_wah_matrix(
  hap_matrix           = d$mh,
  colnames             = attr(d$mh, "block_id"),
  allele_freq_filtered = d$allele_freq
)
wtw_mh <- colSums(wah$W_ah^2)

# marker_type = "multiallelic": W_alpha_h columns, richer per-column information
fit_r_mh_mcmc <- run_bayesr(
  w             = wah$W_ah,
  y             = y,
  wtw_diag      = wtw_mh,
  X             = X_fixed,
  marker_type   = "multiallelic",
  pi_vec        = c(0.95, 0.02, 0.02, 0.01),
  sigma2_ah     = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "mcmc",
  response_type = "gaussian"
)
summary(fit_r_mh_mcmc)
```

## References

-   Erbe, M., Hayes, B. J., Matukumalli, L. K., Goswami, S., Bowman, P.
    J., Reich, C. M., Mason, B. A. & Goddard, M. E. (2012). Improving
    accuracy of genomic predictions within and between dairy cattle
    breeds with imputed high-density single nucleotide polymorphism
    panels. *Journal of Dairy Science*, 95(7), 4114–4129.
    [doi:10.3168/jds.2011-5019](https://doi.org/10.3168/jds.2011-5019).
-   Da, Y. (2015). Multi-allelic haplotype model based on genetic
    partition for genomic prediction and variance component estimation
    using SNP markers. *BMC Genetics*, 16, 144.
    [doi:10.1186/s12863-015-0301-1](https://doi.org/10.1186/s12863-015-0301-1).
