---
title: "BayesA"
---

BayesA assigns each marker its own variance, drawn independently from a
scaled inverse-chi-square distribution. This is the key departure from
GBLUP: instead of every marker being shrunk by the same amount, markers
near large-effect QTLs are allowed to "escape" the prior pull while
markers far from any QTL are shrunk hard toward zero.

## Prior

$$
\beta_j \mid \sigma_{\beta_j}^2 \sim N(0,\, \sigma_{\beta_j}^2),
\qquad
\sigma_{\beta_j}^2 \sim \chi^{-2}(\nu,\, S).
$$ {#eq-bayesa-prior}

Integrating out $\sigma_{\beta_j}^2$ gives a **Student-$t$ marginal** on
$\beta_j$:

$$
\beta_j \;\sim\; t\!\left(\nu,\, 0,\, S\right).
$$

The Student-$t$ has heavier tails than a normal. In practice: SNPs near a
QTL, whose posterior $\beta_j$ is large, pull their own $\sigma_{\beta_j}^2$
upward through the Gibbs update — they are allowed to keep a large effect.
SNPs far from any QTL, whose $\beta_j$ stays close to zero, face strong
shrinkage. The shrinkage is *adaptive and marker-specific*, unlike the
uniform ridge penalty of GBLUP.

**Hyperparameter $S$ — where it comes from.**
The scale $S$ is derived so that the prior expectation of each per-marker
variance equals total genetic variance divided by total design variance:

$$
S \;=\; \frac{\sigma_g^2\,(\nu - 2)}{\nu \cdot \sum_{j} \mathrm{Var}(w_j)},
$$ {#eq-bayesa-scale}

giving $E[\sigma_{\beta_j}^2] = \sigma_g^2 / \sum_j \mathrm{Var}(w_j)$.
This anchors the prior to a biologically interpretable value. The default
degrees of freedom $\nu = 4.5$ gives heavy tails while keeping finite
variance; increasing $\nu$ makes the prior more normal (less heavy-tailed).

::: {.callout-note}
BayesA *never* shrinks $\beta_j$ exactly to zero — the Student-$t$
marginal has support on the entire real line. This makes BayesA less
appropriate than BayesR when the true architecture is genuinely sparse
(many true zeros). Use BayesA when you expect a *continuum* of effect
sizes from near-zero to moderate, with no hard cutoff between "zero"
and "non-zero" markers.
:::

---

## SNP vs Multi-allelic: W Matrix Differences

### SNP Design Matrix

$$
W_{ij}^{\text{SNP}} = M_{ij} - 2p_j,
\qquad
\mathrm{Var}(w_j^{\text{SNP}}) = 2p_j(1-p_j).
$$

Each column represents one biallelic locus, centred on its allele
frequency. The denominator in @eq-bayesa-scale is
$\sum_j 2p_j(1-p_j)$ — summed across all SNPs.

### Microhaplotype Design Matrix

$$
\mathbf{W}_{\alpha h} \;\text{(Da 2015 additive coding)},
\qquad
\mathrm{Var}(w_k^{\text{MH}}) = \text{function of haplotype frequencies}.
$$

Each column represents one haplotype contrast within a block — for a
block with $h$ haplotypes, the block contributes $h - 1$ columns. The
denominator in @eq-bayesa-scale becomes $\sum_k \mathrm{Var}(w_k^{\text{MH}})$,
summed across all haplotype contrast columns.

**Consequence for $S$:** Haplotype columns for rare haplotypes have small
variance — $\mathrm{Var}(w_k) \approx 0$ for $p_k \to 0$. The sum in the
denominator is therefore smaller for a rare-haplotype-rich microhaplotype panel than
for an equivalent SNP panel, so $S$ inflates slightly per column. This
gives the prior more room for rare-haplotype effects to express — the
adjustment is automatic, requiring no manual tuning.

---

## Biological Implications of SNP vs microhaplotypes

A SNP effect $\beta_j$ represents the average additive substitution effect
of one allele copy at locus $j$. A microhaplotype effect $\beta_k$ represents the
average additive substitution effect of one *haplotype* — a specific
combination of 2–5 SNP alleles that are already in phase within a block
of <200 bp.

**Effect concentration.** If a QTL is in LD with a specific haplotype,
BayesA on microhaplotype markers concentrates the posterior on that haplotype's
column: $\hat\beta_k$ is large and $\sigma_{\beta_k}^2$ adapts upward for
that column, while the remaining $h - 2$ columns in the same block stay
near zero. In contrast, BayesA on SNP markers spreads the QTL signal
across 2–5 columns — the individual SNPs that together define the
haplotype — producing several moderate effects rather than one large one.
Microhaplotype effects are therefore fewer but more interpretable.

**Stability under selection.** A haplotype block of <200 bp is rarely
broken by recombination within a single generation. The microhaplotype column tracking
a QTL therefore maintains its effect estimate across generations of
selection, whereas individual SNP columns in the same region gradually
lose LD with the QTL as recombination reshuffles the haplotype — a
process that reduces per-SNP $\hat\beta$ magnitudes and inflates the
number of columns with non-negligible effects over time.

::: {.callout-note}
The BayesA prior, the Gibbs sampler, and the EM algorithm are identical
for SNP and microhaplotype inputs. Only $\mathbf{W}$ changes:
`construct_snp_matrix()` for SNP, `construct_wah_matrix()` for microhaplotypes. The
`marker_type` argument in `run_bayesa()` signals which variance scaling
convention to apply when computing $S$ internally.
:::

---

## See it in Code

**SNP — MCMC**

```{r}
#| eval: false
library(masbayes)

d       <- load_data("large")
y       <- d$pheno$y_cont_qtl_snp
names(y) <- d$pheno$id
X_fixed <- stats::model.matrix(~ sex, data = d$pheno)

W_snp  <- construct_snp_matrix(d$snp, encoding = "vanRaden")$W
wtw_snp <- colSums(W_snp^2)

fit_a_snp_mcmc <- run_bayesa(
  w             = W_snp,
  y             = y,
  wtw_diag      = wtw_snp,
  X             = X_fixed,
  marker_type   = "snp",
  nu            = 4.5,
  sigma2_g      = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "mcmc",
  response_type = "gaussian"
)
summary(fit_a_snp_mcmc)
```

**SNP — EM (MAP estimate, faster)**

```{r}
#| eval: false
fit_a_snp_em <- run_bayesa(
  w             = W_snp,
  y             = y,
  wtw_diag      = wtw_snp,
  X             = X_fixed,
  marker_type   = "snp",
  nu            = 4.5,
  sigma2_g      = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "em",
  response_type = "gaussian"
)
```

**Microhaplotype — MCMC**

```{r}
#| eval: false
wah    <- construct_wah_matrix(
  hap_matrix           = d$mh,
  colnames             = attr(d$mh, "block_id"),
  allele_freq_filtered = d$allele_freq
)
wtw_mh <- colSums(wah$W_ah^2)

# marker_type = "multiallelic" uses the W_alpha_h variance scaling for S
fit_a_mh_mcmc <- run_bayesa(
  w             = wah$W_ah,
  y             = y,
  wtw_diag      = wtw_mh,
  X             = X_fixed,
  marker_type   = "multiallelic",
  nu            = 4.5,
  sigma2_g      = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "mcmc",
  response_type = "gaussian"
)
summary(fit_a_mh_mcmc)
```

## References

- Meuwissen, T. H. E., Hayes, B. J. & Goddard, M. E. (2001). Prediction
  of total genetic value using genome-wide dense marker maps. *Genetics*,
  157(4), 1819–1829.
  [doi:10.1093/genetics/157.4.1819](https://doi.org/10.1093/genetics/157.4.1819).
- Da, Y. (2015). Multi-allelic haplotype model based on genetic partition
  for genomic prediction and variance component estimation using SNP
  markers. *BMC Genetics*, 16, 144.
  [doi:10.1186/s12863-015-0301-1](https://doi.org/10.1186/s12863-015-0301-1).
