---
title: "BayesR"
---

BayesR extends BayesA by adding an explicit zero-effect class. Instead of
each marker having a continuous variance, each marker is assigned to one
of four variance classes — including an atom at zero. This allows exact
variable selection while still accommodating markers with small, moderate,
or large effects.

## Prior

$$
\beta_j \mid \boldsymbol{\pi}, \sigma_g^2
\;\sim\;
\pi_0\,\delta_0
\;+\; \pi_1\,N\!\left(0,\, 0.001\,\sigma_g^2\right)
\;+\; \pi_2\,N\!\left(0,\, 0.01\,\sigma_g^2\right)
\;+\; \pi_3\,N\!\left(0,\, 0.1\,\sigma_g^2\right),
$$ {#eq-bayesr-prior}

with $\sum_k \pi_k = 1$ and a Dirichlet hyperprior:

$$
\boldsymbol{\pi} \sim \mathrm{Dir}(\mathbf{1}).
$$

Default initial mixture proportions in `run_bayesr(pi_vec = c(0.95, 0.02, 0.02, 0.01))`:
95 % of markers start in the zero class, 1 % in the large-variance class.

**What each class represents:**

| Class | Variance | Biological interpretation |
|-------|---------|--------------------------|
| $\pi_0\delta_0$ | 0 | No LD with any QTL |
| $\pi_1 N(0,\, 0.001\sigma_g^2)$ | Very small | Background polygenic signal |
| $\pi_2 N(0,\, 0.01\sigma_g^2)$ | Moderate | Markers in LD with minor QTLs |
| $\pi_3 N(0,\, 0.1\sigma_g^2)$ | Large | Markers tagging major QTLs |

The three non-zero variances span three orders of magnitude — enough to
distinguish background noise from major-QTL signal without requiring a
continuous variance per marker.

::: {.callout-note}
**BayesA vs BayesR — which to choose.**
BayesA gives each marker a *continuous* variance from a heavy-tailed
prior — appropriate when effects form a continuum. BayesR uses *discrete*
mixture classes with an exact zero — appropriate when the architecture is
genuinely sparse. A practical diagnostic: if a BayesR run returns
$\hat\pi_0 > 0.9$, most markers are in the zero class and the sparse
assumption holds. If $\hat\pi_0 \approx 0$ and effects are diffuse,
BayesA or GBLUP is more appropriate.
:::

---

## PIP Diagnostics

The **posterior inclusion probability** for marker $j$ is:

$$
\mathrm{PIP}_j = 1 - P(\delta_j = 0 \mid \mathbf{y}),
$$ {#eq-pip}

where $\hat\pi_0^{(j)}$ is the posterior probability that marker $j$
belongs to the zero class.

| PIP threshold | Interpretation |
|---|---|
| PIP $> 0.5$ | Marker likely associated with the trait |
| PIP $> 0.9$ | High-confidence QTL region |
| PIP $\approx 0$ | No evidence of association |

**Architecture inference from $\hat{\boldsymbol{\pi}}$.** The posterior
distribution over mixture proportions is itself informative. Concentration
of mass in class 3 ($\sigma_3^2 = 0.1\sigma_g^2$) signals a few major
QTLs; diffuse mass across classes 1–2 signals a polygenic architecture
closer to infinitesimal. Inspecting $\hat{\boldsymbol{\pi}}$ before
finalising model choice is good practice.

---

## SNP vs Multi-allelic: W Matrix Differences

### SNP Design Matrix

Each column is one centred biallelic locus:
$W_{ij}^{\text{SNP}} = M_{ij} - 2p_j$. The mixture variances
$\{0.001, 0.01, 0.1\}\sigma_g^2$ scale with $\sigma_g^2$, which is
estimated from the data.

### Microhaplotype Design Matrix

Each column is one haplotype contrast within a block
($\mathbf{W}_{\alpha h}$, Da 2015). A block with $h$ haplotypes
contributes $h - 1$ columns — denser in biological information but
fewer in total count than the equivalent SNP panel.

**Consequence for $\sigma_g^2$ scaling.** For microhaplotype markers,
$\sigma_g^2$ is estimated from a design matrix where each column
already captures a *combination* of SNP alleles in phase. Each column
therefore explains more variance per unit of effect size than a single
SNP column. At the same estimated $\sigma_g^2$, the absolute mixture
variances $\{0.001, 0.01, 0.1\}\sigma_g^2$ translate into *larger*
per-column effect ranges for microhaplotypes than for SNP — giving the prior more
room for haplotype-level effects.

---

## Biological Implications of SNP vs microhaplotypes

**PIP interpretation.** For SNP markers, a high PIP on a column means
that SNP is in LD with a QTL — but the causal variant could be any of
several nearby SNPs. For microhaplotype markers, a high PIP on a haplotype column
means the *specific combination* of alleles encoded by that column is
tagging the QTL. Microhaplotype PIPs are more localised and biologically
interpretable: they point to a haplotype block rather than a single
genomic position.

**$\hat\pi_0$ comparison.** Microhaplotype markers tend to produce lower $\hat\pi_0$
than SNP markers on the same dataset — more columns escape the zero
class. This reflects the richer per-column information content of
microhaplotypes: rare haplotype combinations that individual SNPs miss
(higher PIC than SNP) generate stronger signals for the Dirichlet
sampler to pick up, resulting in a less sparse estimated architecture
even when the true architecture is the same.

**PIP stability under selection.** Under directional selection, LD
between SNP and QTL degrades generation by generation. For BayesR on
SNP markers, columns that had high PIP in the training generation
gradually accumulate recombinant chromosomes — their $\hat\beta$ shrinks
and $\hat\pi_0$ drifts upward over subsequent generations. Haplotype
blocks of <200 bp are rarely broken by recombination within a single
generation, so microhaplotype PIPs remain attached to the same haplotype combination
and stay more stable across rounds of selection.

::: {.callout-note}
The BayesR prior structure, the Dirichlet hyperprior on $\boldsymbol{\pi}$,
and the Gibbs/EM sampler are identical for SNP and microhaplotype inputs. Only
$\mathbf{W}$ changes — `construct_snp_matrix()` for SNP,
`construct_wah_matrix()` for microhaplotypes — and the `marker_type` argument
signals which internal convention to apply.
:::

---

## See it in Code

**SNP — MCMC**

```{r}
#| eval: false
library(masbayes)

d       <- load_data("large")
y       <- d$pheno$y_cont_qtl_snp
names(y) <- d$pheno$id
X_fixed <- stats::model.matrix(~ sex, data = d$pheno)

W_snp   <- construct_snp_matrix(d$snp, encoding = "vanRaden")$W
wtw_snp <- colSums(W_snp^2)

# pi_vec: prior sparsity — 95% of markers start in the zero class
fit_r_snp_mcmc <- run_bayesr(
  w             = W_snp,
  y             = y,
  wtw_diag      = wtw_snp,
  X             = X_fixed,
  marker_type   = "snp",
  pi_vec        = c(0.95, 0.02, 0.02, 0.01),
  sigma2_ah     = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "mcmc",
  response_type = "gaussian"
)
summary(fit_r_snp_mcmc)
```

**SNP — EM (MAP estimate, faster)**

```{r}
#| eval: false
fit_r_snp_em <- run_bayesr(
  w             = W_snp,
  y             = y,
  wtw_diag      = wtw_snp,
  X             = X_fixed,
  marker_type   = "snp",
  pi_vec        = c(0.95, 0.02, 0.02, 0.01),
  sigma2_ah     = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "em",
  response_type = "gaussian"
)
```

**Microhaplotype — MCMC**

```{r}
#| eval: false
wah    <- construct_wah_matrix(
  hap_matrix           = d$mh,
  colnames             = attr(d$mh, "block_id"),
  allele_freq_filtered = d$allele_freq
)
wtw_mh <- colSums(wah$W_ah^2)

# marker_type = "multiallelic": W_alpha_h columns, richer per-column information
fit_r_mh_mcmc <- run_bayesr(
  w             = wah$W_ah,
  y             = y,
  wtw_diag      = wtw_mh,
  X             = X_fixed,
  marker_type   = "multiallelic",
  pi_vec        = c(0.95, 0.02, 0.02, 0.01),
  sigma2_ah     = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "mcmc",
  response_type = "gaussian"
)
summary(fit_r_mh_mcmc)
```

## References

- Erbe, M., Hayes, B. J., Matukumalli, L. K., Goswami, S., Bowman, P. J.,
  Reich, C. M., Mason, B. A. & Goddard, M. E. (2012). Improving accuracy
  of genomic predictions within and between dairy cattle breeds with
  imputed high-density single nucleotide polymorphism panels. *Journal of
  Dairy Science*, 95(7), 4114–4129.
  [doi:10.3168/jds.2011-5019](https://doi.org/10.3168/jds.2011-5019).
- Wang, H., Misztal, I., Aguilar, I., Legarra, A., Fernando, R. L.,
  Vitezica, Z., Okimoto, R., Wing, T., Strasburg, R. & Muir, W. M. (2015).
  Genome-wide association mapping including phenotypes from relatives
  without genotypes in a single-step (ssGWAS) for 6-week body weight in
  broiler chickens. *Frontiers in Genetics*, 5, 134.
  [doi:10.3389/fgene.2014.00134](https://doi.org/10.3389/fgene.2014.00134).
- Da, Y. (2015). Multi-allelic haplotype model based on genetic partition
  for genomic prediction and variance component estimation using SNP
  markers. *BMC Genetics*, 16, 144.
  [doi:10.1186/s12863-015-0301-1](https://doi.org/10.1186/s12863-015-0301-1).
