---
title: "BayesA"
---

BayesA assigns each marker its own variance. That variance is drawn from a scaled inverse-chi-square distribution. This is the key departure from GBLUP, where this model shrinks every marker by the same amount. BayesA lets markers near large-effect QTLs escape the prior pull, and it shrinks markers far from any QTL hard toward zero.

## Prior

The two-layer prior on the marker effects is:

$$
\beta_j \mid \sigma_{\beta_j}^2 \sim N(0,\, \sigma_{\beta_j}^2),
\qquad
\sigma_{\beta_j}^2 \sim \chi^{-2}(\nu,\, S^2).
$$ {#eq-bayesa-prior}

Where:

-   $\beta_j$ : the effect of marker column $j$ on the trait.
-   $\sigma_{\beta_j}^2$ : the variance of $\beta_j$, drawn separately for every marker (this is the locus-specific part).
-   $\nu$ : the degrees-of-freedom hyperparameter of the scaled inverse-chi-square. Larger $\nu$ makes the prior closer to normal. The masbayes default $\nu$ is $4.5$.
-   $S^2$ : the squared-scale hyperparameter. It anchors the prior to a biologically interpretable variance (see @eq-bayesa-scale below). The masbayes implementation calls this quantity `s_squared` and uses the parametrisation $E[\sigma_{\beta_j}^2] = \nu S^2/(\nu-2)$.

Integrating $\sigma_{\beta_j}^2$ out of the hierarchy gives a **Student-**$t$ marginal on $\beta_j$:

$$
\beta_j \;\sim\; t\!\left(\nu,\, 0,\, S\right).
$$

Here $t(\nu, 0, S)$ is the location-scale Student-$t$ with $\nu$ degrees of freedom, location $0$, and scale $S$ (the square root of $S^2$). The Student-$t$ has heavier tails than a normal. The practical consequence is adaptive, marker-specific shrinkage. A SNP near a QTL has a large posterior $\beta_j$. That large $\beta_j$ pulls its own $\sigma_{\beta_j}^2$ upward in the Gibbs update, and the marker is allowed to keep a large effect. For a SNP far from any QTL, $\beta_j$ stays close to zero. The shrinkage on such a marker is strong. masbayes does not apply a uniform rule here. The implementation sets the shrinkage marker by marker.

We pick the squared scale $S^2$ so that the prior expectation of each per-marker variance matches a target ratio. That target is the total genetic variance over the total design variance:

$$
S^2 \;=\; \frac{\sigma_g^2\,(\nu - 2)}{\nu \cdot \sum_{j} \mathrm{Var}(w_j)},
$$ {#eq-bayesa-scale}

Where:

-   $\sigma_g^2$ : the total additive genetic variance of the trait.
-   $\sum_j \mathrm{Var}(w_j)$: the sum of column variances of the marker design matrix $\mathbf{W}$, taken across all $j$ marker columns. This is the total design variance.

We then substitute this $S^2$ into the masbayes parametrisation. The result is $E[\sigma_{\beta_j}^2] = \sigma_g^2 / \sum_j \mathrm{Var}(w_j)$. The prior is now tied to something we can interpret biologically. The default $\nu = 4.5$ keeps the tails heavy and the variance finite. Raising $\nu$ pulls the prior toward a normal and softens the heavy tail.

::: callout-note
BayesA *never* shrinks $\beta_j$ exactly to zero. The Student-$t$ marginal has positive density on the whole real line. BayesA is less suited than BayesR when the true architecture is genuinely sparse (many true zeros). Use BayesA when you expect a continuum of effect sizes from near-zero to moderate, with no hard cutoff between "zero" and "non-zero" markers.
:::

------------------------------------------------------------------------

## SNP vs Multi-allelic: W matrix differences

### SNP design matrix

For a biallelic SNP $j$ with reference allele frequency $p_j$, the VanRaden-centred entry is:

$$
W_{ij}^{\text{SNP}} = M_{ij} - 2p_j,
\qquad
\mathrm{Var}(w_j^{\text{SNP}}) = 2p_j(1-p_j).
$$

Where:

-   $M_{ij}$ : the raw genotype dosage of individual $i$ at SNP $j$, coded $\{0, 1, 2\}$ for the count of reference-allele copies.
-   $p_j$ : the population frequency of the reference allele at SNP $j$.
-   $W_{ij}^{\text{SNP}}$ : the centred entry that enters $\mathbf{W}$.
-   $\mathrm{Var}(w_j^{\text{SNP}}) = 2p_j(1-p_j)$ : the variance of column $j$ of $\mathbf{W}$ under Hardy–Weinberg, used as one term in the $\sum_j \mathrm{Var}(w_j)$ that appears in @eq-bayesa-scale.

Each column of $\mathbf{W}$ represents one biallelic locus. The denominator in @eq-bayesa-scale is $\sum_j 2p_j(1-p_j)$, summed across all SNPs.

### Microhaplotype design matrix

For a microhaplotype block, masbayes follows the Da (2015) additive coding. For individual $i$ at microhaplotype $k$ with frequency $p_k$, the entry depends on how many copies of haplotype $k$ the individual carries:

$$
W_{ah}^{ij,k} = \begin{cases}
  -2(1-p_k), & \text{if } i = j = k \quad\text{(homozygous for } k\text{)} \\
  -(1-2p_k), & \text{if } i = k \text{ or } j = k \quad\text{(heterozygous, one copy)} \\
  2p_k, & \text{if } i, j \neq k \quad\text{(no copies of } k\text{)}
\end{cases}
$$

Where:

-   $i, j$ : the two haplotypes carried by the individual at this block (one from each parent).
-   $k$ : the haplotype indexed by the current column.
-   $p_k$ : the population frequency of haplotype $k$ in this block.
-   $W_{ah}^{ij,k}$ : the entry placed into the column-$k$ slot for an individual whose genotype is $(i, j)$.

Under Hardy–Weinberg genotype frequencies, the expectation of every column is zero. The column is then already allele-frequency-centred without an extra step. masbayes drops the haplotype with the highest frequency in each block as the baseline. This keeps $\mathbf{W}_{ah}$ full rank. The dropped effect is then absorbed into the intercept $\mu$. The column variance $\mathrm{Var}(w_k^{\text{MH}})$ is a function of the haplotype frequencies in the block.

Each column of $\mathbf{W}_{ah}$ represents one haplotype contrast within a block. A block with $h$ haplotypes contributes $h - 1$ columns. The denominator in @eq-bayesa-scale becomes $\sum_k \mathrm{Var}(w_k^{\text{MH}})$, summed across all haplotype contrast columns in the panel.

**Consequence for** $S$. Columns for rare haplotypes have small variance. As $p_k \to 0$, $\mathrm{Var}(w_k) \to 0$. As a result, the denominator sum is smaller for a microhaplotype block rich in rare haplotypes than for an equivalent SNP panel. The numerator of @eq-bayesa-scale does not change. Per column, $S$ then inflates slightly. The prior gains more room for rare-haplotype effects to express themselves. The adjustment falls out of the construction. We do not tune it by hand.

------------------------------------------------------------------------

## Biological Implications of SNP vs microhaplotypes

A SNP effect $\beta_j$ represents the average additive substitution effect of one allele copy at locus $j$. A microhaplotype effect $\beta_k$ represents the average additive substitution effect of one *haplotype*, which is a specific combination of 2–5 SNP alleles that are already in phase within a block of \<200 bp.

Suppose a QTL is in LD with one specific haplotype. With microhaplotype markers, masbayes concentrates the per-QTL posterior on that haplotype's column. The estimate $\hat\beta_k$ is large, and the per-marker variance $\sigma_{\beta_k}^2$ adapts upward for that column. The remaining $h - 2$ columns in the same block stay near zero. With SNP markers, this is different. The same QTL signal is spread across the SNP columns that together define the haplotype.

A haplotype block of \<200 bp is rarely broken by recombination within a single generation. The microhaplotype column that tracks a QTL therefore keeps its effect estimate across generations of selection. Individual SNP columns in the same region gradually lose LD with the QTL as recombination reshuffles the haplotype. Per-SNP $\hat\beta$ magnitudes shift across generations. Prediction accuracy decays by roughly 5% per generation.

::: callout-note
The BayesA prior, the Gibbs sampler, and the EM algorithm are identical for SNP and microhaplotype inputs. Only $\mathbf{W}$ changes: `construct_snp_matrix()` for SNP, `construct_wah_matrix()` for microhaplotypes. The `marker_type` argument in `run_bayesa()` signals which variance scaling convention to apply when computing $S$ internally.
:::

------------------------------------------------------------------------

## See it in Code

**SNP — MCMC**

```{r}
#| eval: false
library(masbayes)

d       <- load_data("large")
y       <- d$pheno$y_cont_qtl_snp
names(y) <- d$pheno$id
X_fixed <- stats::model.matrix(~ sex, data = d$pheno)

W_snp  <- construct_snp_matrix(d$snp, encoding = "vanRaden")$W

fit_a_snp_mcmc <- run_bayesa(
  w             = W_snp,
  y             = y,
  X             = X_fixed,
  marker_type   = "snp",
  nu            = 4.5,
  sigma2_g      = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "mcmc",
  response_type = "gaussian"
)
summary(fit_a_snp_mcmc)
```

**SNP — EM (MAP estimate, faster)**

```{r}
#| eval: false
fit_a_snp_em <- run_bayesa(
  w             = W_snp,
  y             = y,
  X             = X_fixed,
  marker_type   = "snp",
  nu            = 4.5,
  sigma2_g      = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "em",
  response_type = "gaussian"
)
```

**Microhaplotype — MCMC**

```{r}
#| eval: false
wah    <- construct_wah_matrix(
  hap_matrix           = d$mh,
  colnames             = attr(d$mh, "block_id"),
  allele_freq_filtered = d$allele_freq
)

# marker_type = "multiallelic" uses the W_alpha_h variance scaling for S
fit_a_mh_mcmc <- run_bayesa(
  w             = wah$W_ah,
  y             = y,
  X             = X_fixed,
  marker_type   = "multiallelic",
  nu            = 4.5,
  sigma2_g      = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "mcmc",
  response_type = "gaussian"
)
summary(fit_a_mh_mcmc)
```

## References

-   Meuwissen, T. H. E., Hayes, B. J. & Goddard, M. E. (2001). Prediction of total genetic value using genome-wide dense marker maps. *Genetics*, 157(4), 1819–1829. [doi:10.1093/genetics/157.4.1819](https://doi.org/10.1093/genetics/157.4.1819).
-   Da, Y. (2015). Multi-allelic haplotype model based on genetic partition for genomic prediction and variance component estimation using SNP markers. *BMC Genetics*, 16, 144. [doi:10.1186/s12863-015-0301-1](https://doi.org/10.1186/s12863-015-0301-1).
