---
title: "Biallelic & Multi-allelic"
---

## 1. What is the difference between biallelic and multi-allelic?

**Polymorphism** is the existence of two or more variants of a DNA
sequence at the same physical location (locus) in a population.
Without polymorphism there is nothing to predict from — every
individual would carry the same alleles and the genetic component of
phenotypic variance would be zero. Quantitative-genetic analyses
therefore rely on **genetic markers**: positions in the genome where
multiple alleles segregate, so that allelic state correlates (through
linkage or causation) with the trait being predicted.

A marker is **biallelic** when exactly two alleles segregate in the
population, and **multi-allelic** when three or more do.

| Marker family | Alleles per locus | Coding | Typical width |
|---|---|---|---|
| **SNP** (single-nucleotide polymorphism) — biallelic | 2 (e.g. A vs G at a single base) | dosage $\{0, 1, 2\}$ of the alternate allele | one base pair |
| **Microhaplotype** — multi-allelic | up to $2^k$ for a $k$-SNP window, typically 4–20 observed | integer-encoded haplotype id per strand | tens to hundreds of base pairs |

The strengths and limits of each:

| | SNP (biallelic) | Microhaplotype (multi-allelic) |
|---|---|---|
| **Pros** | Cheap to genotype and impute; massive panels (100k–1M markers); robust to single-SNP errors; simple 0/1/2 coding feeds every published mixed-model and Bayesian routine | Captures within-block linkage phase that per-SNP coding loses; each locus carries more bits of information; larger expected effect per locus → easier to detect causal variants; fewer parameters than the corresponding SNP set when within-block LD is strong |
| **Limits** | Each locus only weakly tags a causal variant unless in tight LD; many small effects → polygenic / infinitesimal regime; sensitive to allele-frequency drift across populations | Requires phased reads (Beagle / FImpute style); a single mis-called SNP within a block can flip the whole haplotype id; population-specific haplotype frequencies make cross-population transfer harder than SNP transfer |

The model machinery (BLUP family, Bayesian alphabet) is identical in
form for both — what changes is **how the genotype matrix is built**.
Sections 2–4 below explain that construction; the model itself is
covered in
[Mixed Model & BLUP Family](mixed-model-blup-family.qmd) and
[Bayesian Alphabet](bayesian-alphabet.qmd).

## 2. What is the form of raw data from SNPs and microhaplotypes?

### SNP genotypes

Raw SNP data is a wide integer table: one row per individual, one
column per SNP, values in $\{0, 1, 2\}$ counting the alternate-allele
copies.

| ID | SNP001 | SNP002 | SNP003 | … |
|---|---|---|---|---|
| ind_1 | 1 | 0 | 2 | … |
| ind_2 | 2 | 1 | 0 | … |
| ind_3 | 0 | 2 | 1 | … |

Both packages consume it as an integer matrix
($\texttt{rownames()}$ = individual IDs, $\texttt{colnames()}$ = SNP
IDs). See [Input Data → SNP](../input-data/snp.qmd) for the bundled
demo dataset.

### Microhaplotype genotypes

For microhaplotypes, the final genotype data will be:

```
ID       hap_1_1  hap_1_1  hap_1_2  hap_1_2  hap_1_3  hap_1_3
ind_1        2        1        1        2        3        3
ind_2        1        3        2        2        3        2
ind_3        3        1        2        2        2        3
```

This genotype data consists of numerical microhaplotype alleles in
diploid format, where each number represents a unique combination of
SNP sequences in a specific genomic block. Each locus is presented in
two side-by-side columns to show the pair of haplotypes inherited from
each individual's parents.

### Why microhaplotype genotype data needs $\mathbf{W}_{\alpha h}$

To construct this complex matrix,
[Da, Y. (2015)](https://link.springer.com/article/10.1186/s12863-015-0301-1)
introduced a special coding system that serves to balance the
statistical influence of each variant based on its frequency. This
system mathematically assigns larger deviation values to rare alleles
and smaller values to common alleles. This aims to ensure that rare
variants have a proportional influence on genomic predictions, while
ensuring that the average additive effect across the population
remains zero to keep the model balanced. Technically, the coding rule
is applied to each individual $i$ for haplotype $k$ with population
frequency $p_k$ as follows:

$$
W_{i,k} = \begin{cases}
  -2(1-p_k) & \text{if genotype } k/k \text{ (homozygous: both copies are } k\text{)} \\
  -(1-2p_k) & \text{if genotype } k/\ell \text{ (heterozygous: one copy of } k\text{)} \\
  2p_k & \text{if genotype } \ell/m \text{ (non-carrier: zero copies of } k\text{)}
\end{cases}
$$ {#eq-wah-coding}

where $k \neq \ell \neq m$ are distinct alleles. Note that we drop the
most frequent allele (baseline) from each block, keeping only $h-1$
alleles to provide unique solutions.

Allele frequencies are calculated from phased haplotypes (the two DNA
copies each individual inherited from their parents), so tools like
[Beagle](https://faculty.washington.edu/browning/beagle/beagle.html)
(population-based phasing/imputation) and
[FImpute](https://animalbiosciences.uoguelph.ca/~msargol/fimpute/)
(pedigree-based phasing/imputation) are crucial. For each haplotype
block, we count how many times each allele appears across all
individuals and divide by total haplotypes ($2n$ for $n$ individuals).
For example, if allele 3 appears in 5 out of 200 haplotypes
(100 individuals), its frequency is $p_k = 5/200 = 0.025$ (2.5%).

This standardisation ensures three critical properties. First, the
matrix is mean-centered with $\mathbb{E}[W_k] = 0$, meaning that
positive and negative deviations balance out across the population.
Second, variance scales with $\text{Var}(W_k) \propto 2p_k(1-p_k)$,
matching Hardy-Weinberg genetic expectations where intermediate-
frequency alleles contribute most variance. Third, the haplotype
genomic relationship matrix $\mathbf{G} = \mathbf{W}\mathbf{W}^\top /
k_{\alpha h}$ (where $k_{\alpha h} = \text{tr}(\mathbf{G}) / n$)
becomes comparable to SNP-based GRM by
[VanRaden, P. M. (2008)](https://www.journalofdairyscience.org/article/S0022-0302(08)70990-1/fulltext),
enabling proven statistical methods like GBLUP to work directly with
multi-allelic markers.

#### Example

Consider 4 individuals genotyped at **Locus 1_1** which has 4 distinct
alleles. Allele 2 is the most frequent (60%), chosen as baseline and
dropped. Allele 1 (37.5%) and Allele 3 (2.5% — rare) are included in
the matrix. Allele 4 (0%) is ignored as it is monomorphic in this
sample.

**Step 1: Phased genotypes (maternal/paternal)**

- ID1: 1/3 → carries two different non-baseline alleles, alleles 1 and 3
- ID2: 1/2 → carries one non-baseline and one baseline allele, allele 1 and 2 (baseline)
- ID3: 2/2 → homozygous for the baseline allele (allele 2)
- ID4: 3/3 → homozygous for the rare allele (allele 3)

**Step 2: Apply coding rule**

|     | Genotype | allele1 calculation | allele1 value | allele3 calculation | allele3 value |
|-----|----------|---------------------|---------------|---------------------|---------------|
| ID1 | 1/3      | $-(1-2 \times 0.375) = -0.25$ | **−0.25** | $-(1-2 \times 0.025) = -0.95$ | **−0.95** |
| ID2 | 1/2      | $-(1-2 \times 0.375) = -0.25$ | **−0.25** | $2 \times 0.025 = 0.05$ | **0.05** |
| ID3 | 2/2      | $2 \times 0.375 = 0.75$ | **0.75** | $2 \times 0.025 = 0.05$ | **0.05** |
| ID4 | 3/3      | $2 \times 0.375 = 0.75$ | **0.75** | $-2(1-0.025) = -1.95$ | **−1.95** |

**Step 3: Final $\mathbf{W}_{\alpha h}$ matrix**

|     | hap_1_1_allele1 | hap_1_1_allele3 |
|-----|-----------------|-----------------|
| ID1 | −0.25           | −0.95           |
| ID2 | −0.25           | 0.05            |
| ID3 | 0.75            | 0.05            |
| ID4 | 0.75            | −1.95           |

The matrix $\mathbf{W}_{\alpha h}$ uses allele frequencies to "weight"
the genotypes. Here is how to interpret the resulting values:

- **Rare alleles (e.g., Allele 3 at 2.5%):** Because this allele is
  scarce, its presence creates a massive statistical "deviation."
    - **Carriers** (ID1 & ID4) get large negative values (**−0.95** to
      **−1.95**), making them stand out sharply.
    - **Non-carriers** (ID2 & ID3) get a value near zero (**0.05**),
      meaning they stay close to the population average.
- **Common alleles (e.g., Allele 1 at 37.5%):** Because this allele
  is common, its presence is less surprising to the model.
    - **Carriers** (ID1 & ID2) receive moderate negative values
      (**−0.25**).
    - **Non-carriers** (ID3 & ID4) receive moderate positive values
      (**0.75**).
    - The gap between carrier and non-carrier is smaller here because
      the allele is less "informative" than a rare one.

This weighting ensures rare, potentially high-impact genetic variants
contribute more to genomic predictions than common background
variation.

However, compared to biallelic markers (SNPs) that have a simple
0/1/2 coding and only one effect per marker to estimate, multi-allelic
markers, such as haplotypes and microhaplotypes, have a more expanded
parameter space. Each haplotype block can have multiple alleles
(often 10–20 or more), and after dropping the baseline, we must
estimate separate effects for each remaining allele. So, while the
linear model equation seems similar:

$$
\mathbf{y} = \mathbf{W}_{\alpha h}\boldsymbol{\beta}_{\alpha h} + \mathbf{e}
$$ {#eq-wah-regression}

where $\mathbf{y}$ is the phenotype vector, $\mathbf{W}_{\alpha h}$ is
our multi-allelic matrix, $\boldsymbol{\beta}_{\alpha h}$ contains all
allele effects, and $\mathbf{e}$ is residual error. The key difference
is that $\boldsymbol{\beta}$ now has thousands or tens of thousands of
parameters instead of hundreds of thousands of SNPs, but each
parameter carries more biological information.

## 3. Matrix for models

Different model families consume the marker matrix in different forms.

### Frequency-based models (BLUP family) need a GRM

Methods in the [Mixed Model & BLUP Family](mixed-model-blup-family.qmd)
treat the additive genetic value as a random effect with covariance
proportional to a **relationship matrix** $\mathbf{K}$. With genotype
data, $\mathbf{K} = \mathbf{G}$ is the **genomic relationship matrix
(GRM)** — an $n \times n$ symmetric positive (semi-)definite matrix
whose entry $G_{ij}$ captures the realised additive-genetic similarity
between individuals $i$ and $j$. GBLUP's mixed-model equations
(see [Henderson MME](mixed-model-blup-family.qmd#eq-mme)) require
$\mathbf{G}^{-1}$ directly; the structure of $\mathbf{G}$ — not the
raw marker matrix — is what enters the solve.

**For SNPs (VanRaden 2008, Method 1):**
$$
\mathbf{G}_u = \frac{\mathbf{W}\mathbf{W}^\top}{2 \sum_{j=1}^{p} p_j (1 - p_j)},
\qquad
\mathbf{W}_{ij} = M_{ij} - 2 p_j,
\quad M_{ij} \in \{0, 1, 2\}.
$$ {#eq-grm-snp}

The denominator scales the mean diagonal of $\mathbf{G}_u$ to
$\approx 1 + F$, matching the pedigree $\mathbf{A}$.

**For microhaplotypes:** the same construction with
$\mathbf{W}_{\alpha h}$ from @eq-wah-coding in place of
$\mathbf{W}$, with denominator
$\sum_b \sum_{a=1}^{A_b - 1} 2 f_{b,a}(1 - f_{b,a})$ (sum across all
non-baseline alleles in all blocks):

$$
\mathbf{G}_{\text{mh}}
= \frac{\mathbf{W}_{\alpha h}\,\mathbf{W}_{\alpha h}^\top}
       {\sum_b \sum_{a=1}^{A_b - 1} 2 f_{b,a}(1 - f_{b,a})}.
$$ {#eq-g-mh}

### Bayesian models need only the centred design matrix

The [Bayesian Alphabet](bayesian-alphabet.qmd) (BayesA, BayesR) does
**not** form a GRM. The regression
$\mathbf{y} = \mathbf{W}\boldsymbol{\beta} + \boldsymbol{\epsilon}$
is fit directly in marker space — the prior on $\boldsymbol{\beta}$
does the shrinkage that, in GBLUP, came from $\mathbf{G}$. The matrix
$\mathbf{W}$ must still be **centred** so the per-marker effects are
on a common scale and the intercept absorbs the mean:

- **SNPs**: VanRaden centring $W_{ij} = M_{ij} - 2 p_j$ (default in
  `masbayes::construct_snp_matrix(encoding = "vanRaden")`), or
  per-column z-score $W_{ij}^z = (M_{ij} - 2 p_j) / \sqrt{2 p_j (1 - p_j)}$
  (`encoding = "zscore"`).
- **Microhaplotypes**: the same $\mathbf{W}_{\alpha h}$ from
  @eq-wah-coding goes straight into `masbayes::run_bayesa()` or
  `run_bayesr()` (Bayesian variants accept multi-allelic design
  matrices natively).

Bayesian methods sidestep the $\mathbf{G}^{-1}$ solve entirely, which
matters when $p \gg n$ (BLUP gets a large dense inverse; Bayesian
samplers iterate per marker).

## 4. Important function

Both packages expose builder functions for these matrices. The
canonical demo dataset (`load_data("large")`) is used here; for
detail, see [Input Data → SNP](../input-data/snp.qmd) and
[Microhaplotypes](../input-data/microhaplotypes.qmd).

### Full GRM for GBLUP (`masreml`)

```{r}
#| eval: false
library(masreml)
d <- load_data("large")

# SNP-based GRM (VanRaden Method 1) — eq @eq-grm-snp
G_snp <- build_G_snp(d$snp)
dim(G_snp)                          # 200 x 200

# Multi-allelic GRM from compact d$mh — eq @eq-g-mh
G_mh  <- build_G_mh(mh_list = d$mh, ids = rownames(d$mh))
dim(G_mh)                           # 200 x 200

# Passed to the BLUP solve via the G slot (named list of relationship matrices):
fit_g_snp <- masreml(
  y     = d$pheno$y_cont_qtl_snp,
  G     = list(snp_add = G_snp),
  trait = "continuous"
)
# Or let masreml build it internally from raw markers:
fit_m_snp <- masreml(
  y       = d$pheno$y_cont_qtl_snp,
  markers = list(snp_add = d$snp),
  trait   = "continuous"
)
```

### Centred matrix for Bayesian methods (`masbayes`)

```{r}
#| eval: false
library(masbayes)
d <- load_data("large")

# SNP design matrix — VanRaden centring (default).
# construct_snp_matrix() returns a list; extract $W for the design matrix.
W_snp <- construct_snp_matrix(d$snp, encoding = "vanRaden")$W
# Alternative: per-column z-score
W_snp_z <- construct_snp_matrix(d$snp, encoding = "zscore")$W

# Multi-allelic W_alpha-h — eq @eq-wah-coding applied per block
wah <- construct_wah_matrix(
  hap_matrix             = d$mh,
  colnames               = attr(d$mh, "block_id"),
  allele_freq_filtered   = d$allele_freq,
  reference_structure    = NULL,
  drop_baseline          = TRUE
)
W_mh <- wah$W_ah                     # n x ~149

# Feed into BayesA / BayesR (see Bayesian Alphabet page for full call):
# masbayes::run_bayesa(w = W_snp, y = ..., wtw_diag = colSums(W_snp^2), ...)
# masbayes::run_bayesr(w = W_mh,  y = ..., wtw_diag = colSums(W_mh ^2), ...)
```

## References

- VanRaden, P. M. (2008). *Efficient methods to compute genomic
  predictions.* Journal of Dairy Science, 91(11), 4414–4423.
  [doi:10.3168/jds.2007-0980](https://doi.org/10.3168/jds.2007-0980).

- Da, Y. (2015). *Multi-allelic haplotype model based on genetic
  partition for genomic prediction and variance component estimation
  using SNP markers.* BMC Genetics, 16, 144.
  [doi:10.1186/s12863-015-0301-1](https://doi.org/10.1186/s12863-015-0301-1).

- Browning, S. R. & Browning, B. L.
  *Beagle 5: rapid genotype imputation and haplotype phasing.*
  [faculty.washington.edu/browning/beagle](https://faculty.washington.edu/browning/beagle/beagle.html).

- Sargolzaei, M., Chesnais, J. P. & Schenkel, F. S. (2014).
  *A new approach for efficient genotype imputation using information
  from relatives.* BMC Genomics, 15, 478. FImpute software:
  [animalbiosciences.uoguelph.ca/~msargol/fimpute](https://animalbiosciences.uoguelph.ca/~msargol/fimpute/).
