---
title: "Descriptive statistics"
---

A breeding programme may record body weight, disease score, or harvest yield
for hundreds of individuals. Before fitting a genomic prediction model, the
breeder needs to know what values are typical, how much individuals differ,
and whether a few unusual records deserve investigation. **[Descriptive
statistics](https://www.itl.nist.gov/div898/handbook/eda/section3/eda3.htm)**
answer these questions by reducing a collection of observations to numerical
and graphical summaries. They describe the recorded sample; by themselves,
they do not explain why the observations differ or predict an unobserved
individual.

## Centre and spread answer different questions

Consider two hypothetical groups of five selection candidates. Both groups
have a mean weight of 10 kg, but their values differ in spread.

| Candidate | Group A (kg) | Group B (kg) |
|---:|---:|---:|
| 1 | 8 | 4 |
| 2 | 9 | 7 |
| 3 | 10 | 10 |
| 4 | 11 | 13 |
| 5 | 12 | 16 |

The equal means do not make the groups equivalent. Group A lies close to 10
kg, whereas Group B includes much lighter and heavier candidates. A measure
of centre identifies a typical value. A measure of spread describes how far
observations tend to lie from that centre. Genomic prediction needs both:
marker effects cannot help distinguish candidates if the target trait has no
variation, and a centre alone hides the amount of variation available to the
model.

## Measures of centre

The **[arithmetic mean](https://www.itl.nist.gov/div898/handbook/eda/section3/eda351.htm)**
of observations $y_1,\ldots,y_n$ is

$$
\bar y = \frac{1}{n}\sum_{i=1}^{n} y_i,
$$

where $n$ is the number of observations and $\bar y$ is their sample
mean. Every observation contributes its magnitude, so one extreme value can
move the mean substantially. This sensitivity is useful when the magnitude
of every measurement matters, but it also means that data errors can distort
the summary.

The **[median](https://www.itl.nist.gov/div898/handbook/eda/section3/eda351.htm)**
is the middle value after sorting the observations. For an even number of
observations, it is conventionally the mean of the two middle values. Because
the median depends on order rather than the distance of every observation
from the centre, it changes less than the mean when one observation is
extreme.

The **[mode](https://www.itl.nist.gov/div898/handbook/eda/section3/eda351.htm)**
is the most frequent observed value or category. A sample can have no repeated
value, one mode, or several equally frequent modes. For continuously measured
traits, exact repetitions may be rare and the mode can depend on measurement
precision or histogram bins. It is often more informative for genotype or
phenotype categories than for precise continuous measurements.

## Measures of position and spread

A **[quantile](https://www.itl.nist.gov/div898/handbook/prc/section2/prc262.htm)**
divides ordered observations by cumulative proportion. The first quartile
$Q_1$ marks the 25th percentile, the median marks the 50th, and the third
quartile $Q_3$ marks the 75th. Software packages use several interpolation
conventions for sample quantiles, so a small sample can yield slightly
different values under different conventions.

The **[range](https://www.itl.nist.gov/div898/handbook/eda/section3/eda356.htm)**
is the maximum minus the minimum. It is easy to interpret but uses only two
observations. The **[interquartile range](https://www.itl.nist.gov/div898/handbook/eda/section3/eda356.htm)**,

$$
\operatorname{IQR}=Q_3-Q_1,
$$

describes the width of the middle half of the data. It is less sensitive than
the range to observations in the tails.

Variance begins with a **[deviation](https://en.wikipedia.org/wiki/Deviation_(statistics))**,
$y_i-\bar y$, the signed distance between one observation and the mean.
Deviations sum to zero, so simply averaging them would report no spread.
Squaring prevents positive and negative deviations from cancelling. The
**[sample variance](https://www.itl.nist.gov/div898/handbook/eda/section3/eda356.htm)**
is

$$
s^2=\frac{1}{n-1}\sum_{i=1}^{n}(y_i-\bar y)^2.
$$

Variance has squared units. If weight is recorded in kilograms, its variance
is measured in square kilograms. The **[sample standard deviation](https://www.itl.nist.gov/div898/handbook/eda/section3/eda356.htm)**
is $s=\sqrt{s^2}$, which restores the original measurement unit.

::: {.callout-note collapse="true" title="Why divide sample variance by n - 1?"}
Assume $Y_1,\ldots,Y_n$ are independent and identically distributed with
finite mean $\mu$ and variance $\sigma^2$. Estimating the mean from the same
observations forces the deviations from $\bar Y$ to sum to zero. This is one
linear constraint, leaving $n-1$ degrees of freedom; it does not mean that
the individual deviations are independent.

The identity

$$
\sum_{i=1}^{n}(Y_i-\bar Y)^2
=\sum_{i=1}^{n}(Y_i-\mu)^2-n(\bar Y-\mu)^2
$$

makes the correction explicit. Taking expectations gives

$$
E\left[\sum_{i=1}^{n}(Y_i-\bar Y)^2\right]
=n\sigma^2-n\operatorname{Var}(\bar Y)
=n\sigma^2-n\left(\frac{\sigma^2}{n}\right)
=(n-1)\sigma^2.
$$

Dividing by $n-1$ therefore makes $s^2$ unbiased for $\sigma^2$ under these
assumptions. If the observations are the entire population of interest rather
than a random sample from it, division by $n$ answers a different descriptive
question.
:::

## Manual calculation with the project data

The recurring dataset in this curriculum is simulated. It contains 200
individuals, their marker genotypes, phenotypes, pedigrees, and true breeding
values. For a calculation that can be checked by hand, take the first five
continuous phenotypes and round them to whole units. Rounding is used only for
this teaching calculation; applied summaries retain the raw values.

```{r}
#| label: descriptive-data
d <- readRDS("../../demo-data/main/demo_data.rds")
y <- round(d$pheno$y_cont_qtl_snp[1:5])
y
```

The values are $(-1,-3,-2,1,-2)$. Their sum is $-7$, so the mean is
$-7/5=-1.4$. Sorting gives $(-3,-2,-2,-1,1)$; the middle value and
median is $-2$. The value $-2$ occurs twice, more often than any other
value, so it is the mode.

Subtracting the mean gives the deviations
$(0.4,-1.6,-0.6,2.4,-0.6)$. Their squares sum to

$$
0.4^2+(-1.6)^2+(-0.6)^2+2.4^2+(-0.6)^2=9.2.
$$

The sample variance is therefore $9.2/(5-1)=2.3$, and the sample standard
deviation is $\sqrt{2.3}=1.517$ units. Under R's default quantile
convention, $Q_1=-2$ and $Q_3=-1$, giving an IQR of 1. The range is
$1-(-3)=4$.

```{r}
#| label: descriptive-manual-check
mode_value <- as.numeric(names(which.max(table(y))))
c(
  mean = mean(y),
  median = median(y),
  mode = mode_value,
  variance = var(y),
  sd = sd(y),
  range = diff(range(y)),
  iqr = IQR(y)
)
```

These summaries describe five rounded observations, not the full simulated
population. The negative centre is also relative to the arbitrary scale used
by the simulation; it does not mean that an individual has a biologically
negative trait. On all 200 raw records, the centre and spread differ because
rounding has been removed and many more individuals contribute.

```{r}
#| label: descriptive-full-data
summary(d$pheno$y_cont_qtl_snp)
c(
  variance = var(d$pheno$y_cont_qtl_snp),
  sd = sd(d$pheno$y_cont_qtl_snp)
)
```

For these 200 simulated records, the raw mean is approximately 0.411 and the
standard deviation is approximately 1.332. These are realised summaries of
one simulated dataset. They are not universal values for a species, trait, or
breeding programme.

## Outliers require investigation, not automatic deletion

An **[outlier](https://www.itl.nist.gov/div898/handbook/prc/section1/prc16.htm)**
is an observation that appears unusually distant from most of the data under
a stated comparison rule. The common boxplot rule flags values below
$Q_1-1.5\operatorname{IQR}$ or above
$Q_3+1.5\operatorname{IQR}$. It is a screening convention, not proof that
an observation is false.

An unusual phenotype could result from a transcription error, a different
measurement protocol, disease, environment, or genuine biology. Removing it
without checking can discard valuable genetic variation and make later
predictions appear more precise than they are. A breeder should examine the
record, metadata, and biological context before deciding whether correction,
exclusion, or retention is justified.

## Exercises

1. Round rows 6 to 10 of `y_cont_qtl_snp`. Calculate their mean, median,
   mode, sample variance, and sample standard deviation. Compare their centre
   and spread with rows 1 to 5.

::: {.callout-note collapse="true" title="Solution"}
The rounded values are $(-2,0,-1,-1,-2)$. Their mean is $-1.2$, the median
is $-1$, and the two modes are $-2$ and $-1$. The sample variance is 0.7,
and the sample standard deviation is approximately 0.837. Compared with rows
1 to 5, this group has a higher mean and a smaller standard deviation. The
comparison is descriptive and does not establish a genetic cause for the
difference.

```{r}
exercise_y <- round(d$pheno$y_cont_qtl_snp[6:10])
exercise_counts <- table(exercise_y)
exercise_modes <- as.numeric(
  names(exercise_counts[exercise_counts == max(exercise_counts)])
)
c(
  mean = mean(exercise_y),
  median = median(exercise_y),
  variance = var(exercise_y),
  sd = sd(exercise_y)
)
exercise_modes
```
:::

2. A hypothetical trait has values $(5,5,6,7,27)$. Compare the mean and
   median. Which measure better represents the four observations clustered
   between 5 and 7?

::: {.callout-note collapse="true" title="Solution"}
The mean is $(5+5+6+7+27)/5=10$, while the median is 6. The value 27 pulls
the mean above four of the five observations. The median better represents
the centre of that cluster, but the value 27 should be investigated rather
than deleted automatically.
:::

3. Two hypothetical groups both have mean 4. Group C is $(3,4,5)$, and
   Group D is $(0,4,8)$. Calculate their sample variances and explain what
   the comparison adds beyond their equal means.

::: {.callout-note collapse="true" title="Solution"}
For Group C, the squared deviations sum to 2, so the sample variance is
$2/(3-1)=1$. For Group D, they sum to 32, so the sample variance is
$32/(3-1)=16$. The equal means identify the same centre, while the
variances show that Group D is much more dispersed.
:::

Descriptive statistics treat the recorded values as fixed observations. The
next step is to describe uncertain outcomes before they are observed.
Probability and random variables provide that language and connect Mendelian
segregation to the statistical models used in genomic prediction.
