---
title: "Linear regression"
---

A breeder may notice that animals carrying more copies of one marker allele
also tend to have different phenotypes. How can that pattern be summarised,
and how much of the observed variation does it describe? **[Linear least
squares regression](https://www.itl.nist.gov/div898/handbook/pmd/section1/pmd141.htm)**
fits a line through paired observations by choosing coefficients that make
the squared discrepancies between the observations and the line as small as
possible. The fitted association can support prediction, but it does not show
that the marker is a causal QTL.

## A line separates the fitted pattern from unexplained variation

For individual $i$, let $y_i$ be a phenotype and $x_i$ the dosage of the
counted allele at one marker. The simple linear regression model is

$$
y_i = \beta_0 + \beta_1x_i + \varepsilon_i,
$$

where $\beta_0$ is the intercept, $\beta_1$ is the slope, and
$\varepsilon_i$ is the unobserved model error. The intercept is the
modelled mean phenotype when $x_i=0$. The slope is the modelled difference in
mean phenotype for a one-copy increase in dosage. Because a biallelic diploid
marker has dosages 0, 1, and 2, the fitted line compares those coding levels.
Changing which allele is counted changes the coding and therefore the
coefficient signs; it does not change the underlying observations.

After the coefficients have been estimated, the **fitted value** for
individual $i$ is

$$
\hat y_i = \hat\beta_0 + \hat\beta_1x_i.
$$

The **[residual](https://www.itl.nist.gov/div898/handbook/pmd/section4/pmd44.htm)**
is $e_i=y_i-\hat y_i$. A positive residual means that the observed phenotype
lies above the fitted line, while a negative residual means that it lies
below. Residuals contain variation not represented by this one-predictor
model. They should not be interpreted as purely environmental effects because
they can also contain omitted genetic effects, measurement error, nonlinear
structure, and other unmodelled sources of variation.

## Least squares chooses the slope and intercept

**Least squares** chooses $\hat\beta_0$ and $\hat\beta_1$ to minimise the
residual sum of squares

$$
\operatorname{RSS}=\sum_{i=1}^{n}e_i^2
=\sum_{i=1}^{n}(y_i-\hat\beta_0-\hat\beta_1x_i)^2,
$$

where $n$ is the number of individuals. When the marker dosage varies, the
least-squares slope can be written using the sample covariance and variance
introduced on the preceding pages:

$$
\hat\beta_1=\frac{s_{xy}}{s_x^2},
\qquad
\hat\beta_0=\bar y-\hat\beta_1\bar x.
$$

The common $n-1$ denominators in $s_{xy}$ and $s_x^2$ cancel. A positive
slope means that phenotype tends to increase with the counted-allele dosage
in the fitted sample; a negative slope means it tends to decrease. If all
individuals have the same dosage, then $s_x^2=0$ and the slope cannot be
estimated from those data.

The coefficient of determination compares the residual sum of squares with
the total sum of squares,

$$
R^2=1-\frac{\operatorname{RSS}}{\operatorname{TSS}},
\qquad
\operatorname{TSS}=\sum_{i=1}^{n}(y_i-\bar y)^2.
$$

For an ordinary least-squares model with an intercept, $R^2$ is the fraction
of sample phenotype variation described by the fitted predictors. It is a
training-data summary, not proof of a sound model or accurate prediction for
new individuals. The NIST model-validation guidance emphasises that **[$R^2$
alone is not enough](https://www.itl.nist.gov/div898/handbook/pmd/section4/pmd44.htm)**;
residual patterns and out-of-sample evaluation can reveal failures hidden by
one summary number.

## One marker in the recurring project data

The following example uses the first six animals in the simulated project
dataset. The phenotype is recorded on the simulation's arbitrary scale, and
`SNP007` is coded as counted-allele dosage 0, 1, or 2. These six observations
make the arithmetic inspectable, but they are not an independent random sample
from a breeding population.

```{r}
#| label: linear-regression-data
d <- readRDS("../../demo-data/main/demo_data.rds")
animals <- data.frame(
  phenotype = d$pheno$y_cont_qtl_snp[1:6],
  genotype_m1 = d$snp[1:6, "SNP007"]
)
animals
```

For this teaching slice, the sample covariance between dosage and phenotype
is approximately 0.2391 and the sample dosage variance is 0.7000. Their ratio
gives

$$
\hat\beta_1=\frac{0.2391}{0.7000}=0.3416.
$$

The mean dosage is 0.5 and the mean phenotype is approximately $-1.5616$, so
the intercept is

$$
\hat\beta_0=-1.5616-(0.3416)(0.5)=-1.7323.
$$

The fitted line is therefore $\hat y=-1.7323+0.3416x$. At dosage 2 it gives
$\hat y=-1.0492$. The observed phenotype for that animal is $-2.1248$, so
its residual is $-2.1248-(-1.0492)=-1.0756$.

R's [`lm()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/lm.html)
fits the same least-squares model and provides the fitted values and
residuals.

```{r}
#| label: linear-regression-fit
fit <- lm(phenotype ~ genotype_m1, data = animals)
cbind(animals, fitted = fitted(fit), residual = resid(fit))
coef(fit)
summary(fit)$r.squared
```

The fitted coefficients agree with the manual calculation. The residual sum
of squares is approximately 9.1479 and the total sum of squares is 9.5562,
giving $R^2=0.0427$. Thus the fitted line describes about 4.3% of phenotype
variation among these six records. That small value does not prove that
`SNP007` is biologically unimportant, just as a large value would not prove
causality. Sampling variation, family structure, linkage disequilibrium, and
other loci all affect a one-marker association.

## Assumptions determine what the line can support

Least squares can always produce coefficients when the design has sufficient
rank, but statistical interpretation needs assumptions. The model assumes
that the conditional mean of the phenotype is linear in the predictors and
that the unobserved model error has conditional mean zero. For the usual
standard errors and tests, errors from independent individuals are assumed to
be independent and to have constant conditional variance. A normal-error
assumption is additionally used for exact small-sample tests and intervals;
it is not required merely to calculate the least-squares line. Fitted
residuals estimate aspects of those unobserved errors for model checking, but
they are not the errors themselves. With an intercept, their sum is zero by
the least-squares calculation, and fitted residuals need not be statistically
independent.

These assumptions are conditional on the variables included in the model.
Relatives can have correlated model errors, management groups can shift the
phenotype mean, and a curved relationship can leave a pattern in residuals.
The six project animals come from the same simulated full-sib family, so this
example is a calculation rather than evidence for independent-population
inference. Residual plots, study design, and biological knowledge are needed
before the line is used beyond this slice.

Prediction also requires care. A fitted value describes the modelled mean at
a supplied predictor value. It does not predict every individual's phenotype
exactly, and extrapolation beyond the dosage range used for fitting has no
support from these observations. Prediction for new breeding candidates must
eventually be assessed on records that were not used to fit the model.

## Exercises

1. In the fitted project example, what do the intercept $-1.7323$ and slope
   $0.3416$ mean, and why is the slope not a causal QTL effect?

::: {.callout-note collapse="true" title="Solution"}
The intercept is the fitted mean phenotype at dosage 0. The slope is the
fitted phenotype difference for a one-copy increase in the counted allele, so
dosage 1 is fitted 0.3416 units above dosage 0. The coefficient summarises an
association in six simulated records. LD, family structure, and omitted loci
can generate that association, so the coefficient does not establish that
`SNP007` causes the trait difference.
:::

2. Use $\hat y=-1.7323+0.3416x$ to calculate the fitted value at dosage 2.
   If the observed phenotype is $-2.1248$, what is its residual?

::: {.callout-note collapse="true" title="Solution"}
The fitted value is $-1.7323+0.3416(2)=-1.0491$, with the difference from the
unrounded page result caused by rounding the coefficients. The residual is
$-2.1248-(-1.0491)=-1.0757$. Using the full coefficients gives fitted value
$-1.0492$ and residual $-1.0756$.
:::

3. A teaching sample has marker-phenotype covariance 0.36 and marker-dosage
   variance 0.80. Calculate the least-squares slope.

::: {.callout-note collapse="true" title="Solution"}
The slope is $s_{xy}/s_x^2=0.36/0.80=0.45$ phenotype units per additional
copy of the counted allele. This calculation assumes that the marker dosage
varies; if its variance were zero, the slope would be undefined.
:::

4. A fitted model has $\operatorname{RSS}=18$ and
   $\operatorname{TSS}=24$. Calculate $R^2$ and give one conclusion that the
   value cannot support by itself.

::: {.callout-note collapse="true" title="Solution"}
$R^2=1-18/24=0.25$, so the fitted predictors describe 25% of the sample
variation around the sample mean. This value alone cannot establish causality,
validate the residual assumptions, or show that prediction will be accurate
for new individuals.
:::

5. Fitted residuals grow more variable as fitted phenotypes increase. Which
   model-error assumption is doubtful, and why should the coefficient table
   be interpreted cautiously?

::: {.callout-note collapse="true" title="Solution"}
The constant conditional error-variance assumption is doubtful. Ordinary
least-squares coefficients can still be calculated, but standard errors and
intervals derived under constant variance may be inappropriate. The pattern
also indicates that a single summary such as $R^2$ is not a complete model
check.
:::

6. Explain why predicting at marker dosage 3 from this fitted line would be
   invalid for the stated biallelic diploid coding.

::: {.callout-note collapse="true" title="Solution"}
The stated dosage counts copies of one allele in a diploid individual, so its
possible values are only 0, 1, and 2. Dosage 3 is outside both the biological
coding and the fitted data range. Substituting it into the equation produces
a number, but not a supported biological prediction.
:::

One marker gives a readable line, but genomic datasets contain many correlated
markers. The next page extends the same model to a design matrix and explains
why regularisation is needed when ordinary least squares becomes unstable or
has no unique coefficient solution.
