---
title: "Linear mixed models"
---

Phenotypes reflect both systematic management conditions and differences among
animals. A **[linear mixed model](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4194460/)** treats those roles differently, allowing animal effects to share information through an explicit covariance structure.

## Fixed and random effects have different modelling roles

A fixed effect represents included comparisons, such as observed management
groups. A random effect represents values treated as draws from a distribution,
such as animal effects in a breeding population. The distinction is about the
model target and covariance assumption, not whether a quantity was measured.

For $n$ phenotype records and $q$ animal effects,

$$
\mathbf y=\mathbf X\boldsymbol\beta+\mathbf Z\mathbf u+
\boldsymbol\varepsilon.
$$

$\mathbf y$ is $n\times1$; $\mathbf X$ is the $n\times k$ fixed-effect design
matrix; $\boldsymbol\beta$ is $k\times1$; $\mathbf Z$ is the $n\times q$
incidence matrix; $\mathbf u$ is $q\times1$; and
$\boldsymbol\varepsilon$ is $n\times1$. A simple animal model assumes

$$
\mathbf u\sim N(\mathbf0,\mathbf G\sigma_u^2),\qquad
\boldsymbol\varepsilon\sim N(\mathbf0,\mathbf I_n\sigma_e^2).
$$

Here $\mathbf G$ is a $q\times q$ relationship covariance template,
$\sigma_u^2$ and $\sigma_e^2$ are variance components, and random effects and
residuals are assumed independent. Off-diagonal entries of $\mathbf G$ allow
related animals to share information. Pedigree and genomic matrices are
introduced later.

## Covariance produces partial pooling

The phenotype covariance is

$$
\operatorname{Var}(\mathbf y)=\mathbf Z\mathbf G\mathbf Z^\mathsf T
\sigma_u^2+\mathbf I_n\sigma_e^2.
$$

An animal with little information is pulled more strongly toward the modelled
population mean than an animal with informative records or relatives. This
**partial pooling** is model-based shrinkage, not evidence that an animal has
no genetic merit. Its strength depends on
$\lambda=\sigma_e^2/\sigma_u^2$: larger residual variation makes individual
records less decisive. Estimating variance components, including REML, comes
later.

## A four-animal mixed-model equation

This simulated teaching example has one intercept, so $\mathbf X$ is
$4\times1$. It assigns an independent random effect to each record, so
$\mathbf Z=\mathbf I_4$ is $4\times4$, and supplies $\lambda=2$. It is
preparation for relationship matrices, not a pedigree or genomic analysis.

```{r}
#| label: mixed-model-data
#| code-fold: true
d <- readRDS("../../demo-data/main/demo_data.rds")
y <- round(d$pheno$y_cont_qtl_snp[1:4])
X <- matrix(1, nrow = 4, ncol = 1)
Z <- diag(4)
lambda <- 2
y
```

With supplied variance components, solve

$$
\begin{bmatrix}
\mathbf X^\mathsf T\mathbf X & \mathbf X^\mathsf T\mathbf Z\\
\mathbf Z^\mathsf T\mathbf X & \mathbf Z^\mathsf T\mathbf Z+\lambda\mathbf I_4
\end{bmatrix}
\begin{bmatrix}\hat{\boldsymbol\beta}\\\hat{\mathbf u}\end{bmatrix}
=
\begin{bmatrix}\mathbf X^\mathsf T\mathbf y\\\mathbf Z^\mathsf T\mathbf y\end{bmatrix}.
$$

```{r}
#| label: mixed-model-solution
#| code-fold: true
lhs <- rbind(
  cbind(crossprod(X), crossprod(X, Z)),
  cbind(crossprod(Z, X), crossprod(Z) + lambda * diag(4))
)
rhs <- rbind(crossprod(X, y), crossprod(Z, y))
solution <- solve(lhs, rhs)
solution
```

The intercept is $-1.25$ and predicted animal deviations are approximately
0.083, $-0.583$, $-0.250$, and 0.750. They are shrunk toward zero relative to
raw deviations because $\lambda=2$ penalises large random effects. They are
not observed breeding values, and this toy model does not use simulated truth.

## BLUP and prediction uncertainty

With known or estimated covariance parameters, $\hat{\mathbf u}$ is the
**[best linear unbiased prediction](https://doi.org/10.2307/2529430)** (BLUP)
under the stated model. “Best” means minimum prediction-error variance among
linear unbiased predictors in that model. It does not mean causal, biologically
true, or guaranteed best in another population.

Prediction error variance describes uncertainty in
$\hat{\mathbf u}-\mathbf u$, not variation among predicted values. It depends
on record number, residual noise, fixed-effect design, and relationships. The
next genomic-prediction pages replace this independent toy covariance with
pedigree and genomic relationships, then explain REML estimation of the
variance components controlling shrinkage.

## Exercises

1. With 120 records, three fixed-effect columns, and 150 animal effects, what are the dimensions of $\mathbf X$ and $\mathbf Z$?

::: {.callout-note collapse="true" title="Solution"}
$\mathbf X$ is $120\times3$ and $\mathbf Z$ is $120\times150$.
:::

2. What does a positive off-diagonal entry in $\mathbf G$ allow the model to do?

::: {.callout-note collapse="true" title="Solution"}
It models the corresponding random effects as positively correlated, allowing
one animal's information to contribute to prediction of the other's effect.
:::

3. Holding other quantities fixed, what generally happens to shrinkage when $\lambda=\sigma_e^2/\sigma_u^2$ increases?

::: {.callout-note collapse="true" title="Solution"}
Individual deviations are usually pulled more strongly toward their modelled
mean because residual variation is larger relative to animal-effect variation.
:::
