---
title: "Multiple regression and regularisation"
---

A breeding programme usually genotypes hundreds or thousands of markers, not
one. How can their information enter one prediction model when markers are
correlated and may outnumber the phenotyped individuals? **[Multiple linear
regression](https://www.itl.nist.gov/div898/handbook/pmd/section1/pmd141.htm)**
places all predictors in one design matrix. **[Ridge
regression](https://doi.org/10.1080/00401706.1970.10488634)** adds a penalty
that stabilises the fitted marker coefficients by shrinking them toward zero.

## One design matrix holds several predictors

For $n$ individuals and $p$ markers, let $\mathbf y$ be the $n\times1$
phenotype vector. Let $\mathbf X$ be the $n\times(p+1)$ design matrix whose
first column is an intercept and whose remaining $p$ columns are marker
dosages. The coefficient vector $\boldsymbol\beta$ is $(p+1)\times1$, and the
unobserved model-error vector $\boldsymbol\varepsilon$ is $n\times1$. The
model is

$$
\mathbf y=\mathbf X\boldsymbol\beta+\boldsymbol\varepsilon.
$$

The fitted vector $\hat{\mathbf y}=\mathbf X\hat{\boldsymbol\beta}$ is
$n\times1$. Each marker coefficient is a **partial effect**: it is the fitted
change associated with that marker while the other columns in the design
matrix are held fixed. This conditional meaning differs from fitting each
marker separately. It also remains statistical rather than molecular. A
nonzero coefficient can reflect LD with a causal locus, shared family
structure, or adjustment for correlated predictors.

Ordinary least squares minimises

$$
\lVert\mathbf y-\mathbf X\boldsymbol\beta\rVert_2^2,
$$

where the squared Euclidean norm is the residual sum of squares. If
$\mathbf X$ has full column rank, the algebraic solution can be written

$$
\hat{\boldsymbol\beta}_{\mathrm{OLS}}
=(\mathbf X^\mathsf T\mathbf X)^{-1}
\mathbf X^\mathsf T\mathbf y.
$$

Software normally solves the corresponding linear system instead of forming
the inverse explicitly. If columns repeat the same information exactly,
$\mathbf X$ is rank deficient and the inverse does not exist. Coefficients are
then not uniquely identified without an additional constraint.

## Correlated markers make coefficients unstable

**[Collinearity](https://itl.nist.gov/div898/software/dataplot/refman1/auxillar/regrdiag.htm)**
occurs when one predictor column is closely approximated by other columns.
Linked markers often carry similar information, so small changes in the
sampled animals or phenotypes can shift the fitted effect among those markers.
The combined predictions may change less than the individual coefficients,
but interpreting one unstable coefficient as a distinct locus effect is
unsafe.

The dimensional problem is even clearer when $p>n$. The full project data
contain 200 phenotyped offspring and 500 SNP columns. Adding an intercept
would give a $200\times501$ design matrix, whose rank cannot exceed 200.
Ordinary least squares therefore cannot identify a unique coefficient for
every column from these records alone.

A flexible model can also **overfit** by adapting to noise and sample-specific
patterns. Its training error may be small because the same observations were
used to choose the coefficients. Test error measures prediction on records
held out from fitting and can increase even while training error decreases.
This is why model complexity and tuning choices must be assessed with a split
that represents the intended candidate population.

## Ridge regression trades some fit for stability

Ridge regression estimates coefficients by minimising

$$
\lVert\mathbf y-\mathbf X\boldsymbol\beta\rVert_2^2
+\lambda\sum_{j=1}^{p}\beta_j^2,
$$

where $\lambda\geq0$ is the penalty strength and the sum excludes the
intercept. With

$$
\mathbf P=\operatorname{diag}(0,1,\ldots,1),
$$

an $(p+1)\times(p+1)$ penalty matrix, the ridge solution is

$$
\hat{\boldsymbol\beta}_{\mathrm{ridge}}
=(\mathbf X^\mathsf T\mathbf X+\lambda\mathbf P)^{-1}
\mathbf X^\mathsf T\mathbf y.
$$

The zero in the first diagonal position leaves the intercept unpenalised.
Positive entries penalise large marker coefficients. At $\lambda=0$, the
criterion reduces to ordinary least squares when its solution is unique. As
$\lambda$ increases, the penalty increasingly favours a smaller coefficient
vector. Its overall size moves toward zero, although an individual coefficient
need not decrease monotonically when predictors are correlated. Ridge usually
does not set coefficients exactly to zero, so it performs shrinkage rather
than automatic marker selection or causal-locus discovery.

This added bias can reduce variation in fitted coefficients and improve
prediction on new data. The **bias-variance tradeoff** describes that exchange:
weak penalisation can leave a high-variance model, whereas excessive
penalisation can erase useful signal and cause underfitting. The value of
$\lambda$ should therefore be chosen using training data and an appropriate
validation design, never from final test performance.

The penalty also depends on predictor units. A coefficient attached to a
large-scale predictor can be penalised differently from one attached to a
small-scale predictor for the same predictive change. Genomic models commonly
centre and sometimes scale marker columns using values derived from the
training data. Those operations are developed later in the genomic-data
encoding page. The teaching calculation below uses raw 0, 1, and 2 dosages so
that its matrix can be inspected directly.

## A three-marker ridge calculation

The recurring teaching view contains six simulated individuals and three
markers. The design matrix $\mathbf X$ has dimensions $6\times4$: one
intercept column followed by `SNP007`, `SNP008`, and `SNP017`. The phenotype
vector $\mathbf y$ is $6\times1$, the coefficient vector is $4\times1$, and
$\mathbf P$ is $4\times4$.

```{r}
#| label: ridge-data
d <- readRDS("../../demo-data/main/demo_data.rds")
marker_view <- d$snp[
  1:6,
  c("SNP007", "SNP008", "SNP017"),
  drop = FALSE
]
X <- cbind(intercept = 1, marker_view)
y <- d$pheno$y_cont_qtl_snp[1:6]
X
y
```

For the supplied teaching value $\lambda=2$, the penalty matrix has diagonal
$(0,1,1,1)$. The calculation uses `solve(A, b)` to solve the system
$\mathbf A\mathbf b=\mathbf c$ directly.

```{r}
#| label: ridge-fit
lambda <- 2
penalty <- diag(c(0, rep(1, ncol(marker_view))))
ridge_beta <- solve(
  crossprod(X) + lambda * penalty,
  crossprod(X, y)
)
ridge_fitted <- drop(X %*% ridge_beta)
ridge_beta
ridge_fitted
```

The fitted intercept is approximately $-1.8728$. The coefficients for
`SNP007`, `SNP008`, and `SNP017` are approximately 0.1274, $-0.3216$, and
0.6186. These are coefficients for one raw-dosage, six-animal teaching model.
They are not estimates of causal effects, and $\lambda=2$ was supplied rather
than selected from validation data.

Comparing several penalty strengths makes the shrinkage visible.

```{r}
#| label: ridge-penalty-comparison
ridge_coefficient <- function(lambda) {
  solve(
    crossprod(X) + lambda * penalty,
    crossprod(X, y)
  )
}

coefficient_comparison <- cbind(
  lambda_0 = ridge_coefficient(0),
  lambda_2 = ridge_coefficient(2),
  lambda_20 = ridge_coefficient(20)
)
coefficient_comparison
```

As $\lambda$ rises from 0 to 20, the sum of squared marker coefficients falls
from approximately 1.388 to 0.026. Over the same training records, RSS rises
from approximately 5.264 to 8.439. The `SNP007` coefficient first changes
from 0.0944 to 0.1274 and then falls to 0.0451, illustrating why shrinkage is
best understood through the penalised vector rather than a monotonic path for
every correlated predictor. Ridge accepts a worse unpenalised training fit in
exchange for a smaller coefficient vector. These training values do not
identify which $\lambda$ predicts new candidates best.

## The bridge to genomic prediction

The same high-dimensional problem appears when genome-wide marker effects are
estimated jointly. Ridge supplies every penalised marker coefficient with the
same quadratic penalty after coding and scaling choices have been fixed. Its
shrinkage principle connects directly to genomic best linear unbiased
prediction, although that connection requires a covariance model and variance
components that have not yet been introduced.

Ridge also does not solve every modelling problem. A linear predictor can
miss dominance, epistasis, nonlinear trends, or management effects if the
design matrix omits them. Correlated relatives can violate an independent-
error assumption, and preprocessing can leak information across a data
split. Regularisation controls coefficient size under a stated model; it does
not repair the sampling design or supply a biological mechanism.

## Exercises

1. A study has 200 individuals and 500 marker columns. After adding an
   intercept, what are the dimensions of $\mathbf X$ and
   $\boldsymbol\beta$, and why can ordinary least squares not identify a
   unique coefficient for every column?

::: {.callout-note collapse="true" title="Solution"}
$\mathbf X$ is $200\times501$, and $\boldsymbol\beta$ is $501\times1$. The
rank of a matrix with 200 rows cannot exceed 200, so the 501 columns cannot be
linearly independent. The ordinary least-squares coefficient vector is
therefore not unique without an additional constraint.
:::

2. In a model containing three correlated markers, what does one marker's
   coefficient describe, and why should it not automatically be called that
   marker's causal effect?

::: {.callout-note collapse="true" title="Solution"}
It describes the fitted change associated with that marker while the other
included columns are held fixed. Correlation among marker columns can move
shared information between their coefficients, and a marker can tag a causal
locus through LD. The partial coefficient is therefore a model-dependent
association, not direct proof of a molecular effect.
:::

3. Suppose one design-matrix marker column is an exact copy of another. What
   happens to rank and the ordinary least-squares solution?

::: {.callout-note collapse="true" title="Solution"}
The copied column adds no independent direction, so the design matrix loses
full column rank. The cross-product matrix is singular and the separate
ordinary least-squares coefficients for those duplicate columns are not
uniquely identified.
:::

4. For the $\lambda=2$ teaching fit, RSS is 5.9205 and the sum of squared
   marker coefficients is 0.5023. Calculate the ridge objective.

::: {.callout-note collapse="true" title="Solution"}
The objective is
$5.9205+2(0.5023)=6.9251$. The intercept is excluded from the squared-
coefficient sum because the first diagonal entry of $\mathbf P$ is zero.
:::

5. When $\lambda$ rises from 0 to 20 in the project slice, the squared marker
   coefficient sum falls while training RSS rises. Explain why this does not
   show that either value of $\lambda$ will predict new candidates better.

::: {.callout-note collapse="true" title="Solution"}
The penalised fit deliberately trades training fit for smaller coefficients.
Training RSS favours the model fitted most closely to the same records, while
the smaller coefficient norm indicates stronger shrinkage. Prediction must be
compared on appropriate held-out records because either weak shrinkage can
overfit or strong shrinkage can underfit.
:::

6. Why is the intercept left unpenalised, and what preprocessing issue must be
   addressed before comparing penalties across predictors measured on
   different scales?

::: {.callout-note collapse="true" title="Solution"}
The intercept represents the baseline level rather than a marker contribution,
so $\mathbf P$ assigns it a zero penalty. Because a quadratic penalty acts on
coefficient magnitude, predictor scale changes how strongly equivalent fitted
changes are penalised. Predictors should be centred and, when the model calls
for it, scaled using quantities estimated from training data and then applied
unchanged to validation or test data.
:::

Multiple regression distinguishes observations, parameters, fitted values,
and residuals, but it has not yet explained how parameters are estimated from
a probability model. The next page develops likelihood, estimators, and
identifiability before linear mixed models add covariance among individuals.
