---
title: "Probability distributions"
---

The previous page assigned probabilities to allele-dosage values from one
Mendelian cross. A genomic prediction analysis contains several kinds of
uncertain quantity: a disease outcome may be binary, a genotype is discrete,
and a residual or marker effect is usually represented on a continuous scale.
No single probability model suits all of them. A **[probability
distribution](https://www.itl.nist.gov/div898/handbook/eda/section3/eda36.htm)**
specifies the possible values of a random variable and how probability is
allocated across those values.

## Discrete and continuous distributions

A discrete random variable takes separate values that can be listed or
counted. Its **[probability mass function](https://en.wikipedia.org/wiki/Probability_mass_function)**,
written $P(X=x)$, assigns a probability to each possible value. The masses
are nonnegative and sum to one.

A continuous random variable can take values throughout an interval. Its
**[probability density function](https://en.wikipedia.org/wiki/Probability_density_function)**
$f(x)$ describes how probability is concentrated. Probability is an area
under the density curve:

$$
P(a\leq X\leq b)=\int_a^b f(x)\,dx.
$$

The height $f(x)$ is not the probability of observing exactly $x$. For a
continuous model, $P(X=x)=0$ for any single point, even where the density is
high. A density may also exceed one as long as its total area is one.

A distribution's **[parameters](https://www.itl.nist.gov/div898/handbook/eda/section3/eda364.htm)**
control properties such as location, spread, or shape. Their meanings depend
on the distribution. For example, $\pi$ is a success probability in a
Bernoulli model, whereas $\mu$ and $\sigma$ locate and scale a normal
model.

## Bernoulli distribution: one binary outcome

A **[Bernoulli distribution](https://en.wikipedia.org/wiki/Bernoulli_distribution)**
describes one outcome coded as 1 for success and 0 for failure. If $X$ is a
disease indicator and $\pi$ is the probability of disease under the model,

$$
P(X=x)=\pi^x(1-\pi)^{1-x}, \qquad x\in\{0,1\}.
$$

Its expectation is $E(X)=\pi$ and its variance is $\pi(1-\pi)$. Calling one
outcome a success is only a mathematical convention. A disease case can be
coded 1 without being desirable.

The Bernoulli model assumes the probability $\pi$ is appropriate for the
individual and setting being considered. If disease risk differs with family,
environment, or management, one common value of $\pi$ may be inadequate.

## Binomial distribution: a count of binary outcomes

A **[binomial distribution](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Binomial.html)**
counts successes in $m$ Bernoulli trials. If the trials are independent and
share a constant success probability $\pi$, then

$$
P(X=x)={m\choose x}\pi^x(1-\pi)^{m-x},
\qquad x=0,1,\ldots,m.
$$

Suppose five offspring independently have disease probability 0.2 under a
hypothetical model. The probability that exactly two are affected is

$$
{5\choose2}(0.2)^2(0.8)^3
=10(0.04)(0.512)=0.2048.
$$

The model can also describe an allele count when allele copies are treated as
independent draws with a common frequency. That assumption may fail in
inbred, structured, or selected populations. Related offspring may likewise
have correlated risks, so a binomial calculation is not automatically valid
for every family.

## Categorical distribution: one of several classes

A **[categorical distribution](https://en.wikipedia.org/wiki/Categorical_distribution)**
assigns one outcome to one of $K$ unordered classes. If class probabilities
are $\pi_1,\ldots,\pi_K$, then each $\pi_k\geq0$ and
$\sum_{k=1}^{K}\pi_k=1$. Genotypes such as $AA$, $Aa$, and $aa$ may
be represented as categories when no numerical dosage structure is intended.

Later, BayesR will use a categorical distribution for a latent marker-effect
class. The class label says which variance component generated an effect; the
integers used to store labels do not imply that one class is numerically twice
another.

## Normal distribution: a symmetric working model

The **[normal distribution](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Normal.html)**
with mean $\mu$ and variance $\sigma^2$ has density

$$
f(x)=\frac{1}{\sigma\sqrt{2\pi}}
\exp\left[-\frac{(x-\mu)^2}{2\sigma^2}\right],
\qquad -\infty<x<\infty.
$$

Its symmetric bell shape places the mean, median, and mode at $\mu$, while
$\sigma$ controls spread. Linear models often assume that residuals or
polygenic effects follow a normal distribution. This is a model assumption,
not a claim that every phenotype or marker effect in nature is normally
distributed. Diagnostics and knowledge of how the trait was measured remain
necessary.

## Student-t distribution: more probability in the tails

The **[Student-t distribution](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/TDist.html)**
is symmetric like the normal distribution but has heavier tails. Its degrees
of freedom parameter $\nu$ controls tail weight: smaller $\nu$ places more
probability far from the centre, and the distribution approaches a normal
shape as $\nu$ increases.

Heavy-tailed distributions can serve as models or priors when most marker
effects are near zero but a few larger effects remain plausible. Choosing a
Student-t model does not establish that the observed marker effects truly
follow that distribution. It states which effect sizes the analysis regards
as plausible before or while learning from data.

```{r}
#| label: fig-normal-student
#| fig-cap: "Normal and Student-t densities with centre 0 and scale 1. The dashed Student-t curve with four degrees of freedom lies above the normal curve in the distant tails; the two curves do not have equal variance under this parameterisation."
#| fig-alt: "A solid normal density and dashed Student-t density are centred at zero. The Student-t curve has a lower central peak and remains higher than the normal curve near minus four and four."
#| fig-width: 7
#| fig-height: 4.5
x <- seq(-4, 4, length.out = 401)
plot(
  x, dnorm(x), type = "l", lwd = 2,
  xlab = "Value", ylab = "Density"
)
lines(x, dt(x, df = 4), lwd = 2, lty = 2)
legend(
  "topright", c("Normal", "Student-t, df = 4"),
  lty = c(1, 2), lwd = 2, bty = "n"
)
```

The curves share a centre and scale parameter. They do not share a variance:
the standard normal variance is 1, whereas a Student-t variable with
$\nu=4$ and scale 1 has variance $\nu/(\nu-2)=2$. The comparison isolates
the practical consequence of heavier tails without asserting a universal
marker-effect distribution.

## Chi-squared distribution: nonnegative squared variation

A **[chi-squared distribution](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Chisquare.html)**
with $\nu$ degrees of freedom describes the sum of squares of $\nu$
independent standard normal variables:

$$
X=Z_1^2+\cdots+Z_\nu^2,
\qquad Z_k\sim N(0,1).
$$

It has support on nonnegative values and is right-skewed for small $\nu$.
Chi-squared distributions appear in variance inference and help motivate
some prior distributions for variance components. Their use depends on
normality and independence assumptions in the derivation; a nonnegative
biological measurement is not chi-squared merely because its values cannot be
negative.

## Calculate masses and densities in R

R's distribution functions use a consistent prefix: `d` evaluates a mass or
density, `p` evaluates a cumulative probability, `q` returns a quantile, and
`r` generates random values. The [R distribution
documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Distributions.html)
lists the available families.

```{r}
#| label: distribution-values
c(
  binomial_mass = dbinom(2, size = 5, prob = 0.2),
  normal_density = dnorm(0, mean = 0, sd = 1.5),
  student_density = dt(0, df = 4),
  chi_squared_density = dchisq(2, df = 3)
)
```

The binomial result 0.2048 is a probability mass. The other three numbers are
density heights and should not be read as probabilities at those exact
points. An interval probability for a normal variable requires subtraction
of cumulative probabilities. For example, if $Y\sim N(0,1.5^2)$, then

```{r}
#| label: normal-interval
pnorm(1, mean = 0, sd = 1.5) - pnorm(-1, mean = 0, sd = 1.5)
```

This calculation gives the model probability that $Y$ falls between -1 and
1. It is valid only to the extent that the specified normal distribution is a
useful model for the quantity being studied.

## Exercises

1. Three offspring each have disease probability 0.1 under an independent,
   constant-risk model. What is the probability that at least one is affected?

::: {.callout-note collapse="true" title="Solution"}
It is simpler to calculate the complement. The probability that none is
affected is $(1-0.1)^3=0.9^3=0.729$. Therefore, the probability that at
least one is affected is $1-0.729=0.271$.
:::

2. A categorical prior assigns marker-effect class probabilities 0.80, 0.15,
   and 0.05. What condition must these values satisfy, and what is the
   probability that a marker belongs to either of the last two classes?

::: {.callout-note collapse="true" title="Solution"}
The probabilities must be nonnegative and sum to one. They sum to
$0.80+0.15+0.05=1$. Because the classes are mutually exclusive, the
probability of either of the last two is $0.15+0.05=0.20$.
:::

3. For a standard normal variable $Z$, use R to calculate
   $P(-1\leq Z\leq1)$. Explain why `dnorm(1)` does not answer the question.

::: {.callout-note collapse="true" title="Solution"}
The interval probability is
`pnorm(1) - pnorm(-1)`, approximately 0.683. `dnorm(1)` is the height of the
density at 1, not an area over the interval and not the probability of the
single value 1.
:::

4. A disease indicator is coded 0 or 1. Which distribution models one such
   outcome, and what assumption is needed before using one common probability?

::: {.callout-note collapse="true" title="Solution"}
A Bernoulli distribution models one binary outcome. Its probability parameter
must be appropriate for the individual and setting; family, environment, or
management differences can make one common value inadequate.
:::

5. Why might a Student-t model assign more prior plausibility to large marker
   effects than a normal model with the same centre and scale?

::: {.callout-note collapse="true" title="Solution"}
The Student-t distribution has heavier tails, especially at smaller degrees
of freedom, so it assigns more probability far from its centre. This is a
model choice, not proof that observed marker effects follow that distribution.
:::

6. Which R prefix returns a cumulative probability, and why is a chi-squared
   variable not negative under the article's construction?

::: {.callout-note collapse="true" title="Solution"}
The `p` prefix returns a cumulative probability. A chi-squared variable is a
sum of squared standard normal variables, and squares cannot be negative.
:::

A probability distribution describes variation in repeated outcomes under a
model. A dataset supplies only one realised sample. The next page examines
how estimates change across samples, how standard errors quantify that
variation, and what a confidence interval means.
