---
title: "Sampling and statistical uncertainty"
---

A genomic prediction model learns from a finite training sample, but breeders
usually want its conclusions to apply to other individuals. A sample mean,
marker effect, or prediction accuracy would change if a different set of
individuals were measured. Statistical uncertainty describes this
sample-to-sample variation. More records often improve precision, but a large
sample from the wrong families, environments, or generations can still answer
the wrong breeding question.

## Population depends on the question

A biological population might mean all members of a herd, strain, or species.
A **[statistical population](https://en.wikipedia.org/wiki/Statistical_population)**
is the collection or probability process about which an analysis seeks a
conclusion. The target population must be stated more precisely than either
label. It might consist of selection candidates from the same breeding
programme next year, including the environments and management conditions in
which they will be evaluated.

A **[sample](https://en.wikipedia.org/wiki/Sampling_(statistics))** is the set
of individuals actually observed. In genomic prediction, the training sample
contains the phenotypes and genotypes used to estimate a model. A sample is
representative when its selection mechanism and biological structure retain
the features relevant to the target population. Sample size alone cannot
guarantee this. Two thousand close relatives from one environment may provide
less information about an unrelated population than a smaller, deliberately
diverse sample.

The examples below use the 200 simulated project individuals as an empirical
population. Resampling from those records is a teaching device. It does not
recreate the pedigree, meiosis, recombination, or environmental process that
generated the original dataset.

## A statistic varies across possible samples

A **[parameter](https://en.wikipedia.org/wiki/Statistical_parameter)** such as
a target-population mean is fixed but unknown under the frequentist framework
used here. A **[statistic](https://en.wikipedia.org/wiki/Statistic)** is
calculated from a sample and therefore changes with the selected observations.
The **[sampling
distribution](https://en.wikipedia.org/wiki/Sampling_distribution)** of a
statistic describes its values over repeated samples drawn by the same
procedure.

This differs from the phenotype distribution. One distribution contains
individual phenotype values. The sampling distribution of the mean contains
one mean from each possible sample. If independent observations come from a
population with mean $\mu$, variance $\sigma^2$, and sample size $n$, then

$$
E(\bar Y)=\mu,
\qquad
\operatorname{Var}(\bar Y)=\frac{\sigma^2}{n}.
$$

The standard deviation of a sampling distribution is called a **[standard
error](https://en.wikipedia.org/wiki/Standard_error)**. For the sample mean,

$$
\operatorname{SE}(\bar Y)=\frac{\sigma}{\sqrt n}.
$$

Because the population standard deviation $\sigma$ is usually unknown, it is
often replaced by the sample standard deviation $s$. The square-root relation
has an important consequence: doubling sample size does not halve the
standard error. Under the same independent-sampling assumptions, four times
as many observations are needed to halve it.

## Manual example

The first five rounded project phenotypes from the descriptive-statistics
page have mean $\bar y=-1.4$ and sample standard deviation $s=1.517$. Treating
them temporarily as independent observations from a normal population gives

$$
\widehat{\operatorname{SE}}(\bar y)
=\frac{1.517}{\sqrt5}=0.678.
$$

If the population variance is estimated from these same five observations, a
two-sided 95% Student-t interval for the population mean is

$$
\bar y \mathbin{\pm} t_{0.975,4}\frac{s}{\sqrt n}
=-1.4 \mathbin{\pm} 2.776(0.678)
=(-3.283,\ 0.483).
$$

This interval is wide because five observations provide little information.
It is only an arithmetic illustration. The first five rows were not obtained
through a documented random-sampling design, and related individuals would
violate the independence assumption. Those limitations prevent the interval
from supporting a population claim here.

```{r}
#| label: small-sample-interval
d <- readRDS("../../demo-data/main/demo_data.rds")
y <- round(d$pheno$y_cont_qtl_snp[1:5])
standard_error <- sd(y) / sqrt(length(y))
critical_value <- qt(0.975, df = length(y) - 1)
confidence_interval <- mean(y) + c(-1, 1) * critical_value * standard_error

c(
  mean = mean(y),
  standard_error = standard_error,
  lower = confidence_interval[1],
  upper = confidence_interval[2]
)
```

## Resampling the project phenotypes

The following simulation repeatedly samples with replacement from the 200
raw continuous phenotypes. Each repeat produces one sample mean. The random
seed makes the teaching result reproducible.

```{r}
#| label: bootstrap-means
set.seed(20260910)
project_population <- d$pheno$y_cont_qtl_snp
sample_means_10 <- replicate(
  2000,
  mean(sample(project_population, 10, replace = TRUE))
)
sample_means_100 <- replicate(
  2000,
  mean(sample(project_population, 100, replace = TRUE))
)

c(
  phenotype_sd = sd(project_population),
  mean_sd_n10 = sd(sample_means_10),
  mean_sd_n100 = sd(sample_means_100)
)
```

The 200 source phenotypes have standard deviation approximately 1.332. Across
the resamples, means based on 10 records have standard deviation approximately
0.417, while means based on 100 records have standard deviation approximately
0.135. Increasing sample size narrows the sampling distribution of the mean;
it does not change the spread of the 200 source phenotypes.

```{r}
#| label: fig-sampling-distributions
#| fig-cap: "Sampling distributions from 2,000 resamples of the simulated project phenotypes. Both panels use the same horizontal scale. Means based on 100 observations cluster more tightly around the source-data mean than means based on 10 observations."
#| fig-alt: "Two stacked histograms share a horizontal scale. The histogram for samples of 100 is much narrower around the vertical source-mean line than the histogram for samples of 10."
#| fig-width: 7
#| fig-height: 7
plot_limits <- range(sample_means_10, sample_means_100)
old_par <- par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
hist(
  sample_means_10, breaks = 30, xlim = plot_limits,
  main = "Sample size 10", xlab = "Sample mean"
)
abline(v = mean(project_population), lwd = 2, lty = 2)
hist(
  sample_means_100, breaks = 30, xlim = plot_limits,
  main = "Sample size 100", xlab = "Sample mean"
)
abline(v = mean(project_population), lwd = 2, lty = 2)
par(old_par)
```

The observed ratio of the two simulation standard deviations is approximately
3.10, close to the theoretical ratio
$\sqrt{100/10}=\sqrt{10}\approx3.16$. Small differences are expected because
the simulation uses a finite number of resamples.

## Confidence intervals describe a procedure

A **[confidence interval](https://www.itl.nist.gov/div898/handbook/eda/section3/eda352.htm)**
combines an estimate with its sampling uncertainty. A 95% confidence procedure
is designed so that, under its assumptions, about 95% of intervals calculated
from repeated samples cover the fixed population parameter. Once one interval
has been calculated, it either covers that parameter or it does not. The
frequentist statement is not that there is a 95% probability that this one
fixed interval contains the fixed parameter.

For an independent normal sample with unknown variance, the interval for a
mean uses a Student-t critical value:

$$
\bar y \mathbin{\pm} t_{1-\alpha/2,n-1}\frac{s}{\sqrt n},
$$

where $1-\alpha$ is the confidence level. Dependence among relatives,
selective sampling, strong skew in a small sample, or unmodelled population
structure can make this simple interval inappropriate. A genomic analysis
must align its uncertainty calculation with its actual sampling and model
structure.

## Bias and precision are different

A **[statistical estimator](https://en.wikipedia.org/wiki/Estimator)**
$\hat\theta$ targets a parameter $\theta$. Its **[statistical
bias](https://en.wikipedia.org/wiki/Bias_of_an_estimator)** is

$$
\operatorname{Bias}(\hat\theta)=E(\hat\theta)-\theta.
$$

An unbiased estimator is correct on average over repeated samples, but one
realised estimate can still be far from the target.
**[Precision](https://en.wikipedia.org/wiki/Accuracy_and_precision)**
describes the spread of the estimator's sampling distribution. Larger samples
often improve precision, but they do not repair bias caused by systematically
missing a subpopulation or measuring the trait differently in training and
deployment groups.

This distinction matters in breeding. A precisely estimated mean for one
family may still be a biased estimate of the mean for all candidate families.
Likewise, random cross-validation within a training population can estimate
one prediction task precisely while giving an optimistic answer for future or
unrelated candidates.

## Exercises

1. Independent observations have standard deviation 10. Compare the standard
   error of the mean for sample sizes 25 and 100.

::: {.callout-note collapse="true" title="Solution"}
For $n=25$, the standard error is $10/\sqrt{25}=2$. For $n=100$, it is
$10/\sqrt{100}=1$. Quadrupling the sample size halves the standard error under
the independence assumption.
:::

2. A 95% confidence interval calculated from one sample is $(12,18)`. Is it
   correct to say that the fixed population mean has probability 0.95 of lying
   in this realised interval?

::: {.callout-note collapse="true" title="Solution"}
No, not under the frequentist interpretation used here. The interval is the
random object before sampling; the parameter is fixed. The 95% refers to the
long-run coverage of the method over repeated samples generated under its
assumptions.
:::

3. A training sample contains many records but only from one family. Explain
   why increasing the number of records may narrow a standard error without
   making the sample representative of unrelated families.

::: {.callout-note collapse="true" title="Solution"}
Additional records can reduce sample-to-sample variation for the family that
was sampled. They do not add the allele frequencies, linkage patterns, or
environments present in unobserved families. The estimate may therefore be
precise for the sampled family but systematically unrepresentative of the
broader target population.
:::

Sampling uncertainty concerns the behaviour of one statistic across possible
samples. The next mathematical topic examines how two variables vary
together. Covariance and correlation provide the first tools for describing
connections among phenotypes, breeding values, and later predictions.
