---
title: "Statistical models and estimation"
---

A breeder records phenotypes and wants a defensible estimate of average
performance. A **[statistical model](https://www.itl.nist.gov/div898/handbook/pri/section1/pri11.htm)** turns that question into assumptions and unknown quantities that can be estimated from records.

## From observations to an estimate

For a first model, let phenotype $y_i$ of animal $i$ among $n$ observed
animals satisfy

$$
y_i=\mu+\varepsilon_i, \qquad \varepsilon_i\sim N(0,\sigma^2).
$$

The parameter $\mu$ is the fixed but unknown population mean and $\sigma$ is
the assumed residual standard deviation. The normal, independent, common-
variance residual model is an assumption, not a biological fact. Family,
management, or environmental structure can make it inadequate.

An *estimator* is a rule applied to possible datasets, while an *estimate* is
the number it returns for this dataset. Thus the sample mean $\bar y$ is an
estimator of $\mu$, and $\hat\mu$ is its realised estimate. A *prediction*
instead concerns an unobserved outcome, such as a future offspring phenotype.
Estimating a mean and predicting a new record are related but different tasks.

```{r}
#| label: estimation-data
#| code-fold: true
d <- readRDS("../../demo-data/main/demo_data.rds")
y <- d$pheno$y_cont_qtl_snp[1:5]
mean(y)
```

The result is about $-1.518$. It estimates the mean under this simple model;
it is neither the true mean of every possible animal nor a prediction that all
five animals should have the same phenotype. The data are simulated project
data, not measurements from an empirical trial.

## Likelihood holds data fixed and compares parameters

A probability model asks which data could arise if $\mu$ and $\sigma$ were
known. A **[likelihood](https://www.itl.nist.gov/div898/handbook/apr/section1/apr161.htm)** uses the same formula after data are observed, holding
$\mathbf y$ fixed and comparing candidate parameter values:

$$
L(\mu;\mathbf y)=\prod_{i=1}^{n}
\frac{1}{\sqrt{2\pi\sigma^2}}
\exp\left\{-\frac{(y_i-\mu)^2}{2\sigma^2}\right\}.
$$

Likelihood is not a probability distribution over $\mu$: in frequentist
estimation, the parameter is fixed and unknown. Its maximiser is the
maximum-likelihood estimate. Software commonly uses log-likelihoods because
they turn products into sums.

```{r}
#| label: normal-likelihood
#| code-fold: true
mu_grid <- seq(-4, 2, length.out = 401)
log_likelihood <- vapply(
  mu_grid,
  function(mu) sum(dnorm(y, mean = mu, sd = 3, log = TRUE)),
  numeric(1)
)
mu_grid[which.max(log_likelihood)]
```

The grid gives $-1.525$, close to the exact sample mean because grid points
are 0.015 apart. With known $\sigma$, maximising this normal likelihood gives
$\hat\mu=\bar y$. Here $\sigma=3$ is supplied to focus on estimating $\mu$.

The same estimate minimises squared-error loss,

$$Q(\mu)=\sum_{i=1}^{n}(y_i-\mu)^2.$$

That agreement depends on the chosen normal model and loss. Absolute-error
loss targets a median, and another probability model can yield a different
estimator.

## Identifiability limits what records can separate

An **[identifiable parameter](https://www.itl.nist.gov/div898/handbook/pri/section5/pri541.htm)** has a unique implication for the distribution of observed data. If two parameter values always give the same fitted values, the data cannot distinguish them.

For $\mathbf y=\mathbf X\boldsymbol\beta+\boldsymbol\varepsilon$, suppose
two columns of the $n\times p$ marker matrix $\mathbf X$ are exact copies.
Increasing one coefficient and decreasing the other by the same amount leaves
$\mathbf X\boldsymbol\beta$ unchanged. Their separate coefficients are not
identified, although their combined fitted contribution can be. Near-copy
markers create collinearity: effects may be identifiable but unstable. A
marker coefficient is therefore not proof of a unique causal locus.

Identifiability also does not establish out-of-sample prediction quality. A
well-estimated training-population parameter can still predict poorly in a new
generation or environment. The next page adds fixed management effects and
correlated animal effects before genomic relationship matrices and REML.

## Exercises

1. A rule returns 2.1 for one dataset while a model parameter is $\theta$. Which is the estimator and which is the estimate?

::: {.callout-note collapse="true" title="Solution"}
The estimator is the rule mapping data to a value. The value 2.1 is its
estimate for this dataset; $\theta$ is the fixed but unknown parameter.
:::

2. After observing data, what varies in a likelihood comparison?

::: {.callout-note collapse="true" title="Solution"}
The observations are held fixed and candidate parameter values vary.
Likelihood ranks how well those values explain the same observed data.
:::

3. Can two exact-copy marker columns have separately identified regression coefficients from phenotypes alone?

::: {.callout-note collapse="true" title="Solution"}
No. Opposite coefficient changes can leave every fitted value unchanged, so an
extra constraint or representation is needed to separate the effects.
:::
