---
title: "Demo Datasets"
---

```{r}
#| include: false
library(DT)
library(tibble)

d       <- readRDS("main/demo_data.rds")
d_small <- readRDS("toy/demo_data_small.rds")

dt_table <- function(data, page = 15, scroll_x = FALSE) {
  DT::datatable(
    data,
    rownames = FALSE,
    class    = "cell-border compact stripe hover",
    options  = list(
      pageLength = page,
      lengthMenu = c(10, 15, 25, 50, 100),
      scrollX    = scroll_x,
      autoWidth  = FALSE,
      dom        = "frtip"
    )
  )
}

dt_static <- function(data) {
  DT::datatable(
    data,
    rownames = FALSE,
    class    = "cell-border compact stripe hover",
    options  = list(dom = "t", paging = FALSE, ordering = FALSE)
  )
}
```

Two pre-generated synthetic datasets accompany this site. **Every code
example in the [Tasks](../tasks/index.qmd) section uses these**, so the
numbers you see in tutorials match what you get when you run them locally
or via `system.file("extdata/demo_data.rds", package = "masbayes")`.

| Dataset | Individuals | SNPs | MH blocks | $W_{ah}$ columns | QTL | Use |
|---|---:|---:|---:|---:|---:|---|
| **main** | 200 (10 sires × 20 offspring) | 100 | 50 | ~149 | 10 | All standard tutorials |
| **toy**  | 100 (10 sires × 10 offspring) | 50  | 25 | ~75  | 5  | Quick-start (runs in seconds) |

Both are simulated with `set.seed(42)` for the genetic stream and
`set.seed(41)` for the within-family train/test split; re-running
`generate.R` produces bit-identical output.

::: {.callout-tip}
## One `.rds` per size, byte-identical to the sibling packages

Each dataset is a single named R list serialised as `.rds`. The files at
`demo-data/main/demo_data.rds` and `demo-data/toy/demo_data_small.rds`
are **byte-identical** to the same-named files shipped under
`masbayes/inst/extdata/` and `masreml/inst/extdata/`. The canonical
generator lives in the workspace as `tools/make_demo_data.R`; the local
mirror is `demo-data/generate.R` (output routed only into
`demo-data/<size>/`, never into either sibling-package `inst/extdata/`).
:::

## Loading

```{r}
#| eval: false
# From the masgenomics-docs working directory:
d <- readRDS("demo-data/main/demo_data.rds")

# Or from either installed package (byte-identical):
d <- readRDS(system.file("extdata", "demo_data.rds", package = "masbayes"))
d <- readRDS(system.file("extdata", "demo_data.rds", package = "masreml"))
```

`d` is a named list with the following slots:

```{r}
#| echo: false
slots <- tibble(
  slot          = c("snp", "mh", "allele_freq", "pheno", "pedigree",
                    "qtl", "meta", "family_id", "train_idx", "test_idx",
                    "map_snp", "map_mh"),
  class         = vapply(c("snp", "mh", "allele_freq", "pheno", "pedigree",
                           "qtl", "meta", "family_id", "train_idx", "test_idx",
                           "map_snp", "map_mh"),
                         function(nm) paste(class(d[[nm]]), collapse = "/"),
                         character(1)),
  shape_or_size = c(
    paste(nrow(d$snp),     "x", ncol(d$snp)),
    paste(nrow(d$mh),      "x", ncol(d$mh)),
    paste(length(d$allele_freq$haplotype), "rows (long form)"),
    paste(nrow(d$pheno),   "x", ncol(d$pheno)),
    paste(nrow(d$pedigree),"x", ncol(d$pedigree)),
    "4 sub-elements (snp_idx, mh_idx, effects_snp, effects_mh)",
    paste0(length(d$meta), " entries"),
    paste("length", length(d$family_id)),
    paste("length", length(d$train_idx)),
    paste("length", length(d$test_idx)),
    paste(nrow(d$map_snp), "x", ncol(d$map_snp)),
    paste(nrow(d$map_mh),  "x", ncol(d$map_mh))
  ),
  description   = c(
    "Diploid SNP dosage matrix; rownames = individual IDs, colnames = SNP IDs",
    "Microhaplotype matrix (integer-encoded alleles, 2 cols per block)",
    "Per-block allele frequency table (long form: haplotype / allele / freq)",
    "Phenotypes (SNP- and MH-architecture, continuous + binary; plus sex, true breeding values)",
    "3-column pedigree (id / sire / dam) including founders with NA parents",
    "True QTL positions and standardised effects under both architectures",
    "Simulation metadata (sizes, h2 target, seeds, sex effect betas)",
    "Family label per offspring (length n)",
    "Within-family training-set indices (16/family in main; 8/family in toy)",
    "Within-family test-set indices (4/family in main; 2/family in toy)",
    "Physical-position map for SNPs (SNP / CHROM / POS, synthetic)",
    "Physical-position map for MH blocks (block_id / chr / start_pos / end_pos / n_snps)"
  )
)
dt_static(slots)
```

## Simulation metadata (`d$meta`)

```{r}
#| echo: false
meta_df <- tibble(
  key   = names(d$meta),
  value = vapply(d$meta, function(x) format(x, digits = 4), character(1))
)
dt_static(meta_df)
```

Heritability target is $h^2 = 0.5$; realised values for the canonical
`main` dataset are $h^2_{SNP} \approx 0.516$ and $h^2_{MH} \approx 0.474$
(printed by `generate.R` on run). Binary phenotypes are obtained by
thresholding the continuous trait at its median, so prevalence is
exactly 0.5 by construction.

## Download

::: {.grid}
::: {.g-col-6}
**main dataset** (≈ 23 KB compressed)

[📦 demo_data.rds](main/demo_data.rds){.btn .btn-sm .btn-primary} &nbsp;
[📦 ZIP (whole `main/`)](main.zip){.btn .btn-sm .btn-outline-primary}
:::
::: {.g-col-6}
**toy dataset** (≈ 9 KB compressed)

[📦 demo_data_small.rds](toy/demo_data_small.rds){.btn .btn-sm .btn-secondary} &nbsp;
[📦 ZIP (whole `toy/`)](toy.zip){.btn .btn-sm .btn-outline-secondary}
:::
:::

## Phenotypes (`d$pheno`)

`n = 200` rows, eight columns: `id`, `sex` (factor F/M), the continuous
and binary phenotypes under both SNP and MH architectures, and the true
breeding values used to simulate each one.

```{r}
#| echo: false
dt_table(d$pheno, page = 10, scroll_x = TRUE)
```

## Pedigree (`d$pedigree`)

220 rows (10 sires + 10 dams + 200 offspring). Sires and dams have
`NA` parents; offspring carry the engineered sire/dam pairing.

```{r}
#| echo: false
dt_table(d$pedigree, page = 10)
```

## SNP map (`d$map_snp`)

Synthetic but deterministic physical positions; 100 SNPs evenly
distributed across 5 chromosomes at 100 kb spacing starting from
1 Mb on each chromosome.

```{r}
#| echo: false
dt_table(d$map_snp, page = 10)
```

## MH block map (`d$map_mh`)

50 blocks, 2 SNPs per block, 10 blocks per chromosome. Schema matches
the maspipeline `microhaplotype_coordinates.csv` format
(`block_id / chr / start_pos / end_pos / n_snps`) so production
pipelines can consume `d$map_mh` without translation.

```{r}
#| echo: false
dt_table(d$map_mh, page = 10)
```

## SNP genotypes (`d$snp`)

Integer matrix; `0/1/2` dosage. Preview below shows the first 10
individuals × first 12 SNPs of the `main` dataset.

```{r}
#| echo: false
snp_preview <- as.data.frame(d$snp[1:10, 1:12])
snp_preview <- cbind(id = rownames(snp_preview), snp_preview)
dt_table(snp_preview, page = 10, scroll_x = TRUE)
```

## Microhaplotype genotypes (`d$mh`)

Integer matrix; 2 columns per block holding the polynomial-encoded
allele ids for haplotypes 1 and 2. The `block_id` attribute maps each
column to its block:

```{r}
#| eval: false
attr(d$mh, "block_id")[1:6]
#> "block_1" "block_1" "block_2" "block_2" "block_3" "block_3"
```

Preview: first 10 individuals × first 6 columns (= blocks 1–3):

```{r}
#| echo: false
mh_preview <- as.data.frame(d$mh[1:10, 1:6])
colnames(mh_preview) <- paste0(rep(c("blk1", "blk2", "blk3"), each = 2),
                               c("_h1", "_h2"))
mh_preview <- cbind(id = rownames(mh_preview), mh_preview)
dt_table(mh_preview, page = 10)
```

## Per-chromosome MH genotype files (derived)

Alongside `demo_data.rds`, `generate.R` writes one MicrohapsSel-compatible
tab-delimited file per chromosome at
`demo-data/<size>/mh_genotypes/hap_geno_<chr>`. These are **derived from
`d$mh` + `d$map_mh`** and can be plugged into MicrohapsSel-pipeline
tooling unchanged; the canonical `.rds` remains the source of truth.

Header layout per file: `ID`, then 4 columns per local block —
`hap_<i>_1, hap_<i>_1, hap_<i>_2, hap_<i>_2` — with values in `{1, 2}`
(per-SNP haplotype calls, recovered by decoding the polynomial-encoded
block ids in `d$mh`). Local block numbering restarts at 1 in each
per-chr file.

Preview of `main/mh_genotypes/hap_geno_1` (first 10 individuals × first
3 blocks = 13 columns):

```{r}
#| echo: false
mh1 <- read.table("main/mh_genotypes/hap_geno_1", header = TRUE,
                  sep = "\t", check.names = FALSE,
                  nrows = 10, stringsAsFactors = FALSE)
mh1 <- mh1[, 1:13]
dt_table(mh1, page = 10, scroll_x = TRUE)
```

Download per-chromosome files (5 chromosomes per dataset):

[main chr 1](main/mh_genotypes/hap_geno_1){.btn .btn-sm .btn-outline-primary}
[main chr 2](main/mh_genotypes/hap_geno_2){.btn .btn-sm .btn-outline-primary}
[main chr 3](main/mh_genotypes/hap_geno_3){.btn .btn-sm .btn-outline-primary}
[main chr 4](main/mh_genotypes/hap_geno_4){.btn .btn-sm .btn-outline-primary}
[main chr 5](main/mh_genotypes/hap_geno_5){.btn .btn-sm .btn-outline-primary}
&nbsp;
[toy chr 1](toy/mh_genotypes/hap_geno_1){.btn .btn-sm .btn-outline-secondary}
[toy chr 2](toy/mh_genotypes/hap_geno_2){.btn .btn-sm .btn-outline-secondary}
[toy chr 3](toy/mh_genotypes/hap_geno_3){.btn .btn-sm .btn-outline-secondary}
[toy chr 4](toy/mh_genotypes/hap_geno_4){.btn .btn-sm .btn-outline-secondary}
[toy chr 5](toy/mh_genotypes/hap_geno_5){.btn .btn-sm .btn-outline-secondary}

## Allele frequencies (`d$allele_freq`)

Long-form table consumed by `masbayes::construct_wah_matrix()` when no
reference structure is supplied:

```{r}
#| echo: false
af <- tibble(
  haplotype = d$allele_freq$haplotype,
  allele    = d$allele_freq$allele,
  freq      = round(d$allele_freq$freq, 4)
)
dt_table(af, page = 15)
```

## QTL ground truth (`d$qtl`)

Ten QTL under each architecture, with unit-length-normalised effects.
**Use these for QTL recovery analysis only**, not as model input.

```{r}
#| echo: false
qtl_summary <- tibble(
  architecture = c("SNP", "MH"),
  positions    = c(paste(d$qtl$snp_idx, collapse = ", "),
                   paste(d$qtl$mh_idx,  collapse = ", ")),
  effect_min   = round(c(min(d$qtl$effects_snp),
                         min(d$qtl$effects_mh)), 4),
  effect_max   = round(c(max(d$qtl$effects_snp),
                         max(d$qtl$effects_mh)), 4),
  sum_sq       = round(c(sum(d$qtl$effects_snp ^ 2),
                         sum(d$qtl$effects_mh  ^ 2)), 4)
)
dt_static(qtl_summary)
```

## Within-family train/test split

The split is stratified by family: each of the 10 families contributes
16 offspring to the training set and 4 to the test set (8 / 2 in `toy`).
This keeps full-sibs out of the test partition's predictor pool without
losing any family-level relatedness.

```{r}
#| echo: false
split_summary <- tibble(
  set    = c("train", "test"),
  n      = c(length(d$train_idx), length(d$test_idx)),
  per_family = c(d$meta$n_per_family - 4L, 4L)
)
dt_static(split_summary)
```

## Regenerating locally

`generate.R` requires `library(masbayes)` (uses `construct_wah_matrix()`).

```bash
Rscript demo-data/generate.R main toy   # both sizes
Rscript demo-data/generate.R main       # large only
Rscript demo-data/generate.R toy        # small only
```

::: {.callout-note collapse="true"}
## Show generate.R

```{r}
#| eval: false
#| file: generate.R
```
:::
