---
title: "Demo Datasets"
---

```{r}
#| include: false
library(reactable)
library(tibble)
library(dplyr)
library(readr)
```

Two pre-generated synthetic datasets accompany this site. **Every code
example in the [Tasks](../tasks/index.qmd) section uses these**, so the
numbers you see in tutorials match what you get when you run them locally.

| Dataset | Individuals | SNPs | MH blocks | Hap blocks | Use |
|---|---:|---:|---:|---:|---|
| **main** | 1,000 | 5,000 | ~5,000 | ~500 | All standard tutorials |
| **toy**  | 200   | 500   | ~500   | ~220 | Quick-start (runs in seconds) |

Both are simulated from `generate.R` with `set.seed(42)` —
re-running produces bit-identical files.

::: {.callout-tip}
## Layout matches `MicrohapsSel/simulation/model_fitting`

The file structure mirrors what the user's MicrohapsSel pipeline expects.
You can plug these CSVs into `main-mh-fit-eval.R` or `main-snp-fit-eval.R`
unchanged.
:::

## Simulation parameters

- **`set.seed(42)`** at the top of `generate.R`; no other seed calls.
- **Genetic architecture:** randomly drawn QTL (50 SNPs for `main`, 20 for `toy`) with effects from $N(0, 1)$.
- **Heritability:** $h^2 = 0.4$.
- **Phenotypes:** continuous (`pheno_continuous_G1.csv`) and binary (threshold at median, `pheno_binary_G1.csv`).
- **Pedigree:** 4 generations; ~10% founders (NA sire/dam), random mating among earlier individuals; family lineages traced via the sire.
- **MH blocks:** built from sliding 3-SNP windows (max 125bp span).
- **Haplotype blocks:** built from 100kb windows.

## Download

::: {.grid}
::: {.g-col-6}
**main dataset (~76 MB uncompressed)**

[📦 Download main as ZIP](main.zip){.btn .btn-primary} &nbsp;
:::
::: {.g-col-6}
**toy dataset (~2 MB uncompressed)**

[📦 Download toy as ZIP](toy.zip){.btn .btn-secondary} &nbsp;
:::
:::

## Pedigree

`pedigree_G1.csv` — full table, 1,000 rows. Columns: `individual_id`,
`family_id`, `population_type`, `sire`, `dam`, `generation`.

```{r}
#| echo: false
ped <- read_csv("main/pedigree_G1.csv", show_col_types = FALSE)
reactable(
  ped,
  searchable = TRUE, sortable = TRUE,
  pagination = TRUE, defaultPageSize = 15,
  highlight = TRUE, compact = TRUE, bordered = TRUE
)
```

[Download pedigree_G1.csv](main/pedigree_G1.csv){.btn .btn-sm .btn-primary}

## Phenotypes

### Continuous trait (`pheno_continuous_G1.csv`)

```{r}
#| echo: false
pheno_c <- read_csv("main/pheno_continuous_G1.csv", show_col_types = FALSE)
reactable(
  pheno_c,
  searchable = TRUE, sortable = TRUE,
  pagination = TRUE, defaultPageSize = 15,
  highlight = TRUE, compact = TRUE, bordered = TRUE
)
```

[Download pheno_continuous_G1.csv](main/pheno_continuous_G1.csv){.btn .btn-sm .btn-primary}

### Binary trait (`pheno_binary_G1.csv`)

```{r}
#| echo: false
pheno_b <- read_csv("main/pheno_binary_G1.csv", show_col_types = FALSE)
reactable(
  pheno_b,
  searchable = TRUE, sortable = TRUE,
  pagination = TRUE, defaultPageSize = 15,
  highlight = TRUE, compact = TRUE, bordered = TRUE
)
```

[Download pheno_binary_G1.csv](main/pheno_binary_G1.csv){.btn .btn-sm .btn-primary}

## SNP genotypes

### `geno_G1.csv` — wide CSV (preview only)

`individual_id`, `generation`, `population_type`, then 5,000 SNP dosage
columns (0/1/2). The preview below shows the first 50 individuals × first
20 SNPs.

```{r}
#| echo: false
geno_preview <- read_csv("main/geno_G1.csv", n_max = 50, show_col_types = FALSE)
geno_preview <- geno_preview[, 1:23]   # id + generation + pop_type + first 20 SNPs
reactable(
  geno_preview,
  searchable = TRUE, sortable = TRUE,
  pagination = TRUE, defaultPageSize = 10,
  highlight = TRUE, compact = TRUE, bordered = TRUE,
  defaultColDef = colDef(minWidth = 80)
)
```

[Download geno_G1.csv](main/geno_G1.csv){.btn .btn-sm .btn-primary}
&nbsp;
[Download salmon_60k.vcf](main/salmon_60k.vcf){.btn .btn-sm .btn-secondary}

### SNP map (`snp_map.txt`)

Tab-delimited; columns `snp_id`, `chr`, `position`. Used by QTL recovery
analysis and downstream tools.

```{r}
#| echo: false
snp_map <- read_tsv("main/snp_map.txt", show_col_types = FALSE)
reactable(
  snp_map,
  searchable = TRUE, sortable = TRUE,
  pagination = TRUE, defaultPageSize = 15,
  highlight = TRUE, compact = TRUE, bordered = TRUE
)
```

[Download snp_map.txt](main/snp_map.txt){.btn .btn-sm .btn-primary}

## Microhaplotype genotypes

### Per-chromosome MH genotype files

Each chromosome has one file under `mh_genotypes/hap_geno_<chr>`. Tab-delimited.
Format: first column is `ID`; subsequent columns are 4-per-block
(`hap_<idx>_1, hap_<idx>_1, hap_<idx>_2, hap_<idx>_2`) — duplicated allele
columns matching the MicrohapsSel pipeline expectation.

Preview of `hap_geno_1` (first 50 individuals × first 20 columns):

```{r}
#| echo: false
mh1 <- read.table("main/mh_genotypes/hap_geno_1",
                  header = TRUE, sep = "\t", check.names = FALSE,
                  nrows = 50)
mh1 <- mh1[, 1:min(21, ncol(mh1))]
reactable(
  mh1,
  searchable = TRUE, sortable = TRUE,
  pagination = TRUE, defaultPageSize = 10,
  highlight = TRUE, compact = TRUE, bordered = TRUE,
  defaultColDef = colDef(minWidth = 80)
)
```

Download per-chromosome files (chromosomes 1–10):

[chr1](main/mh_genotypes/hap_geno_1){.btn .btn-sm .btn-outline-primary}
[chr2](main/mh_genotypes/hap_geno_2){.btn .btn-sm .btn-outline-primary}
[chr3](main/mh_genotypes/hap_geno_3){.btn .btn-sm .btn-outline-primary}
[chr4](main/mh_genotypes/hap_geno_4){.btn .btn-sm .btn-outline-primary}
[chr5](main/mh_genotypes/hap_geno_5){.btn .btn-sm .btn-outline-primary}
[chr6](main/mh_genotypes/hap_geno_6){.btn .btn-sm .btn-outline-primary}
[chr7](main/mh_genotypes/hap_geno_7){.btn .btn-sm .btn-outline-primary}
[chr8](main/mh_genotypes/hap_geno_8){.btn .btn-sm .btn-outline-primary}
[chr9](main/mh_genotypes/hap_geno_9){.btn .btn-sm .btn-outline-primary}
[chr10](main/mh_genotypes/hap_geno_10){.btn .btn-sm .btn-outline-primary}

### Microhaplotype coordinates (`mh_info_ld_haploblock_G0/stats/microhaplotype_coordinates.csv`)

```{r}
#| echo: false
mh_coords <- read_csv("main/mh_info_ld_haploblock_G0/stats/microhaplotype_coordinates.csv",
                      show_col_types = FALSE)
reactable(
  mh_coords,
  searchable = TRUE, sortable = TRUE,
  pagination = TRUE, defaultPageSize = 15,
  highlight = TRUE, compact = TRUE, bordered = TRUE
)
```

[Download mh coordinates](main/mh_info_ld_haploblock_G0/stats/microhaplotype_coordinates.csv){.btn .btn-sm .btn-primary}

## QTL information (true causal variants)

`qtl/` contains RDS files documenting the true causal variants used in
phenotype simulation. **Do not use these as model input** — they are ground
truth for QTL recovery analysis only.

```{r}
#| echo: false
qtl_snp_ids    <- readRDS("main/qtl/qtl_snp_ids.rds")
qtl_mh_blocks  <- readRDS("main/qtl/qtl_mh_blocks.rds")
qtl_hap_blocks <- readRDS("main/qtl/qtl_hap_blocks.rds")
qtl_summary <- tibble(
  file        = c("qtl_snp_ids.rds", "qtl_mh_blocks.rds", "qtl_hap_blocks.rds"),
  description = c("SNP IDs flagged as QTL",
                  "MH block IDs flagged as QTL",
                  "Haplotype block IDs flagged as QTL"),
  n           = c(length(qtl_snp_ids), length(qtl_mh_blocks), length(qtl_hap_blocks))
)
reactable(qtl_summary, sortable = FALSE, compact = TRUE, bordered = TRUE)
```

[Download qtl_snp_ids.rds](main/qtl/qtl_snp_ids.rds){.btn .btn-sm .btn-outline-primary}
[Download qtl_mh_blocks.rds](main/qtl/qtl_mh_blocks.rds){.btn .btn-sm .btn-outline-primary}
[Download qtl_hap_blocks.rds](main/qtl/qtl_hap_blocks.rds){.btn .btn-sm .btn-outline-primary}
[Download snp_meta.rds](main/qtl/snp_meta.rds){.btn .btn-sm .btn-outline-primary}

## Generation script

The full simulation script is committed to the repo. To regenerate locally:

```bash
Rscript demo-data/generate.R main   # 1000 ind, 10 chr, 500 SNPs/chr
Rscript demo-data/generate.R toy    # 200  ind, 5  chr, 100 SNPs/chr
```

::: {.callout-note collapse="true"}
## Show generate.R

```{r}
#| eval: false
#| file: generate.R
```
:::
