---
title: "Microhaplotypes"
---

```{r}
#| include: false
library(DT)
library(tibble)

d <- readRDS("../demo-data/main/demo_data.rds")

dt_table <- function(data, page = 10, scroll_x = FALSE) {
  DT::datatable(
    data,
    rownames = FALSE,
    class    = "cell-border compact stripe hover",
    options  = list(
      pageLength = page,
      scrollX    = scroll_x,
      autoWidth  = FALSE,
      dom        = "frtip"
    )
  )
}
```

## Schema

Microhaplotypes are short genomic windows in which the **phased
combination** of SNP alleles defines a multi-allelic locus. Both
packages consume microhaplotypes in **two interchangeable layouts**:

1. **On-disk file layout (maspipeline / MicrohapsSel format)** — one
   tab-delimited file per chromosome at
   `mh_genotypes/hap_geno_<chr>`. The header is
   `ID\thap_<i>_1\thap_<i>_1\thap_<i>_2\thap_<i>_2\t...` (block `<i>`'s
   two haplotype-strand names repeated, one column per SNP within the
   block). Values are per-SNP allele codes `1` or `2`. Block coordinates
   live in a companion `microhaplotype_coordinates.csv`.

2. **In-memory compact layout (`d$mh`)** — an integer matrix
   $n \times (2 \cdot n_{blocks})$. Two columns per block, one per
   strand. Each cell is a **base-3 polynomial-encoded** integer
   identifier of the block's haplotype on that strand
   (`sum(snp_calls * 3 ^ (seq_len(k) - 1))` for `k` SNPs per block).
   `attr(d$mh, "block_id")` maps each column to its block label.

`d$allele_freq` carries the parallel vectors `haplotype`, `allele`,
`freq` used by `masbayes::construct_wah_matrix()` when no reference
population is supplied.

The companion **MH coordinate map** (`d$map_mh` /
`microhaplotype_coordinates.csv`):

| Column | Type | Description |
|---|---|---|
| `block_id` | character | Block identifier, joins `unique(attr(mh, "block_id"))` |
| `chr` | integer | Chromosome |
| `start_pos` / `end_pos` | integer | bp coordinates of the spanned SNPs |
| `n_snps` | integer | SNPs per block (2 in the demo) |

## Example (from the demo dataset)

`d$mh` is `r nrow(d$mh)` × `r ncol(d$mh)` (main; 250 blocks × 2 strands
× 200 individuals) and 100 × 50 (toy). Each value is one of `{4, 5, 7,
8}` because each block has 2 SNPs and each SNP allele is in `{1, 2}` —
the four possible encoded haplotype ids.

First 8 individuals × first 6 columns (= blocks 1–3) of `d$mh`:

```{r}
#| echo: false
mh_prev <- as.data.frame(d$mh[1:8, 1:6])
colnames(mh_prev) <- paste0(rep(c("blk1", "blk2", "blk3"), each = 2),
                            c("_h1", "_h2"))
mh_prev <- cbind(id = rownames(mh_prev), mh_prev)
dt_table(mh_prev, page = 8)
```

Block-id attribute (first 6 entries):

```{r}
#| echo: false
ba <- tibble(
  col       = 1:6,
  block_id  = attr(d$mh, "block_id")[1:6]
)
DT::datatable(ba, rownames = FALSE,
              options = list(dom = "t", paging = FALSE))
```

The matching coordinate rows from `d$map_mh`:

```{r}
#| echo: false
dt_table(d$map_mh[1:5, ], page = 5)
```

The full `d$map_mh` is also written as a standalone CSV with the same
schema as the maspipeline `microhaplotype_coordinates.csv`:

[Download microhaplotype_coordinates.csv (large)](../demo-data/main/microhaplotype_coordinates.csv){.btn .btn-sm .btn-primary}
&nbsp;
[Download microhaplotype_coordinates.csv (small)](../demo-data/toy/microhaplotype_coordinates.csv){.btn .btn-sm .btn-outline-secondary}

## Per-chromosome microhaplotype genotype files (MicrohapsSel format)

Alongside the canonical `.rds`, `generate.R` emits one
MicrohapsSel-compatible TSV per chromosome at
`mh_genotypes/hap_geno_<chr>`. Header layout: `ID`, then 4 columns per
local block (`hap_<i>_1, hap_<i>_1, hap_<i>_2, hap_<i>_2`), values
`{1, 2}` (per-SNP haplotype calls). Local block numbering restarts at
1 in each per-chr file. These files are derived from `d$mh` + `d$map_mh`
(round-trip decode-then-re-encode verified) — the `.rds` remains the
source of truth.

Preview of `main/mh_genotypes/hap_geno_1` (first 10 individuals × first
3 local blocks = 13 columns):

```{r}
#| echo: false
mh1 <- read.table("../demo-data/main/mh_genotypes/hap_geno_1",
                  header = TRUE, sep = "\t", check.names = FALSE,
                  nrows = 10, stringsAsFactors = FALSE)
mh1 <- mh1[, 1:13]
dt_table(mh1, page = 10, scroll_x = TRUE)
```

Download per-chromosome files (5 chromosomes per size):

[main chr 1](../demo-data/main/mh_genotypes/hap_geno_1){.btn .btn-sm .btn-outline-primary}
[main chr 2](../demo-data/main/mh_genotypes/hap_geno_2){.btn .btn-sm .btn-outline-primary}
[main chr 3](../demo-data/main/mh_genotypes/hap_geno_3){.btn .btn-sm .btn-outline-primary}
[main chr 4](../demo-data/main/mh_genotypes/hap_geno_4){.btn .btn-sm .btn-outline-primary}
[main chr 5](../demo-data/main/mh_genotypes/hap_geno_5){.btn .btn-sm .btn-outline-primary}
&nbsp;
[toy chr 1](../demo-data/toy/mh_genotypes/hap_geno_1){.btn .btn-sm .btn-outline-secondary}
[toy chr 2](../demo-data/toy/mh_genotypes/hap_geno_2){.btn .btn-sm .btn-outline-secondary}
[toy chr 3](../demo-data/toy/mh_genotypes/hap_geno_3){.btn .btn-sm .btn-outline-secondary}
[toy chr 4](../demo-data/toy/mh_genotypes/hap_geno_4){.btn .btn-sm .btn-outline-secondary}
[toy chr 5](../demo-data/toy/mh_genotypes/hap_geno_5){.btn .btn-sm .btn-outline-secondary}

## Allele frequencies (`d$allele_freq`)

Long-form table consumed by `masbayes::construct_wah_matrix()` when no
reference structure is supplied:

```{r}
#| echo: false
af <- tibble(
  haplotype = d$allele_freq$haplotype,
  allele    = d$allele_freq$allele,
  freq      = round(d$allele_freq$freq, 4)
)
dt_table(af, page = 15)
```

## Loading

### From the bundled demo

```{r}
#| eval: false
library(masreml)   # or masbayes
d <- load_data("large")

dim(d$mh)                       # 200 500
attr(d$mh, "block_id")[1:6]     # "block_1" "block_1" "block_2" "block_2" ...
head(d$map_mh)
```

### From the per-chromosome file layout (maspipeline output)

```{r}
#| eval: false
chr1 <- read.table("demo-data/main/mh_genotypes/hap_geno_1",
                   header = TRUE, sep = "\t", check.names = FALSE)
# First column is "ID"; subsequent columns are 4-per-block
# (hap_<i>_1, hap_<i>_1, hap_<i>_2, hap_<i>_2) with values 1 or 2.
```

To convert the on-disk per-chr layout into the compact `d$mh` shape,
collapse each block's 4 columns into 2 base-3 polynomial-encoded ids:

```{r}
#| eval: false
# Per block: snp1_h1, snp2_h1, snp1_h2, snp2_h2 -> two encoded ids
encode_block <- function(s1_h1, s2_h1, s1_h2, s2_h2) {
  cbind(
    h1 = as.integer(s1_h1) + 3L * as.integer(s2_h1),
    h2 = as.integer(s1_h2) + 3L * as.integer(s2_h2)
  )
}
```

### Building model inputs from a microhaplotype matrix

```{r}
#| eval: false
library(masreml)
library(masbayes)

# masreml: microhaplotype-based GRM. build_G_mh auto-detects the compact layout.
G_mh <- build_G_mh(mh_list = d$mh, ids = rownames(d$mh))

# masbayes: W_alpha-h design matrix for BayesA / BayesR on microhaplotype alleles.
wah <- construct_wah_matrix(
  hap_matrix             = d$mh,
  colnames               = attr(d$mh, "block_id"),
  allele_freq_filtered   = d$allele_freq,
  reference_structure    = NULL,
  drop_baseline          = TRUE
)
str(wah)            # W_ah is n x (sum(n_alleles_b - 1))
```

## Common pitfalls

::: {.callout-warning}
**`block_id` length must equal `ncol(mh)`.** The block-id vector is
repeated **twice per block** (once per strand), so it has length
$2 \cdot n_{blocks}$, not $n_{blocks}$. `masbayes::construct_wah_matrix()`
consumes it via the `colnames` argument. `masreml::build_G_mh()` does
not read this attribute — it infers block boundaries positionally
from `ncol(mh) / 2` (paired strands), so block ordering in `mh` must
match the genotype-pipeline output.
:::

::: {.callout-warning}
**Allele-frequency reference matters for unrelated populations.** If
the prediction set is genetically distinct from the training set, the
frequencies bundled in `d$allele_freq` (computed on the bundled
population) are not representative. Pass an explicit
`reference_structure` to `construct_wah_matrix()` so the encoding
matches the population whose effects you want to estimate.
:::

::: {.callout-warning}
**Partial blocks / missing strands.** Real maspipeline output may emit
NA for a block on an individual whose phasing failed. Neither package
imputes these — drop or impute before fitting. The demo data has no
missingness.
:::

::: {.callout-tip}
**Why compact (`d$mh`) instead of the per-chr file format?** The
compact layout is 4× smaller in memory and exposes the same biology
($2 \cdot n_{blocks}$ haplotype slots). All downstream microhaplotype-aware code in
`masreml` / `masbayes` accepts both; the on-disk format only matters
for pipeline-level interop.
:::

📐 **Related theory:** [Biallelic & Multi-allelic](../theory/biallelic-and-multiallelic.qmd) ·
[Mixed Model & BLUP Family](../theory/mixed-model-blup-family.qmd) ·
[Bayesian Alphabet](../theory/bayesian-alphabet.qmd)

🛠 **Used in:** [Genomic Prediction](../tasks/genomic-prediction/index.qmd) ·
[GWAS → GWABLUP](../tasks/gwas/masreml.qmd#step-6-gwas-assisted-prediction-gwablup)
