---
title: "Pedigree"
---

```{r}
#| include: false
library(DT)
library(tibble)

d <- readRDS("../demo-data/main/demo_data.rds")

dt_table <- function(data, page = 10, scroll_x = FALSE) {
  DT::datatable(
    data,
    rownames = FALSE,
    class    = "cell-border compact stripe hover",
    options  = list(
      pageLength = page,
      scrollX    = scroll_x,
      autoWidth  = FALSE,
      dom        = "frtip"
    )
  )
}
```

## Schema

Pedigree is consumed by **masreml only** — `masbayes` is a
marker-based package and does not use a numerator-relationship matrix.

The data frame has one row per individual *including founders*:

| Column | Type | Description |
|---|---|---|
| `id` | character | Individual identifier; joins to `rownames(snp)` for offspring |
| `sire` | character or `NA` | Sire identifier, `NA` for founders without recorded sire |
| `dam` | character or `NA` | Dam identifier, `NA` for founders without recorded dam |

The demo dataset has 220 rows (large) / 120 rows (small): 10 sire
founders + 10 dam founders (all with `NA` sire/dam) plus the offspring
with their two-parent ancestry recorded.

## Example (from the demo dataset)

Full pedigree, with founders surfaced first:

```{r}
#| echo: false
dt_table(d$pedigree, page = 10)
```

Founder vs. offspring count:

```{r}
#| echo: false
summary_df <- tibble(
  category   = c("Sire founders",
                 "Dam founders",
                 "Offspring (full-sib families)"),
  count      = c(sum(is.na(d$pedigree$sire) & is.na(d$pedigree$dam) &
                     grepl("^SIRE", d$pedigree$id)),
                 sum(is.na(d$pedigree$sire) & is.na(d$pedigree$dam) &
                     grepl("^DAM",  d$pedigree$id)),
                 sum(!is.na(d$pedigree$sire) & !is.na(d$pedigree$dam))),
  example_id = c(d$pedigree$id[grepl("^SIRE", d$pedigree$id)][1],
                 d$pedigree$id[grepl("^DAM",  d$pedigree$id)][1],
                 d$pedigree$id[grepl("^IND",  d$pedigree$id)][1])
)
DT::datatable(summary_df, rownames = FALSE,
              options = list(dom = "t", paging = FALSE))
```

## Loading

`masreml::build_A_ped()` expects the sire/dam columns as **integer
row-indices** (with `0` meaning "unknown"), not as character IDs. The
demo pedigree stores them as character strings (e.g. `"SIRE05"`), so a
small conversion step is required before calling the builder:

```{r}
#| eval: false
library(masreml)
d <- masreml::load_data("large")

# Map every id to its 1-based row index in the pedigree frame.
id_to_row <- setNames(seq_len(nrow(d$pedigree)), d$pedigree$id)

ped_int <- data.frame(
  id   = d$pedigree$id,
  sire = ifelse(is.na(d$pedigree$sire), 0L, id_to_row[d$pedigree$sire]),
  dam  = ifelse(is.na(d$pedigree$dam),  0L, id_to_row[d$pedigree$dam]),
  stringsAsFactors = FALSE
)

A <- masreml::build_A_ped(ped_int)
str(A)                          # 220 x 220 numeric, rownames = ids
```

### Using A inside the mixed model

`masreml::masreml()` does not accept a pedigree argument directly.
Build $\mathbf{A}$ first with `build_A_ped()`, then pass it through the
`G` slot under one of the recognised keys (`snp_add`, `snp_dom`, or
`mh_add` — the key is the random-effect label, not a content check):

```{r}
#| eval: false
y <- d$pheno$y_cont_qtl_snp
names(y) <- d$pheno$id
X_fixed <- stats::model.matrix(~ sex, data = d$pheno)

# Build A on the full pedigree (220 individuals = founders + offspring)
A_full <- masreml::build_A_ped(ped_int)

# Restrict to the phenotyped offspring before fitting.
A_pheno <- A_full[d$pheno$id, d$pheno$id]

# A-BLUP: the pedigree-derived relationship matrix as the single
# random-effect kernel. The slot key here is `snp_add` simply because
# that is the recognised label for an additive-relationship slot in
# masreml's G list — its contents are A, not a SNP-derived G.
fit_a <- masreml::masreml(
  y     = y,
  X     = X_fixed,
  G     = list(snp_add = A_pheno),
  trait = "continuous"
)
```

::: {.callout-note}
**Single-step GBLUP (combining $\mathbf{A}$ and $\mathbf{G}$).** masreml's
`G` slot recognises only the three keys above; it does not provide a
dedicated pedigree-plus-genomic kernel. For single-step prediction,
build the combined $\mathbf{H}$ matrix
([Misztal et al. 2009](https://doi.org/10.3168/jds.2009-2064))
externally and pass it as `G = list(snp_add = H)`.
:::

## Common pitfalls

::: {.callout-warning}
**Founders must appear in the pedigree.** A common mistake is including
only offspring rows. `build_A_ped()` walks the matrix bottom-up; if a
sire row is missing, its descendants lose their genetic-relationship
columns and inbreeding estimates collapse to 0. The demo pedigree's
220 rows include 20 founder rows for this reason.
:::

::: {.callout-warning}
**Missing parents must be `NA` (or `0` after conversion), not the empty
string or `"NA"`.** `is.na()` is the only check that triggers founder
handling — the literal text `"NA"` is not the same as `NA_character_`.
:::

::: {.callout-warning}
**Pedigree row order matters for the integer-index conversion shown
above.** Each parent must appear in the data frame *before* (i.e. at a
lower row index than) any of its offspring. The demo pedigree
satisfies this — sires/dams come first, then offspring. If your real
pedigree has cycles or is unordered, sort topologically before passing
to `build_A_ped()`:

```{r}
#| eval: false
# Toy topological sort assuming no cycles
sort_pedigree <- function(ped) {
  done <- character()
  out  <- ped[0, ]
  while (nrow(ped) > 0) {
    ready <- is.na(ped$sire) | (ped$sire %in% done)
    ready <- ready & (is.na(ped$dam) | (ped$dam %in% done))
    if (!any(ready)) stop("Pedigree has cycles or missing parent rows.")
    out  <- rbind(out, ped[ready, ])
    done <- c(done, ped$id[ready])
    ped  <- ped[!ready, ]
  }
  out
}
```
:::

::: {.callout-tip}
**Pedigree is optional for marker-based prediction.** If you only need
GBLUP / BayesA / BayesR on markers, skip the pedigree entirely and use
`build_G_snp()` or `build_G_mh()` on the genotype matrix. The pedigree
matters when you have un-genotyped individuals (then `A` covers them)
or when you want to compare $A$ vs $G$ partitioning.
:::

📐 **Related theory:** [Mixed Model & BLUP Family](../theory/mixed-model-blup-family.qmd)

🛠 **Used in:** [Genomic Prediction](../tasks/genomic-prediction.qmd)
