---
title: "Phenotype"
---

```{r}
#| include: false
library(DT)
library(tibble)

d <- readRDS("../demo-data/main/demo_data.rds")

dt_table <- function(data, page = 10, scroll_x = FALSE) {
  DT::datatable(
    data,
    rownames = FALSE,
    class    = "cell-border compact stripe hover",
    options  = list(
      pageLength = page,
      scrollX    = scroll_x,
      autoWidth  = FALSE,
      dom        = "frtip"
    )
  )
}
```

## Schema

Both packages consume phenotypes as a **data frame** with one row per
individual and at least three column kinds:

| Column kind | Type | Required by | Description |
|---|---|---|---|
| Identifier | character | both | Joins to `rownames(snp)` / `rownames(mh)` |
| Fixed-effect covariates | numeric or factor | both | e.g. `sex`, `age`, batch |
| Response(s) | numeric or integer 0/1 | both | One column per trait — continuous OR binary |
| (Optional) `tbv_*` | numeric | none | True breeding values; only meaningful for simulation studies |

The demo dataset bundles **both** continuous and binary versions of two
traits (SNP-architecture and MH-architecture). Standalone CSVs split
them out:

| File | Columns |
|---|---|
| `pheno_continuous_G1.csv` | `id, sex, y_cont_qtl_snp, y_cont_qtl_mh, tbv_qtl_snp, tbv_qtl_mh` |
| `pheno_binary_G1.csv`     | `id, sex, y_bin_qtl_snp, y_bin_qtl_mh, tbv_qtl_snp, tbv_qtl_mh` |

`d$pheno` carries all eight columns in one frame.

## Example (from the demo dataset)

```{r}
#| echo: false
dt_table(d$pheno, page = 10, scroll_x = TRUE)
```

Quick summary of the binary trait (median-thresholded, so prevalence
is 0.5 by construction):

```{r}
#| echo: false
bal <- tibble(
  trait        = c("y_bin_qtl_snp", "y_bin_qtl_mh"),
  n_cases      = c(sum(d$pheno$y_bin_qtl_snp == 1),
                   sum(d$pheno$y_bin_qtl_mh  == 1)),
  n_controls   = c(sum(d$pheno$y_bin_qtl_snp == 0),
                   sum(d$pheno$y_bin_qtl_mh  == 0)),
  prevalence   = c(mean(d$pheno$y_bin_qtl_snp),
                   mean(d$pheno$y_bin_qtl_mh))
)
DT::datatable(bal, rownames = FALSE,
              options = list(dom = "t", paging = FALSE))
```

## Loading

```{r}
#| eval: false
library(masreml)
d <- masreml::load_data("large")

# Continuous trait, SNP architecture
y_cont <- d$pheno$y_cont_qtl_snp

# Binary trait, MH architecture
y_bin  <- d$pheno$y_bin_qtl_mh

# Or load the standalone CSVs directly:
pheno_c <- read.csv("demo-data/main/pheno_continuous_G1.csv")
pheno_b <- read.csv("demo-data/main/pheno_binary_G1.csv")
```

### Passing phenotypes to each package

```{r}
#| eval: false
# masreml: REML mixed model with sex as fixed effect
fit <- masreml::masreml(
  formula = y_cont_qtl_snp ~ sex,
  data    = d$pheno,
  markers = list(snp_add = d$snp),
  pedigree = d$pedigree
)

# masbayes: BayesA / BayesR on a continuous trait
W   <- masbayes::construct_snp_matrix(d$snp, encoding = "vanRaden")
fit <- masbayes::run_bayesa(y = d$pheno$y_cont_qtl_snp, X = W)

# Binary trait via threshold model in masreml
fit_bin <- masreml::masreml(
  formula = y_bin_qtl_snp ~ sex,
  data    = d$pheno,
  family  = "binomial",                    # liability-scale model
  markers = list(snp_add = d$snp)
)
```

## Common pitfalls

::: {.callout-warning}
**Row order must match the genotype matrix.** `masreml` aligns on `id`
when both `data$id` and `rownames(markers$snp_add)` are present, but
this is the most common source of silent bugs when ID conventions
differ. Always verify before fitting:

```{r}
#| eval: false
stopifnot(identical(d$pheno$id, rownames(d$snp)))
```
:::

::: {.callout-warning}
**Factor levels must be consistent across train and test.** If your
fixed-effect covariate is a factor (e.g. `sex`), make sure both
train and test splits see all levels — otherwise the design-matrix
column count diverges and prediction breaks.
:::

::: {.callout-warning}
**Binary phenotypes go through a threshold model, not regression on
0/1.** Use `family = "binomial"` in `masreml::masreml()` so the
liability-scale variance is identified. Naive Gaussian fitting of
binary data produces unbiased EBVs in some regimes but distorts
heritability estimates.
:::

::: {.callout-tip}
**`tbv_*` columns are ground truth, not input.** They are only useful
for simulation: comparing $\widehat{u}$ against `tbv` is how the
[Cross-validation](../tasks/cross-validation.qmd) tutorial computes
accuracy. Never include `tbv_*` as a covariate.
:::

📐 **Related theory:** [Mixed Models & BLUP](../theory/mixed-models-blup.qmd) ·
[Threshold Model](../theory/threshold-model.qmd)

🛠 **Used in:** [Genomic Prediction](../tasks/genomic-prediction.qmd) ·
[Binary trait analysis](../tasks/binary-trait.qmd) ·
[Cross-validation](../tasks/cross-validation.qmd)
