---
title: "Phenotype"
---

```{r}
#| include: false
library(DT)
library(tibble)

d <- readRDS("../demo-data/main/demo_data.rds")

dt_table <- function(data, page = 10, scroll_x = FALSE) {
  DT::datatable(
    data,
    rownames = FALSE,
    class    = "cell-border compact stripe hover",
    options  = list(
      pageLength = page,
      scrollX    = scroll_x,
      autoWidth  = FALSE,
      dom        = "frtip"
    )
  )
}
```

## Schema

Both packages consume phenotypes as a **data frame** with one row per
individual and at least three column kinds:

| Column kind | Type | Required by | Description |
|---|---|---|---|
| Identifier | character | both | Joins to `rownames(snp)` / `rownames(mh)` |
| Fixed-effect covariates | numeric or factor | both | e.g. `sex`, `age`, batch |
| Response(s) | numeric or integer 0/1 | both | One column per trait — continuous OR binary |
| (Optional) `tbv_*` | numeric | none | True breeding values; only meaningful for simulation studies |

The demo dataset bundles **both** continuous and binary versions of two
traits (SNP-architecture and MH-architecture). Standalone CSVs split
them out:

| File | Columns |
|---|---|
| `pheno_continuous_G1.csv` | `id, sex, y_cont_qtl_snp, y_cont_qtl_mh, tbv_qtl_snp, tbv_qtl_mh` |
| `pheno_binary_G1.csv`     | `id, sex, y_bin_qtl_snp, y_bin_qtl_mh, tbv_qtl_snp, tbv_qtl_mh` |

`d$pheno` carries all eight columns in one frame.

## Example (from the demo dataset)

### Continuous trait (`pheno_continuous_G1.csv`)

Columns: `id, sex, y_cont_qtl_snp, y_cont_qtl_mh, tbv_qtl_snp,
tbv_qtl_mh`.

```{r}
#| echo: false
pheno_c <- read.csv("../demo-data/main/pheno_continuous_G1.csv",
                    stringsAsFactors = FALSE)
dt_table(pheno_c, page = 10, scroll_x = TRUE)
```

[Download pheno_continuous_G1.csv (large)](../demo-data/main/pheno_continuous_G1.csv){.btn .btn-sm .btn-primary}
&nbsp;
[Download pheno_continuous_G1.csv (small)](../demo-data/toy/pheno_continuous_G1.csv){.btn .btn-sm .btn-outline-secondary}

### Binary trait (`pheno_binary_G1.csv`)

Same schema but with the median-thresholded `y_bin_*` columns in place
of the continuous ones. Prevalence is 0.5 by construction.

```{r}
#| echo: false
pheno_b <- read.csv("../demo-data/main/pheno_binary_G1.csv",
                    stringsAsFactors = FALSE)
dt_table(pheno_b, page = 10, scroll_x = TRUE)
```

[Download pheno_binary_G1.csv (large)](../demo-data/main/pheno_binary_G1.csv){.btn .btn-sm .btn-primary}
&nbsp;
[Download pheno_binary_G1.csv (small)](../demo-data/toy/pheno_binary_G1.csv){.btn .btn-sm .btn-outline-secondary}

### Case/control balance

```{r}
#| echo: false
bal <- tibble(
  trait        = c("y_bin_qtl_snp", "y_bin_qtl_mh"),
  n_cases      = c(sum(d$pheno$y_bin_qtl_snp == 1),
                   sum(d$pheno$y_bin_qtl_mh  == 1)),
  n_controls   = c(sum(d$pheno$y_bin_qtl_snp == 0),
                   sum(d$pheno$y_bin_qtl_mh  == 0)),
  prevalence   = c(mean(d$pheno$y_bin_qtl_snp),
                   mean(d$pheno$y_bin_qtl_mh))
)
DT::datatable(bal, rownames = FALSE,
              options = list(dom = "t", paging = FALSE))
```

## QTL ground truth

Ten QTL under each architecture (SNP and MH), with
unit-length-normalised effects ($\sum \beta_i^2 = 1$). **Use these for
QTL recovery analysis only**, not as model input. Four standalone
`.rds` files mirror `d$qtl` with human-readable identifiers — SNP
names and MH block ids — so downstream tools can join against
`d$map_snp` / `d$map_mh` directly.

### Causal SNPs

```{r}
#| echo: false
qtl_snp_ids <- readRDS("../demo-data/main/qtl/qtl_snp_ids.rds")
effects_snp <- readRDS("../demo-data/main/qtl/effects_snp.rds")
qtl_snp_tbl <- tibble(
  rank     = seq_along(qtl_snp_ids),
  snp_id   = qtl_snp_ids,
  chrom    = d$map_snp$CHROM[match(qtl_snp_ids, d$map_snp$SNP)],
  position = d$map_snp$POS  [match(qtl_snp_ids, d$map_snp$SNP)],
  effect   = round(effects_snp, 4)
)
DT::datatable(qtl_snp_tbl, rownames = FALSE,
              options = list(dom = "t", paging = FALSE, ordering = FALSE))
```

### Causal MH blocks

`d$qtl$mh_idx` indexes columns of $W_{ah}$ rather than block columns
directly, because the causal "allele" within a multi-allelic block is
itself a free parameter. The derived `qtl_mh_blocks.rds` reduces each
$W_{ah}$ column to its parent block id; if two QTL fall in the same
block, the block id appears twice.

```{r}
#| echo: false
qtl_mh_blocks <- readRDS("../demo-data/main/qtl/qtl_mh_blocks.rds")
effects_mh    <- readRDS("../demo-data/main/qtl/effects_mh.rds")
qtl_mh_tbl <- tibble(
  rank      = seq_along(qtl_mh_blocks),
  block_id  = qtl_mh_blocks,
  chr       = d$map_mh$chr      [match(qtl_mh_blocks, d$map_mh$block_id)],
  start_pos = d$map_mh$start_pos[match(qtl_mh_blocks, d$map_mh$block_id)],
  end_pos   = d$map_mh$end_pos  [match(qtl_mh_blocks, d$map_mh$block_id)],
  effect    = round(effects_mh, 4)
)
DT::datatable(qtl_mh_tbl, rownames = FALSE,
              options = list(dom = "t", paging = FALSE, ordering = FALSE))
```

[Download qtl_snp_ids.rds](../demo-data/main/qtl/qtl_snp_ids.rds){.btn .btn-sm .btn-outline-primary}
[Download qtl_mh_blocks.rds](../demo-data/main/qtl/qtl_mh_blocks.rds){.btn .btn-sm .btn-outline-primary}
[Download effects_snp.rds](../demo-data/main/qtl/effects_snp.rds){.btn .btn-sm .btn-outline-primary}
[Download effects_mh.rds](../demo-data/main/qtl/effects_mh.rds){.btn .btn-sm .btn-outline-primary}

## Loading

```{r}
#| eval: false
library(masreml)
d <- masreml::load_data("large")

# Continuous trait, SNP architecture
y_cont <- d$pheno$y_cont_qtl_snp

# Binary trait, MH architecture
y_bin  <- d$pheno$y_bin_qtl_mh

# Or load the standalone CSVs directly:
pheno_c <- read.csv("demo-data/main/pheno_continuous_G1.csv")
pheno_b <- read.csv("demo-data/main/pheno_binary_G1.csv")
```

### Passing phenotypes to each package

`masreml::masreml()` takes the response as a named numeric vector `y`,
the fixed-effects design matrix `X` explicitly (build it with
`stats::model.matrix()`), and `markers` as a named list
(`snp_add`, `snp_dom`, `mh_add`). There is no `formula` argument and no
direct `pedigree` argument — for pedigree-based prediction, precompute
$\mathbf{A}$ via `build_A_ped()` and pass it through the `G` slot.

```{r}
#| eval: false
# Continuous trait, SNP-additive GBLUP with sex as fixed effect.
y <- d$pheno$y_cont_qtl_snp
names(y) <- d$pheno$id

X_fixed <- stats::model.matrix(~ sex, data = d$pheno)

fit <- masreml::masreml(
  y       = y,
  X       = X_fixed,
  markers = list(snp_add = d$snp),
  trait   = "continuous"
)

# masbayes: BayesA on the same trait. The wrapper takes the
# pre-standardised design matrix `w` plus colSums(w^2) as inputs;
# sigma2_g and sigma2_e_init are required hyperparameters.
W   <- masbayes::construct_snp_matrix(d$snp, encoding = "vanRaden")
fit <- masbayes::run_bayesa(
  w             = W,
  y             = y,
  wtw_diag      = colSums(W^2),
  X             = X_fixed,
  marker_type   = "snp",
  nu            = 4.5,
  sigma2_g      = var(y) * 0.5,
  sigma2_e_init = var(y) * 0.5,
  method        = "mcmc"
)

# Binary trait via threshold model in masreml (Laplace approximation).
y_bin <- d$pheno$y_bin_qtl_snp
names(y_bin) <- d$pheno$id
fit_bin <- masreml::masreml(
  y       = y_bin,
  X       = X_fixed,
  markers = list(snp_add = d$snp),
  trait   = "binary",
  link    = "logit"
)
```

## Common pitfalls

::: {.callout-warning}
**Row order must match the genotype matrix.** `masreml::masreml()`
aligns on `names(y)` against `rownames(markers$snp_add)` (and rownames
of `X`). This is the most common source of silent bugs when ID
conventions differ. Always verify before fitting:

```{r}
#| eval: false
stopifnot(identical(d$pheno$id, rownames(d$snp)))
```
:::

::: {.callout-warning}
**Factor levels must be consistent across train and test.** If your
fixed-effect covariate is a factor (e.g. `sex`), make sure both
train and test splits see all levels before calling
`stats::model.matrix()` — otherwise the design-matrix column count
diverges and prediction breaks.
:::

::: {.callout-warning}
**Binary phenotypes go through a threshold model, not regression on
0/1.** Pass `trait = "binary"` (with `link = "logit"` or `"probit"`) in
`masreml::masreml()` so the liability-scale variance is identified.
Naive Gaussian fitting of binary data (`trait = "continuous"`)
produces unbiased EBVs in some regimes but distorts heritability
estimates.
:::

::: {.callout-tip}
**`tbv_*` columns are ground truth, not input.** They are only useful
for simulation: comparing $\widehat{u}$ against `tbv` is how the
[Cross-validation](../tasks/cross-validation.qmd) tutorial computes
accuracy. Never include `tbv_*` as a covariate.
:::

📐 **Related theory:** [Mixed Model & BLUP Family](../theory/mixed-model-blup-family.qmd) ·
[Continuous and Binary Trait](../theory/continuous-and-binary-trait.qmd)

🛠 **Used in:** [Genomic Prediction](../tasks/genomic-prediction.qmd) ·
[Binary trait analysis](../tasks/binary-trait.qmd) ·
[Cross-validation](../tasks/cross-validation.qmd)
