---
title: "SNP genotypes"
---

```{r}
#| include: false
library(DT)
library(tibble)

d <- readRDS("../demo-data/main/demo_data.rds")

dt_table <- function(data, page = 10, scroll_x = FALSE) {
  DT::datatable(
    data,
    rownames = FALSE,
    class    = "cell-border compact stripe hover",
    options  = list(
      pageLength = page,
      scrollX    = scroll_x,
      autoWidth  = FALSE,
      dom        = "frtip"
    )
  )
}
```

## Schema

Both packages consume SNP genotypes as an **integer dosage matrix** with
one row per individual and one column per SNP.

| Field | Type | Description |
|---|---|---|
| Row dimension | $n$ individuals | `rownames()` must be set to the individual IDs |
| Column dimension | $p$ SNPs | `colnames()` set to SNP IDs (must match `map_snp$SNP`) |
| Cell values | integer `0/1/2` | Count of the alternate allele |
| Missingness | `NA` | Imputed upstream — neither package estimates a missingness mechanism |

The SNP matrix is paired with a **SNP map** (`d$map_snp` or `snp_map.txt`)
with one row per SNP:

| Column | Type | Description |
|---|---|---|
| `SNP` | character | SNP identifier; joins to `colnames(snp)` |
| `CHROM` | integer | Chromosome (1..5 in demo) |
| `POS` | integer | Physical position in base pairs |

## Example (from the demo dataset)

Demo dimensions: `r nrow(d$snp)` × `r ncol(d$snp)` (main) and 100 × 50 (toy).
Rownames are `IND001..IND200`; colnames are `SNP001..SNP100`.

First 8 individuals × first 10 SNPs of `main/demo_data.rds`:

```{r}
#| echo: false
preview <- as.data.frame(d$snp[1:8, 1:10])
preview <- cbind(id = rownames(preview), preview)
dt_table(preview, page = 8, scroll_x = TRUE)
```

The full $200 \times 100$ matrix is also written as a wide CSV with an
`id` column for plain-CSV consumers:

[Download geno_G1.csv (large)](../demo-data/main/geno_G1.csv){.btn .btn-sm .btn-primary}
&nbsp;
[Download geno_G1.csv (small)](../demo-data/toy/geno_G1.csv){.btn .btn-sm .btn-outline-secondary}

The matching map rows for those SNPs:

```{r}
#| echo: false
dt_table(d$map_snp[1:10, ], page = 10)
```

`snp_map.txt` is the same data as `d$map_snp`, tab-delimited so it can
be read with `read.table()` defaults:

[Download snp_map.txt (large)](../demo-data/main/snp_map.txt){.btn .btn-sm .btn-primary}
&nbsp;
[Download snp_map.txt (small)](../demo-data/toy/snp_map.txt){.btn .btn-sm .btn-outline-secondary}

## Loading

`load_data()` is exported by both `masreml` and `masbayes` and returns
the same byte-identical bundle:

```{r}
#| eval: false
# Either package's load_data() returns the canonical list.
library(masreml)
d <- load_data("large")                   # or "small"
str(d$snp)                                # int [1:200, 1:100] 1 2 2 ...
head(d$map_snp)                           # SNP / CHROM / POS

# Equivalent via masbayes:
# library(masbayes); d <- load_data("large")

# Or load the standalone CSVs derived alongside the .rds:
geno <- read.csv("demo-data/main/geno_G1.csv", check.names = FALSE)
snp_map <- read.table("demo-data/main/snp_map.txt",
                      header = TRUE, sep = "\t")
```

For your **own** SNP data — a CSV with an `id` column then one column
per SNP — convert to a matrix once and pass it on:

```{r}
#| eval: false
geno_df <- read.csv("my_genotypes.csv", check.names = FALSE)
geno    <- as.matrix(geno_df[, -1])
rownames(geno) <- geno_df$id
storage.mode(geno) <- "integer"
```

`build_G_snp()` (masreml) and `construct_snp_matrix()` (masbayes)
expect the matrix in this shape.

### Building model inputs from a SNP matrix

```{r}
#| eval: false
library(masreml)
library(masbayes)

# masreml: SNP-based GRM for BLUP / GBLUP
G_snp <- build_G_snp(d$snp)
str(G_snp)                                # 200 x 200 numeric

# masbayes: design matrix W for BayesA / BayesR.
# construct_snp_matrix() returns a list (W, freq, n, p); extract $W.
W <- construct_snp_matrix(d$snp, encoding = "vanRaden")$W
str(W)
```

## Common pitfalls

::: {.callout-warning}
**Allele coding direction.** Both packages assume the dosage counts the
**alternate** allele (0 = homozygous reference, 2 = homozygous
alternate). If your upstream pipeline reports reference-allele dosages,
you can either flip with `2L - geno` or rely on the per-marker centring
that `build_G_snp()` / `construct_snp_matrix(encoding = "vanRaden")`
applies — the resulting GRM is invariant to a uniform flip, but
per-marker effect estimates change sign.
:::

::: {.callout-warning}
**Row order must match the phenotype frame.** Neither package re-orders
for you. Always reconcile before passing to `masreml(...)` or
`run_bayesa()` / `run_bayesr()`:

```{r}
#| eval: false
stopifnot(identical(rownames(d$snp), d$pheno$id))
```
:::

::: {.callout-warning}
**Missingness.** `NA` cells must be imputed before fitting. The demo
data has no missingness by construction. For real data, common choices
are mean imputation per SNP (cheap, biased toward the population mean)
or model-based imputation (Beagle / BEAGLE for biobanks). Whatever you
choose, do it before splitting train/test so the imputed values do not
leak between sets.
:::

::: {.callout-tip}
**Minor allele frequency (MAF).** Markers with very low MAF carry
almost no information and inflate the GRM's leading eigenvalues. A
conservative pre-filter is `MAF >= 0.05`. The demo dataset's MAF is
uniformly drawn from `[0.1, 0.5]` so no filter is needed.
:::

📐 **Related theory:** [Mixed Model & BLUP Family](../theory/mixed-model-blup-family.qmd) ·
[Biallelic & Multi-allelic](../theory/biallelic-and-multiallelic.qmd) ·
[Bayesian Alphabet](../theory/bayesian-alphabet.qmd)

🛠 **Used in:** [Genomic Prediction](../tasks/genomic-prediction/index.qmd) ·
[GWAS](../tasks/gwas/index.qmd)
