---
title: "GWAS with masbayes"
---

📐 **Theory:** [BayesR](../../theory/bayes-bayesr.qmd) ·
[MCMC & EM](../../theory/bayes-mcmc-em.qmd)

This tutorial extracts GWAS-style output from a BayesR MCMC fit:
**per-marker posterior inclusion probabilities (PIPs)** and
**windowed posterior probabilities of association (WPPA)** over
physical-position windows. Both come for free from the posterior
mixture-class samples — no separate scan is needed. The full source
is adapted from
[`masbayes/examples/04_gwas.R`](https://github.com/bowo1698/masbayes/blob/main/examples/04_gwas.R).

::: {.callout-important}
**MCMC required.** PIPs and WPPA are derived from the posterior class
trajectories (`gamma_samples`). The EM path of BayesR produces only
MAP point estimates; passing `method = "em"` with a `map` argument
errors out. Use `method = "mcmc"` for the entire workflow.
:::

::: {.callout-important}
**BayesA has no equivalent.** The scaled-$t$ marginal places no mass
on exactly zero, so PIP is not defined for BayesA. `run_bayesa(..., map = ...)`
errors out by design. For frequentist GWAS, see the
[`masreml` page](masreml.qmd).
:::

## Setup

```{r}
#| message: false
#| warning: false
library(masbayes)
library(knitr)

d <- load_data("large")

# Tutorial-scale MCMC params; production should use n_iter = 20000+
mcmc <- list(n_iter = 2000L, n_burn = 1000L, n_thin = 5L, seed = 42L)
```

## SNP-based variable selection with BayesR

### Step 1 — Build the design matrix

```{r}
y_snp   <- d$pheno$y_cont_qtl_snp
snp     <- construct_snp_matrix(d$snp, encoding = "zscore")
W_snp   <- snp$W
wtw_snp <- colSums(W_snp^2)

map_snp <- d$map_snp
```

### Step 2 — Fit BayesR with the marker map attached

Passing a `map` data frame plus either `windsize` (bp window) or
`windnum` (consecutive-marker window) tells `run_bayesr()` to also
compute PIPs and WPPA from the posterior samples. The Rust sampler is
untouched; the extra output is post-processed on the R side from the
existing `gamma_samples` matrix.

```{r}
fit_snp <- run_bayesr(
  w             = W_snp,
  y             = y_snp,
  wtw_diag      = wtw_snp,
  marker_type   = "snp",
  sigma2_e_init = var(y_snp) * 0.5,
  sigma2_ah     = var(y_snp) * 0.5,
  method        = "mcmc",
  response_type = "gaussian",
  mcmc_params   = mcmc,
  map           = map_snp,
  windsize      = 5e5,
  save_rds      = FALSE,
  verbose       = FALSE
)
```

### Step 3 — Top windows by WPPA

```{r}
top10_snp <- fit_snp$gwas[order(-fit_snp$gwas$WPPA), ][seq_len(10L), ]
kable(top10_snp, digits = 3,
      caption = "Top 10 SNP windows by WPPA")
```

WPPA is the fraction of posterior samples in which **at least one**
marker inside the window had non-zero effect. It is the natural
Bayesian analogue of the windowed-posterior column produced by
[`masreml::run_gwas()`](masreml.qmd).

### Step 4 — Manhattan: $-\log_{10}(1 - \mathrm{WPPA})$

```{r}
#| fig-width: 10
#| fig-height: 4
#| dpi: 150
#| message: false
#| warning: false
wppa_cap <- 1 - 1e-3
snp_wppa_df <- data.frame(
  Wind     = fit_snp$gwas$Wind,
  Chr      = fit_snp$gwas$Chr,
  Pos      = (fit_snp$gwas$Start + fit_snp$gwas$End) / 2,
  `1-WPPA` = 1 - pmin(fit_snp$gwas$WPPA, wppa_cap),
  check.names = FALSE
)
snp_wppa_hl <- as.character(fit_snp$gwas$Wind[fit_snp$gwas$WPPA > 0.5])
if (length(snp_wppa_hl) == 0L) snp_wppa_hl <- NULL

CMplot::CMplot(
  snp_wppa_df,
  type           = "h",
  plot.type      = "m",
  LOG10          = TRUE,
  threshold      = 0.5,
  threshold.lty  = 2,
  threshold.col  = "red",
  amplify        = FALSE,
  highlight      = snp_wppa_hl,
  highlight.text = snp_wppa_hl,
  highlight.col  = NULL,
  file.output    = FALSE,
  verbose        = FALSE,
  ylab           = expression(-log[10](1 - italic(WPPA)))
)
```

### Step 5 — Manhattan: per-SNP PIP

```{r}
#| fig-width: 10
#| fig-height: 4
#| dpi: 150
#| message: false
#| warning: false
snp_pip_df <- data.frame(
  SNP    = map_snp$SNP,
  Chr    = map_snp$CHROM,
  Pos    = map_snp$POS,
  `1-PIP` = 1 - pmin(fit_snp$pip, wppa_cap),
  check.names = FALSE
)
snp_pip_hl <- as.character(map_snp$SNP[fit_snp$pip > 0.5])
if (length(snp_pip_hl) == 0L) snp_pip_hl <- NULL

CMplot::CMplot(
  snp_pip_df,
  type           = "h",
  plot.type      = "m",
  LOG10          = TRUE,
  threshold      = 0.5,
  threshold.lty  = 2,
  threshold.col  = "red",
  amplify        = FALSE,
  highlight      = snp_pip_hl,
  highlight.text = snp_pip_hl,
  highlight.col  = NULL,
  file.output    = FALSE,
  verbose        = FALSE,
  ylab           = expression(-log[10](1 - italic(PIP)))
)
```

Common PIP interpretation rules:

| PIP range | Interpretation |
|---|---|
| < 0.05 | No evidence of effect |
| 0.05 – 0.10 | Weak / suggestive |
| 0.10 – 0.50 | Moderate — worth follow-up |
| > 0.50 | Strong — likely true positive |
| > 0.95 | Decisive — calibration limit of the prior |

## Microhaplotype-based variable selection

For phased microhaplotype input, the map schema changes — one row per
block (not per allele).

```{r}
y_mh  <- d$pheno$y_cont_qtl_mh
bid   <- attr(d$mh, "block_id")
wah   <- construct_wah_matrix(d$mh, bid, d$allele_freq)
W_mh  <- wah$W_ah

map_mh <- d$map_mh
```

```{r}
fit_mh <- run_bayesr(
  w             = W_mh,
  y             = y_mh,
  wtw_diag      = colSums(W_mh^2),
  sigma2_e_init = var(y_mh) * 0.5,
  sigma2_ah     = var(y_mh) * 0.5,
  method        = "mcmc",
  response_type = "gaussian",
  mcmc_params   = mcmc,
  map           = map_mh,
  windsize      = 5e6,
  save_rds      = FALSE,
  verbose       = FALSE
)

top10_mh <- fit_mh$gwas[order(-fit_mh$gwas$WPPA), ][seq_len(10L), ]
kable(top10_mh, digits = 3,
      caption = "Top 10 microhaplotype windows by WPPA")
```

::: {.callout-note}
**Block-level PIP via max-per-block.** `fit_mh$pip_block` is a *union*
statistic (probability that at least one allele in the block is
non-zero); on small block sizes this inflates and can put every block
above the 0.5 threshold. We use **max-per-block PIP** instead —
bounded by the maximum per-allele PIP within the block — which gives
sharper QTL vs non-QTL discrimination.
:::

```{r}
mh_allele_block <- attr(W_mh, "block_id")
max_pip_block <- vapply(map_mh$block_id, function(b) {
  max(fit_mh$pip[mh_allele_block == b])
}, numeric(1L))
```

```{r}
#| fig-width: 10
#| fig-height: 4
#| dpi: 150
#| message: false
#| warning: false
mh_wppa_df <- data.frame(
  Wind     = fit_mh$gwas$Wind,
  Chr      = fit_mh$gwas$Chr,
  Pos      = (fit_mh$gwas$Start + fit_mh$gwas$End) / 2,
  `1-WPPA` = 1 - pmin(fit_mh$gwas$WPPA, wppa_cap),
  check.names = FALSE
)
mh_wppa_hl <- as.character(fit_mh$gwas$Wind[fit_mh$gwas$WPPA > 0.5])
if (length(mh_wppa_hl) == 0L) mh_wppa_hl <- NULL

CMplot::CMplot(
  mh_wppa_df,
  type           = "h",
  plot.type      = "m",
  LOG10          = TRUE,
  threshold      = 0.5,
  threshold.lty  = 2,
  threshold.col  = "red",
  amplify        = FALSE,
  highlight      = mh_wppa_hl,
  highlight.text = mh_wppa_hl,
  highlight.col  = NULL,
  file.output    = FALSE,
  verbose        = FALSE,
  ylab           = expression(-log[10](1 - italic(WPPA)))
)
```

```{r}
#| fig-width: 10
#| fig-height: 4
#| dpi: 150
#| message: false
#| warning: false
mh_pip_df <- data.frame(
  Block       = map_mh$block_id,
  Chr         = map_mh$chr,
  Pos         = (map_mh$start_pos + map_mh$end_pos) / 2,
  `1-PIP_max` = 1 - pmin(max_pip_block, wppa_cap),
  check.names = FALSE
)
mh_pip_hl <- as.character(map_mh$block_id[max_pip_block > 0.5])
if (length(mh_pip_hl) == 0L) mh_pip_hl <- NULL

CMplot::CMplot(
  mh_pip_df,
  type           = "h",
  plot.type      = "m",
  LOG10          = TRUE,
  threshold      = 0.5,
  threshold.lty  = 2,
  threshold.col  = "red",
  amplify        = FALSE,
  highlight      = mh_pip_hl,
  highlight.text = mh_pip_hl,
  highlight.col  = NULL,
  file.output    = FALSE,
  verbose        = FALSE,
  ylab           = expression(-log[10](1 - italic(PIP)[max]))
)
```

## QTL recovery sanity check

The demo data ships with ground-truth QTL positions, so we can verify
the BayesR scan recovers them — median PIP at true QTL should be
substantially higher than at non-QTL markers.

```{r}
unit_index_mh <- match(attr(W_mh, "block_id"), map_mh$block_id)
qtl_blocks    <- unique(unit_index_mh[d$qtl$mh_idx])

kable(data.frame(
  path        = c("SNP (per-marker PIP)",
                  "Microhaplotype (per-block max PIP)"),
  pip_at_QTL  = c(median(fit_snp$pip[d$qtl$snp_idx]),
                  median(max_pip_block[qtl_blocks])),
  pip_non_QTL = c(median(fit_snp$pip[-d$qtl$snp_idx]),
                  median(max_pip_block[-qtl_blocks]))
), digits = 3,
   caption = "Median PIP at true QTL vs non-QTL markers")
```

## Accessing the fit object

```{r}
#| eval: false
fit_snp$pip            # per-marker PIP (length = ncol(W_snp))
fit_snp$pip_block      # per-block PIP (length = nrow(map))
fit_snp$gwas           # window data frame: Wind, Chr, N, Start, End, WPPA
fit_snp$gwas_meta      # call metadata: windsize, windnum, n_windows, ...
fit_snp$beta_samples   # full posterior trace, n_kept x ncol(W) (MCMC only)
```

## See also

- [`run_bayesr()`](../../reference/masbayes/run_bayesr.qmd) — full
  signature including the `map`, `windsize`, `windnum` arguments
- [Theory → BayesR](../../theory/bayes-bayesr.qmd) — mixture prior,
  PIP diagnostics, $\hat\pi_0$ interpretation
- [GWAS with masreml](masreml.qmd) — frequentist EMMAX-style alternative
- [Genomic prediction with masbayes](../genomic-prediction/masbayes.qmd) —
  use the same BayesR fit for prediction
- [`masbayes/examples/04_gwas.R`](https://github.com/bowo1698/masbayes/blob/main/examples/04_gwas.R) — full source for this tutorial
