---
title: "GWAS with masreml"
---

📐 **Theory:** [Mixed Model & BLUP Family](../../theory/mixed-model-blup-family.qmd) ·
[GWABLUP](../../theory/blup-gwablup.qmd)

This tutorial runs an EMMAX-style mixed-model GWAS with `masreml`,
inspects the results, validates them against the bundled ground-truth
QTL positions, and feeds them into GWABLUP — the GWAS-assisted
re-weighting of the GRM that lifts GBLUP accuracy when the trait has a
non-infinitesimal component. The full code is adapted from
[`masreml/examples/04_gwas.R`](https://github.com/bowo1698/masreml/blob/main/examples/04_gwas.R).

## Setup

```{r}
#| message: false
#| warning: false
library(masreml)
library(knitr)

d        <- load_data("large")
ids      <- d$pheno$id
map_snp  <- d$map_snp
map_mh   <- d$map_mh
n_snp    <- nrow(map_snp)
n_blocks <- nrow(map_mh)
# Per-block midpoint for the multi-allelic Manhattan
mh_pos   <- as.integer((map_mh$start_pos + map_mh$end_pos) / 2)
```

## SNP-based GWAS

### Step 1 — Fit the mixed model

`run_gwas()` consumes a fitted `masreml` object: it pulls variance
components and the genomic relationship matrix from there. Fit the
baseline GBLUP first.

```{r}
y_snp <- setNames(d$pheno$y_cont_qtl_snp, ids)
W_snp <- d$snp
storage.mode(W_snp) <- "double"
rownames(W_snp)     <- ids

fit_snp_g <- masreml(
  y       = y_snp,
  markers = list(snp_add = W_snp),
  method  = "auto",
  trait   = "continuous",
  verbose = FALSE
)
summary(fit_snp_g)
```

### Step 2 — Run the GWAS scan

```{r}
gwas_snp <- run_gwas(
  markers     = list(snp_add = W_snp),
  y           = y_snp,
  masreml_fit = fit_snp_g,
  ref_markers = list(snp_add = W_snp)
)

str(gwas_snp)
```

The output is a list with per-marker columns including:

- `pval` — frequentist $p$ from the EMMAX-style score test
- `pp` — Bayesian model-averaged posterior probability of non-zero
  effect (used by GWABLUP for re-weighting)
- effect estimates and likelihood ratios

### Step 3 — Top peaks

```{r}
top10_snp_idx <- order(-gwas_snp$pp)[seq_len(10L)]
kable(data.frame(
  SNP   = map_snp$SNP[top10_snp_idx],
  CHROM = map_snp$CHROM[top10_snp_idx],
  POS   = map_snp$POS[top10_snp_idx],
  pval  = signif(gwas_snp$pval[top10_snp_idx], 3),
  pp    = round(gwas_snp$pp[top10_snp_idx], 3)
), caption = "Top 10 SNPs ranked by posterior probability (pp)")
```

### Step 4 — Manhattan plot ($-\log_{10}(p)$)

`CMplot` (Yin et al.) provides publication-quality genome-wide plots
with chromosome banding and threshold lines. We render in-chunk via
`file.output = FALSE`.

```{r}
#| fig-width: 10
#| fig-height: 4
#| dpi: 150
#| message: false
#| warning: false
pval_cap <- 1e-12
thr_snp_p <- 0.05 / n_snp

snp_pval_df <- data.frame(
  SNP  = map_snp$SNP,
  Chr  = map_snp$CHROM,
  Pos  = map_snp$POS,
  pval = pmax(gwas_snp$pval, pval_cap),
  check.names = FALSE
)
snp_pval_hl <- as.character(map_snp$SNP[gwas_snp$pval < thr_snp_p])
if (length(snp_pval_hl) == 0L) snp_pval_hl <- NULL

CMplot::CMplot(
  snp_pval_df,
  type           = "h",
  plot.type      = "m",
  LOG10          = TRUE,
  threshold      = thr_snp_p,
  threshold.lty  = 2,
  threshold.lwd  = 1,
  threshold.col  = "red",
  amplify        = FALSE,
  highlight      = snp_pval_hl,
  highlight.text = snp_pval_hl,
  highlight.col  = NULL,
  file.output    = FALSE,
  verbose        = FALSE,
  ylab           = expression(-log[10](italic(p)))
)
```

### Step 5 — Manhattan plot of posterior probability

`pp` is smoothed and shrunk toward zero — the Manhattan of
$-\log_{10}(1 - pp)$ emphasises the markers that GWABLUP will
upweight.

```{r}
#| fig-width: 10
#| fig-height: 4
#| dpi: 150
#| message: false
#| warning: false
pp_cap <- 1 - 1e-3
snp_pp_df <- data.frame(
  SNP    = map_snp$SNP,
  Chr    = map_snp$CHROM,
  Pos    = map_snp$POS,
  `1-pp` = 1 - pmin(gwas_snp$pp, pp_cap),
  check.names = FALSE
)
snp_pp_hl <- as.character(map_snp$SNP[gwas_snp$pp > 0.5])
if (length(snp_pp_hl) == 0L) snp_pp_hl <- NULL

CMplot::CMplot(
  snp_pp_df,
  type           = "h",
  plot.type      = "m",
  LOG10          = TRUE,
  threshold      = 0.5,
  threshold.lty  = 2,
  threshold.lwd  = 1,
  threshold.col  = "red",
  amplify        = FALSE,
  highlight      = snp_pp_hl,
  highlight.text = snp_pp_hl,
  highlight.col  = NULL,
  file.output    = FALSE,
  verbose        = FALSE,
  ylab           = expression(-log[10](1 - italic(pp)))
)
```

### Step 6 — GWAS-assisted prediction (GWABLUP)

GWABLUP re-weights the genomic relationship matrix using the per-SNP
`pp` values from the GWAS scan. Markers with high `pp` get heavier
weights; the resulting weighted GRM enters a second mixed-model fit.

```{r}
fit_snp_wa <- gwablup(
  y           = y_snp,
  markers     = list(snp_add = W_snp),
  gwas_result = gwas_snp,
  ref_markers = list(snp_add = W_snp),
  trait       = "continuous"
)

kable(data.frame(
  model = c("GBLUP", "GWABLUP"),
  h2    = c(as.numeric(fit_snp_g$varcomp$h2["snp_add"]),
            as.numeric(fit_snp_wa$varcomp$h2["snp_add"]))
), digits = 4,
   caption = "SNP architecture — h² captured by GBLUP vs GWABLUP")
```

::: {.callout-tip}
**Graceful degradation.** When the training-set GWAS is underpowered,
the `pp` values are near-uniform and the re-weighted GRM is
numerically close to the baseline GBLUP G. `gwablup()` then collapses
to GBLUP — it does not hurt accuracy in underpowered regimes, it just
stops helping.
:::

## Microhaplotype-based GWAS

The same flow applies with `markers = list(mh_add = d$mh)`. Each
microhaplotype block produces one statistic per *block* (not per
allele); the Manhattan x-axis uses physical positions from
`d$map_mh`.

```{r}
y_mh <- setNames(d$pheno$y_cont_qtl_mh, ids)
mh   <- d$mh
rownames(mh) <- ids

fit_mh_g <- masreml(
  y       = y_mh,
  markers = list(mh_add = mh),
  method  = "auto",
  trait   = "continuous",
  verbose = FALSE
)

gwas_mh <- run_gwas(
  markers     = list(mh_add = mh),
  y           = y_mh,
  masreml_fit = fit_mh_g,
  ref_markers = list(mh_add = mh)
)
```

```{r}
top10_mh_idx <- order(-gwas_mh$pp)[seq_len(10L)]
kable(data.frame(
  Block = map_mh$block_id[top10_mh_idx],
  CHR   = map_mh$chr[top10_mh_idx],
  POS   = mh_pos[top10_mh_idx],
  pval  = signif(gwas_mh$pval[top10_mh_idx], 3),
  pp    = round(gwas_mh$pp[top10_mh_idx], 3)
), caption = "Top 10 microhaplotype blocks ranked by posterior probability")
```

```{r}
#| fig-width: 10
#| fig-height: 4
#| dpi: 150
#| message: false
#| warning: false
thr_mh_p <- 0.05 / n_blocks
mh_pval_df <- data.frame(
  Block = map_mh$block_id,
  Chr   = map_mh$chr,
  Pos   = mh_pos,
  pval  = pmax(gwas_mh$pval, pval_cap),
  check.names = FALSE
)
mh_pval_hl <- as.character(map_mh$block_id[gwas_mh$pval < thr_mh_p])
if (length(mh_pval_hl) == 0L) mh_pval_hl <- NULL

CMplot::CMplot(
  mh_pval_df,
  type           = "h",
  plot.type      = "m",
  LOG10          = TRUE,
  threshold      = thr_mh_p,
  threshold.lty  = 2,
  threshold.col  = "red",
  amplify        = FALSE,
  highlight      = mh_pval_hl,
  highlight.text = mh_pval_hl,
  highlight.col  = NULL,
  file.output    = FALSE,
  verbose        = FALSE,
  ylab           = expression(-log[10](italic(p)))
)
```

```{r}
#| fig-width: 10
#| fig-height: 4
#| dpi: 150
#| message: false
#| warning: false
mh_pp_df <- data.frame(
  Block  = map_mh$block_id,
  Chr    = map_mh$chr,
  Pos    = mh_pos,
  `1-pp` = 1 - pmin(gwas_mh$pp, pp_cap),
  check.names = FALSE
)
mh_pp_hl <- as.character(map_mh$block_id[gwas_mh$pp > 0.5])
if (length(mh_pp_hl) == 0L) mh_pp_hl <- NULL

CMplot::CMplot(
  mh_pp_df,
  type           = "h",
  plot.type      = "m",
  LOG10          = TRUE,
  threshold      = 0.5,
  threshold.lty  = 2,
  threshold.col  = "red",
  amplify        = FALSE,
  highlight      = mh_pp_hl,
  highlight.text = mh_pp_hl,
  highlight.col  = NULL,
  file.output    = FALSE,
  verbose        = FALSE,
  ylab           = expression(-log[10](1 - italic(pp)))
)
```

GWABLUP on multi-allelic blocks:

```{r}
fit_mh_wa <- gwablup(
  y           = y_mh,
  markers     = list(mh_add = mh),
  gwas_result = gwas_mh,
  ref_markers = list(mh_add = mh),
  trait       = "continuous"
)

kable(data.frame(
  model = c("Microhaplotype-GBLUP", "Microhaplotype-GWABLUP"),
  h2    = c(as.numeric(fit_mh_g$varcomp$h2["mh_add"]),
            as.numeric(fit_mh_wa$varcomp$h2["mh_add"]))
), digits = 4,
   caption = "Microhaplotype architecture — h² captured by GBLUP vs GWABLUP")
```

## QTL recovery sanity check

Because the demo data ships with the true QTL positions in
`d$qtl$snp_idx` and `d$qtl$mh_idx`, we can verify that the GWAS scan
recovers them — the median `pp` at QTL markers should be substantially
higher than at non-QTL markers.

```{r}
# Map allele-level QTL indices to block ids for the MH path
qtl_block_names <- unique(d$allele_freq$haplotype[d$qtl$mh_idx])
qtl_blocks      <- as.integer(sub("block_", "", qtl_block_names))

kable(data.frame(
  path        = c("SNP", "Microhaplotype"),
  pp_at_QTL   = c(median(gwas_snp$pp[d$qtl$snp_idx]),
                  median(gwas_mh$pp[qtl_blocks])),
  pp_non_QTL  = c(median(gwas_snp$pp[-d$qtl$snp_idx]),
                  median(gwas_mh$pp[-qtl_blocks]))
), digits = 3,
   caption = "Median posterior probability at true QTL vs non-QTL markers")
```

The ratio between the two columns is a discrimination diagnostic — a
ratio of ~2× or better indicates the GWAS resolves the QTL signal
above background noise on this demo. On real data, expect a similar
or stronger separation when the trait is well powered and the marker
density is appropriate.

## Working with the GWAS output

```{r}
# Markers passing a Bonferroni threshold
bonferroni <- 0.05 / n_snp
sum(gwas_snp$pval < bonferroni)

# Markers passing a posterior-probability threshold
sum(gwas_snp$pp > 0.5)
```

`run_gwas()` does **not** apply multiple-testing correction itself —
the choice (Bonferroni / FDR / posterior threshold) is left to the
analyst, since the appropriate method depends on whether you want a
strict family-wise error guarantee, an expected-FDR control, or a
Bayesian posterior threshold.

## See also

- [`run_gwas()`](../../reference/masreml/run_gwas.qmd) ·
  [`gwablup()`](../../reference/masreml/gwablup.qmd)
- [Theory → GWABLUP](../../theory/blup-gwablup.qmd) — Meuwissen 2024 framework
- [GWAS with masbayes](masbayes.qmd) — BayesR PIP-based alternative
- [`masreml/examples/04_gwas.R`](https://github.com/bowo1698/masreml/blob/main/examples/04_gwas.R) — full source for this tutorial
