---
title: "Genomic data encoding"
---

A prediction model cannot compare a training animal with a candidate unless
their genotypes use exactly the same representation. This page turns marker
states into matrices while keeping a coding convention separate from biological
meaning.

## SNP dosage counts a chosen allele

At a biallelic SNP, let $M_{ij}$ be the number of copies of a chosen counted
allele carried by individual $i$ at marker $j$. Thus $M_{ij}$ is 0, 1, or 2.
The reference and counted alleles are labels: reversing them changes the
numbers and coefficient signs, but not the individual's DNA or the evidence
that a locus is causal. For $n$ animals and $p$ SNPs, $\mathbf M$ is an
$n\times p$ dosage matrix and the allele frequency is

$$f_j=\frac{1}{2n}\sum_{i=1}^{n}M_{ij}.$$

Centred coding removes the expected dosage under the chosen reference
population,

$$W_{ij}=M_{ij}-2f_j,$$

giving an $n\times p$ matrix $\mathbf W$. Optional variance scaling changes
the relative weighting of marker columns and must be stated. Missing calls
need a documented imputation or model strategy; silently treating them as zero
copies changes allele frequencies and similarity.

```{r}
#| label: centre-snp-dosages
#| code-fold: true
d <- readRDS("../../demo-data/main/demo_data.rds")
M <- d$snp[1:6, c("SNP007", "SNP008", "SNP017"), drop = FALSE]
allele_frequency <- colMeans(M) / 2
W <- sweep(M, 2, 2 * allele_frequency, FUN = "-")
M
allele_frequency
colMeans(W)
```

The three frequencies in this six-animal teaching view are 0.25, 0.417, and
0.417. The centred column means are zero apart from floating-point rounding.
For prediction, frequencies and imputation values must be estimated from the
training population and then applied unchanged to candidates. Re-estimating
them in a test set changes the fitted representation and can leak information.

## A microhaplotype is categorical before it is encoded

A phased microhaplotype records one local haplotype state on each chromosome
copy. Its stored identifiers are category labels, not dosages: identifier 8 is
not twice identifier 4. A useful representation first creates one column per
observed allele, then counts its 0, 1, or 2 copies for every individual. This
one-hot or allele-count design retains multiallelic states without inventing an
order among them.

```{r}
#| label: encode-microhaplotype
#| code-fold: true
mh <- d$mh[1:6, 5:6, drop = FALSE]
colnames(mh) <- c("block_3_copy_1", "block_3_copy_2")
alleles <- sort(unique(as.vector(mh)))
mh_counts <- sapply(alleles, function(a) rowSums(mh == a))
mh
mh_counts
```

The project representation contains labels 4, 7, and 8 in this block. The
count matrix says how many copies of each label occur; it does not say that
allele 8 is quantitatively larger than allele 4. The haplotype and
[microhaplotype input guide](../../input-data/microhaplotypes.qmd) explains
how phase and compact input objects are obtained.

## Encoding fixes the comparison population

Centred SNP columns and multiallelic allele-count columns are both statistical
representations. They depend on marker order, allele definitions, phase, and
the population used for frequency estimates. They do not by themselves prove
identity by descent, establish a causal variant, or make one marker type
universally better. The next page uses the centred SNP matrix to describe
realised genomic similarity through a genomic relationship matrix.

## Exercises

1. A SNP has counted-allele dosage 2. What happens to its dosage if the other allele becomes the counted allele?

::: {.callout-note collapse="true" title="Solution"}
It becomes 0. The coding changes from $M$ to $2-M$, but the diploid genotype
and biological evidence do not change.
:::

2. Why must a candidate use allele frequencies estimated in the training population?

::: {.callout-note collapse="true" title="Solution"}
The fitted parameters refer to the training-derived centring. Re-estimating
frequencies in candidates changes the columns supplied to the model and can
leak information during evaluation.
:::

3. Why is it invalid to regress directly on a compact haplotype label such as 4, 7, or 8?

::: {.callout-note collapse="true" title="Solution"}
The labels identify categories, not ordered measurements. A direct numeric
slope would impose artificial distances between haplotype states.
:::
