---
title: "GBLUP — Genomic BLUP"
---

GBLUP replaces the expected pedigree kinship of **A** with *realised*
allele sharing measured directly from markers. Two sub-methods share
the same MME; only the relationship matrix **K** changes.

## SNP-GBLUP (VanRaden 2008) {#snp-gblup-vanraden-2008}

$$
\mathbf{K} = \mathbf{G}_u
  = \frac{\mathbf{W}\mathbf{W}^\top}{2\sum_{j=1}^{p} p_j(1-p_j)},
\qquad
W_{ij} = M_{ij} - 2p_j,
$$ {#eq-gblup-grm}

where $M_{ij} \in \{0,1,2\}$ is the SNP dosage for individual $i$ at
locus $j$, and $p_j$ is the allele frequency at locus $j$.

### Why Centering with $2p_j$ Matters

Subtracting $2p_j$ from each column centres the SNP matrix at zero
mean, which ensures the GRM diagonal averages to 1 (comparable to
**A**) and gives *greater relative weight to rare alleles*. A rare
allele ($p_j$ small) produces a large deviation $W_{ij}$ when
present, so two individuals who share that rare allele appear more
similar in **G**. This is biologically appropriate: shared rare
alleles are more likely to be IBD than shared common alleles.

### Allele Frequency Choice

::: {.callout-warning}
**Use base-population allele frequencies.**
VanRaden (2008) showed that allele frequencies from the *base
(unselected founder) population* — rather than from the genotyped
cohort — produce a **G** whose diagonal correctly reflects inbreeding
relative to the founder generation and maximises correlation with
the pedigree **A**. Using sample frequencies from a selected
population biases $G_{ij}$ and can produce genomic inbreeding
coefficients that are negative or inverted relative to pedigree
expectations. `masreml` computes $p_j$ from the training set by
default; re-centring for a new prediction population requires
supplying the reference frequencies explicitly.
:::

### Three Equivalent Formulations

<details>
<summary>Full derivation (advanced)</summary>

VanRaden (2008) showed three algebraically equivalent ways to obtain
GEBVs $\hat{\mathbf{a}}$ — useful to know because each scales
differently with $n$ (individuals) versus $p$ (markers):

**Method 1 — Standard GRM (above).** $\mathbf{G} = \mathbf{ZZ'} /
[2\sum p_j(1-p_j)]$. Invert the $n \times n$ matrix **G**;
efficient when $n \ll p$.

**Method 2 — Per-marker weighted GRM.**
$\mathbf{G} = \mathbf{Z}\mathbf{D}\mathbf{Z}^\top$ with
$D_{jj} = 1/[p \cdot 2p_j(1-p_j)]$. Each marker is weighted by
the inverse of its expected variance, giving equal-variance
SNP-BLUP. This is the foundation for the diagonal weight matrix
used in GWABLUP.

**Method 3 — Regression on A.**
$\mathbf{MM}^\top = g_0\mathbf{11}^\top + g_1\mathbf{A} + \mathbf{E}$,
then $\mathbf{G} = (\mathbf{MM}^\top - g_0\mathbf{11}^\top)/g_1$.
Does not require allele frequency estimates; useful when frequencies
are unknown or unreliable.

All three produce the same GEBV ranking when the training set is
large. Method 1 is the standard in `masreml` for computational
efficiency and interpretability.
</details>

---

## Microhaplotype-GBLUP (Da 2015) {#mh-gblup-da-2015}

When markers are microhaplotypes — short multi-allelic blocks (<200 bp
spanning 2–5 tightly linked SNPs) — the biallelic centering formula of
VanRaden no longer applies. Da (2015) derived a multi-allelic
extension that treats each haplotype block as a single locus with $h$
alleles.

$$
\mathbf{K} = \mathbf{A}_h = \mathbf{T}_{\alpha h}\mathbf{T}_{\alpha h}^\top,
\qquad
\mathbf{T}_{\alpha h} = \frac{\mathbf{W}_{\alpha h}}{k_{\alpha h}^{1/2}},
$$ {#eq-mh-grm}

where $\mathbf{W}_{\alpha h}$ is the additive coding matrix for
haplotype block $h$ (see
[Biallelic & Multi-allelic](biallelic-and-multiallelic.qmd#mh-design-matrix-w_alpha-h)
for the full coding rule) and $k_{\alpha h} =
\overline{\mathrm{diag}(\mathbf{W}_{\alpha h}\mathbf{W}_{\alpha h}^\top)}$
is the scaling constant that keeps $E[A_{h,ii}] = 1$, consistent
with a non-inbred individual in the pedigree.

### Why Microhaplotypes Preserve More Signal Under Selection

A microhaplotype block groups 2–5 SNPs within <200 bp — a window so
short that recombination *within* the block is rare. The block
haplotype therefore stays intact across generations longer than any
individual SNP allele would. This means $\mathbf{A}_h$ captures IBD
at finer resolution than a single SNP, and is more robust to LD
erosion under directional selection — a key advantage in breeding
programmes where marker-trait LD degrades generation by generation.
Simulation evidence (Wibowo, Chapter 1) shows that GBLUP with microhaplotypes maintains
prediction accuracy more stably across generations of selection than
SNP-GBLUP under equivalent marker density.

### Key Assumptions

::: {.callout-note}
GBLUP with microhaplotypes assumes phasing is reliable. Phase errors introduce noise
into $\mathbf{W}_{\alpha h}$ and propagate into $\mathbf{A}_h$. For
blocks longer than 5 SNPs, recombination within the block is no
longer negligible and $E[\mathbf{A}_h]$ drifts below the expected
**A** value — block size is a tuning parameter and should be validated
empirically for the population and marker panel at hand.
:::

---

## See it in Code

**SNP-GBLUP**

```{r}
#| eval: false
library(masreml)

d <- load_data("large")
y <- d$pheno$y_cont_qtl_snp
names(y) <- d$pheno$id
X_fixed <- stats::model.matrix(~ sex, data = d$pheno)

# G_u is built internally from the marker matrix using VanRaden Method 1
fit_gblup <- masreml(
  y       = y,
  X       = X_fixed,
  markers = list(snp_add = d$snp),
  trait   = "continuous"
)
summary(fit_gblup)
```

**GBLUP with microhaplotypes**

```{r}
#| eval: false
library(masreml)

d <- load_data("large")
y <- d$pheno$y_cont_qtl_snp
names(y) <- d$pheno$id
X_fixed <- stats::model.matrix(~ sex, data = d$pheno)

# W_alpha_h coding constructed internally by masreml from phased microhaplotype genotypes
fit_mhblup <- masreml(
  y       = y,
  X       = X_fixed,
  markers = list(mh_add = d$mh),
  trait   = "continuous"
)
summary(fit_mhblup)
```

## References

- VanRaden, P. M. (2008). Efficient methods to compute genomic
  predictions. *Journal of Dairy Science*, 91(11), 4414–4423.
  [doi:10.3168/jds.2007-0980](https://doi.org/10.3168/jds.2007-0980).
- Da, Y. (2015). Multi-allelic haplotype model based on genetic
  partition for genomic prediction and variance component estimation
  using SNP markers. *BMC Genetics*, 16, 144.
  [doi:10.1186/s12863-015-0301-1](https://doi.org/10.1186/s12863-015-0301-1).
