---
title: "Genomic prediction with masreml"
---

📐 **Theory:** [Mixed Model & BLUP Family](../../theory/mixed-model-blup-family.qmd) ·
[REML Algorithms](../../theory/blup-reml.qmd) ·
[GBLUP](../../theory/blup-gblup.qmd) ·
[Continuous and Binary Trait](../../theory/continuous-and-binary-trait.qmd)

This tutorial fits genomic prediction models with `masreml` end-to-end:
GBLUP on SNP, Microhaplotype-GBLUP on phased microhaplotypes,
cross-validation, and the threshold model for binary traits. All code
chunks evaluate against the bundled demo dataset, so the tables and
plots below are the actual output you get from running the code.

## Setup

```{r}
#| message: false
#| warning: false
library(masreml)
library(ggplot2)
library(knitr)

d <- load_data("large")

# Continuous trait, SNP architecture
y <- setNames(d$pheno$y_cont_qtl_snp, d$pheno$id)

# Fixed-effect design matrix — sex as the single covariate
X_fixed <- stats::model.matrix(~ sex, data = d$pheno)
```

## SNP workflow

### Full-data fit — GBLUP

`masreml()` fits the standard mixed model
$\mathbf{y} = \mathbf{X}\mathbf{b} + \mathbf{Z}\mathbf{u} + \mathbf{e}$,
estimates variance components by REML (default `method = "auto"` —
HE-initialised AI-REML for $n < 50{,}000$), and returns a `masreml`
object.

```{r}
fit_gblup <- masreml(
  y       = y,
  X       = X_fixed,
  markers = list(snp_add = d$snp),
  trait   = "continuous",
  verbose = FALSE
)

summary(fit_gblup)
```

Inspect the GEBV distribution:

```{r}
#| fig-width: 7
#| fig-height: 4
ggplot(data.frame(gebv = fit_gblup$gebv$snp_add),
       aes(x = gebv)) +
  geom_histogram(bins = 30, fill = "#2c7fb8", colour = "white") +
  labs(x = "GEBV (snp_add)", y = "Count",
       title = "SNP-GBLUP GEBVs — full-data fit") +
  theme_minimal()
```

### Predict on held-out candidates

The bundled demo provides a `train_idx` / `test_idx` split:

```{r}
train_ids <- d$pheno$id[d$train_idx]
test_ids  <- d$pheno$id[d$test_idx]

fit_train <- masreml(
  y       = y[train_ids],
  X       = X_fixed[d$train_idx, , drop = FALSE],
  markers = list(snp_add = d$snp[train_ids, , drop = FALSE]),
  trait   = "continuous",
  verbose = FALSE
)

pred <- predict(
  fit_train,
  markers_new   = list(snp_add = d$snp[test_ids, , drop = FALSE]),
  markers_train = list(snp_add = d$snp[train_ids, , drop = FALSE]),
  train_ids     = train_ids,
  test_ids      = test_ids,
  X_new         = X_fixed[d$test_idx, , drop = FALSE],
  y_new         = y[test_ids]
)

kable(as.data.frame(pred$metrics),
      digits  = 4,
      caption = "Held-out accuracy on the test split")
```

### Cross-validation — `cv_masreml()`

`cv_masreml()` automates K-fold or family-based splits and refits the
model on each training fold. Default is 5-fold random; pass
`scheme = "family"` and a family-id vector for genuine
candidate-selection accuracy.

```{r}
#| message: false
cv <- cv_masreml(
  y       = y,
  X       = X_fixed,
  markers = list(snp_add = d$snp),
  folds   = 5L,
  scheme  = "random",
  seed    = 123L
)

fold_tbl <- data.frame(
  fold     = seq_along(cv$accuracy_fold),
  accuracy = cv$accuracy_fold,
  bias     = cv$bias_fold
)
kable(fold_tbl, digits = 4, caption = "Per-fold test accuracy")
```

```{r}
kable(data.frame(metric = c("Mean accuracy", "Mean bias"),
                 value  = c(cv$accuracy, cv$bias)),
      digits  = 4,
      caption = "Overall 5-fold CV summary")
```

```{r}
#| fig-width: 7
#| fig-height: 3.5
ggplot(fold_tbl, aes(x = factor(fold), y = accuracy)) +
  geom_col(fill = "#2c7fb8") +
  geom_hline(yintercept = mean(fold_tbl$accuracy),
             linetype = "dashed", colour = "red") +
  labs(x = "Fold", y = "Test accuracy (Pearson r)",
       title = "5-fold CV — SNP-GBLUP") +
  theme_minimal()
```

### Binary trait {#binary-trait}

For 0/1 outcomes, set `trait = "binary"` and pick a link function.
`masreml` runs a single-step Laplace approximation with the working
response, returning the posterior mode of the liability-scale GEBV
plus the fitted probability $\hat{p}_i$.

```{r}
y_bin <- setNames(d$pheno$y_bin_qtl_snp, d$pheno$id)

fit_bin <- masreml(
  y       = y_bin,
  X       = X_fixed,
  markers = list(snp_add = d$snp),
  trait   = "binary",
  link    = "logit",
  verbose = FALSE
)

summary(fit_bin)
```

`fit_bin$h2` is reported on the liability scale (denominator
$\sigma_u^2 + \pi^2/3$ for logit). Predictions return both the
liability-scale GEBV and the observed-scale probability:

```{r}
pred_bin <- predict(fit_bin, y_new = y_bin)
kable(as.data.frame(pred_bin$metrics),
      digits  = 4,
      caption = "Training-set metrics, binary trait (AUC = discrimination)")
```

## Microhaplotype workflow

Switching from SNP to phased microhaplotype markers only changes the
`markers` slot. Everything else — fitting, summary, prediction,
cross-validation, binary trait — is identical.

```{r}
y_mh <- setNames(d$pheno$y_cont_qtl_mh, d$pheno$id)

fit_mhblup <- masreml(
  y       = y_mh,
  X       = X_fixed,
  markers = list(mh_add = d$mh),
  trait   = "continuous",
  verbose = FALSE
)
summary(fit_mhblup)
```

Cross-validation on microhaplotype architecture:

```{r}
#| message: false
cv_mh <- cv_masreml(
  y       = y_mh,
  X       = X_fixed,
  markers = list(mh_add = d$mh),
  folds   = 5L,
  scheme  = "random",
  seed    = 123L
)

fold_mh_tbl <- data.frame(
  fold     = seq_along(cv_mh$accuracy_fold),
  accuracy = cv_mh$accuracy_fold,
  bias     = cv_mh$bias_fold
)
kable(fold_mh_tbl, digits = 4,
      caption = "5-fold CV — Microhaplotype-GBLUP")
```

Binary trait on microhaplotype architecture:

```{r}
y_bin_mh <- setNames(d$pheno$y_bin_qtl_mh, d$pheno$id)

fit_bin_mh <- masreml(
  y       = y_bin_mh,
  X       = X_fixed,
  markers = list(mh_add = d$mh),
  trait   = "binary",
  link    = "logit",
  verbose = FALSE
)
summary(fit_bin_mh)
```

The W~αh~ coding (Da 2015) is constructed internally by `masreml`; no
separate preprocessing call is needed when you go through `masreml()`
directly. If you want to inspect or reuse the coded matrix, build it
explicitly with `masbayes::construct_wah_matrix()` — see the
[`masbayes` tutorial](masbayes.qmd) and the
[Microhaplotype input page](../../input-data/microhaplotypes.qmd).

## Working with the fit object

Beyond `summary()` and `predict()`:

```{r}
kable(varcomp(fit_gblup),
      digits  = 4,
      caption = "Variance components (SNP-GBLUP)")
```

```{r}
acc <- compute_accuracy(gebv = fit_gblup$total_gebv, y = y)
kable(as.data.frame(acc),
      digits  = 4,
      caption = "Training-set accuracy (SNP-GBLUP)")
```

The full set of accessor methods:

```{r}
#| eval: false
fit_gblup$gebv          # named list of per-component GEBVs
fit_gblup$total_gebv    # aggregated total GEBV vector
fit_gblup$converged     # convergence flag
fit_gblup$loglik        # REML log-likelihood
fit_gblup$n_iter        # iterations to convergence
```

See [`predict.masreml`](../../reference/masreml/predict.masreml.qmd)
and [`evaluate_prediction`](../../reference/masreml/evaluate_prediction.qmd)
for the full argument lists.

## See also

- [`masreml()`](../../reference/masreml/masreml.qmd) — main fitting function
- [`cv_masreml()`](../../reference/masreml/cv_masreml.qmd) — cross-validation
- [`predict.masreml`](../../reference/masreml/predict.masreml.qmd) — prediction on new individuals
- [Theory → REML Algorithms](../../theory/blup-reml.qmd) — variance-component estimation engines
- [Genomic prediction with masbayes](masbayes.qmd) — Bayesian alternative
